### 2022-02-28 manuscript/manuscript.tex
Add build classifier script

\documentclass[]{article}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{authblk}
\graphicspath{ {./images/} }
\linespread{1.15}
\textwidth=380pt

\title{Variant calling with transformers}
\author[1]{Brendan O'Fallon}
\author[1]{Ashini Bolia}
\author[1]{Jacob Durtschi}
\author[1]{Luobin Yang}

\affil[1]{ARUP Institute for Clinical and Experimental Pathology, Salt Lake City, UT}
\date{}
\begin{document}

\maketitle

\begin{abstract}
	Detection of germline variants in next-generation sequencing data is an essential element of modern genomics. Variant detection tools typically rely on a variety of statistical algorithms and heuristic techniques to identify variants. Here we describe a new approach that replaces the handcrafted statistical methods with a single, end-to-end deep learning model. Our model frames variant detection as a sequence-to-sequence modeling task, and employs a transformer-based architecture to translate alignment columns into predicted haplotype sequences. 
\end{abstract}

\section{Introduction}

Accurate reconstruction of the DNA sequence in a sample is a fundamental component of modern genomics analysis. Next Generation Sequencing (NGS) workflows typically begin by aligning sequenced fragments to a reference genome, optionally performing some alignment post-processing steps such as identification of duplicate fragments, and then identifying sequence variants relative to the reference genome. The variant identification step is challenging for many reasons, including technical artifacts in the sequencing process, improper alignments due to sequence homology or low sequence complexity, and the stochastic nature of the library preparation and amplification steps [citation here would be nice].

Variant discovery tools use a variety of statistical techniques to identify variants. Early callers, such as samtools / mpileup (Li et al. 2009) and the UnifiedGenotyper tool from the Genome Analysis ToolKit (GATK, De Pristo et al. 2011) examined each alignment column individually to determine genotype. Because these tools do not realign reads, they rely on the alignments produced by read mapping tools such as BWA (Li \& Durbin 2009) . Later tools, such as the HaplotypeCaller from the GATK and Strelka 2 (Kim et al. 2018) incorporated a re-alignment of reads in each region of interest, resulting in a significant improvement in the precision and specificity of variant calls. The HaplotypeCaller tool constructs de Bruijn graphs from k-mers present in the reads in a region, realigns reads to the graphs, and uses a pair Hidden Markov Model (HMM) to determine the most likely path through the graph.

More recent tools have incorporated elements of deep learning to further improve accuracy. DeepVariant (Poplin et al. 2018) builds on the HaplotypeCaller approach, but adds a Convolutional Neural Net (CNN) to classify variants as true or false positive detections. HELLO (Ramachandran et al. 2020), designed to work on hybrid short- and long-read datasets, employs a mixture-of-experts approach with separate 1-dimensional convolutions across the read and allele dimensions of the input. While these tools deliver accurate results, they rely heavily on hand-crafted statistical algorithms and heuristics to generate and assess candidate alleles.

Here, we show that none of the bespoke statistical apparatus developed in earlier tools is required for accurate variant identification, and that a robust variant detection model can be learned directly from raw NGS data. To achieve this, we recast variant detection as a sequence-to-sequence translation problem. The series of bases aligning to a given reference position is a single input token, and the sequence of these alignment columns constitutes the input to the model. The output of the model is the two predicted haplotype sequences in the sample. The sequence-to-sequence translation is performed by a transformer-based deep neural network (Vaswani et al. 2017). Transformer models have achieved state-of-the-art results in many sequence-to-sequence problems,

\section{Methods}

\subsection{Model architecture}

Our model consists of a transformer-based encoder and simple, fully connected decoder. The encoder component is an unmodified transformer with GeLU activations as implemented in PyTorch 1.10 (Paszke et al. 2019). We add an additional fully connected layer prior to the transformer encoders which embeds the encoded basecalls in $d$ dimensions, where $d=12$ for the analyses here. The embedded basecalls are 'flattened' along the read dimension, producing an input with size $dr$ where $r$ is the number of reads. Instead of the typical 1-dimensional positional encoding, we employed a 2-dimensional encoding, enabling the model to retain information about which read contained a given sequenced base.

The decoder consists of a single fully-connected layer followed by two additional fully connected components that each generate a single haplotype. Predictions are discrete probability distributions over the four possible bases at each predicted output position. The decoder is of fixed size, and each predicted haplotype contains the same number of predicted bases as the number of positions $g$ in the input.

\subsection{NGS data}

We obtained training data from 17 whole genomes sequenced from 5 Genome-in-a-Bottle cell lines. Eight samples were prepped with Illumina Nextera DNA Flex kit and the remained were prepared with the Illumina TruSeq PCR-free kit. All samples were sequenced on an Illumina NovaSeq 6000 instrument in 2x150 mode, to an approximate read depth of 50. After conversion to fastq, the sequenced reads were aligned to human reference genome GRCh37 with the GEM-mapper (v3, Marco-Sola et al. 2012) and were sorted and converted to CRAM format with samtools version 1.9 (Li et al. 2009). No additional refinements, such as duplicate read marking, base-quality score recalibration, or indel realignment were performed.

\subsection{Training data}

Training tensors were generated by selecting regions 150bp in width from alignments. The encoding process included fetching all reads aligning to the desired region. If more reads were found then a specified maximum read depth (100 for all analyses included in this work), reads were randomly downsampled to not exceed the maximum size. Reads were sorted by the reference coordinate of the first aligned base. For each aligned base in the selected reads nine features were encoded; the first four were the one-hot encoded base call, followed by base quality, two flags indicating if base `consumed' reference base (i.e. was not an insertion) or consumed a sequence read base (was not a deletion), and additional flags indicating sequence read direction and clipping status.  No 'gap' tokens or other special handling was performed for insertions or deletions. In addition to sequenced reads, the first row in each encoded region was the reference sequence. For this special row we inserted a base quality of 100 for every position and did not set any of the other flags. Resulting tensors had dimension $[g, r, 9]$, where $g$ represents genomic position, $r$ represents read, and 9 is the number of features.

Target / label sequences were produced by obtaining truth variants from the Genome-in-a-Bottle VCF files for each sample and inserting the variants into the reference sequence. Two sequences were generated for each region, one representing each haplotype. In situations where phasing of the variants was ambiguous, the reads in the sequenced sample were examined to determine phase status. Briefly, all possible genotypes (pairs of haplotype sequences) were generated and reads were aligned via Smith-Waterman to the possible haplotypes, and the highest scoring genotypes selected as the most likely phasing. This phasing procedure was only performed if multiple heterozygous variants were present in the region.

To select regions to include for training data, we developed a scheme to sample regions in a biased manner, prioritizing regions containing variants and, especially, regions with multiple or complex variants. Regions of the reference genome overlapping the high-confidence regions from Genome-in-a-Bottle were subdivided into 150bp nonoverlapping windows, and these regions labelled according to the presence of variants. Separate labels were generated for regions conataining a single SNV, deletion, or insertion, as well as regions containing multiple insertion-deletions, or those containing variants intersecting  low-complexity regions. We additionally included `true negative' regions where no known variant was present. The total number of regions generated per sample was 325,000, yielding 5.53M total training regions across our 17 samples.

\subsection{Variant detection}

Given an alignment file in BAM or CRAM format,

\section{Results}

\subsection{Model size experiments}

 small (5M) medium (10M), large (25M) pretty big (50M)

 Ablate the 2D positional encoding.

 \section{Discussion}

 Convolutions - do they make sense?

 \section{Availability}

 Source code for is available via git at https://github.com/ARUP-NGS/somewhere

 \section{References}

 \vspace{8pt}
 Conrad, Donald F., et al. "Origins and functional impact of copy number variation in the human genome." Nature 464.7289 (2010): 704.

\vspace{8pt}
DePristo, Mark A., et al. "A framework for variation discovery and genotyping using next-generation DNA sequencing data." Nature genetics 43.5 (2011): 491-498.

\vspace{8pt}
Kim, Sangtae, et al. "Strelka2: fast and accurate calling of germline and somatic variants." Nature methods 15.8 (2018): 591-594.

\vspace{8pt}
Li, Heng, et al. "The sequence alignment/map format and SAMtools." Bioinformatics 25.16 (2009): 2078-2079.

\vspace{8pt}
Li, Heng, and Richard Durbin. "Fast and accurate short read alignment with Burrows–Wheeler transform." Bioinformatics 25.14 (2009): 1754-1760.

\vspace{8pt}
Marco-Sola S., Sammeth M., Guigó R., Ribeca P. "The GEM mapper: fast, accurate and versatile alignment by filtration". Nat Methods. (2012);9(12):1185-1188. doi:10.1038/nmeth.2221

\vspace{8pt}
Poplin, Ryan, et al. "A universal SNP and small-indel variant caller using deep neural networks." Nature biotechnology 36.10 (2018): 983-987.

\vspace{8pt}
Paszke, A., et al. "PyTorch: An Imperative Style, High-Performance Deep Learning Library." In Advances in Neural Information Processing Systems 32 (2019):8024–8035. Curran Associates, Inc. Retrieved from http://papers.neurips.cc/paper/9015-pytorch-an-imperative-style-high-performance-deep-learning-library.pdf

\vspace{8pt}
Ramachandran, Anand, et al. "HELLO: A hybrid variant calling approach." bioRxiv (2020).

 \vspace{8pt}
 \subsection{Ablation experiments}
Vaswani, Ashish, et al. "Attention is all you need." Advances in neural information processing systems 30 (2017).

\end{document}

### 2022-02-28 scripts/buildclf.py

@@ -1,0 +1,67 @@
+
+import sys
+import yaml
+import numpy as np
+import scipy.stats as stats
+import pysam
+import pickle
+import sklearn
+from sklearn.ensemble import RandomForestClassifier
+
+import logging
+
+logger = logging.getLogger(__name__)
+logging.basicConfig(format='[%(asctime)s]  %(name)s  %(levelname)s  %(message)s',
+                    datefmt='%m-%d %H:%M:%S',
+                    level=logging.INFO)
+
+def extract_feats(vcf):
+    allfeats = []
+    for var in pysam.VariantFile(vcf, ignore_truncation=True):
+        feats = []
+        feats.append(var.qual)
+        feats.append(1 if "PASS" in var.filter else 0)
+        feats.append(1 if "LowCov" in var.filter else 0)
+        feats.append(1 if "SingleCallHet" in var.filter else 0)
+        feats.append(1 if "SingleCallHom" in var.filter else 0)
+        feats.append(min(var.info['QUALS']))
+        feats.append(max(var.info['QUALS']))
+        feats.append(var.info['WIN_VAR_COUNT'][0])
+        feats.append(var.info['WIN_CIS_COUNT'][0])
+        feats.append(var.info['WIN_TRANS_COUNT'][0])
+        feats.append(var.info['STEP_COUNT'][0])
+        feats.append(min(var.info['WIN_OFFSETS']))
+        feats.append(max(var.info['WIN_OFFSETS']))
+        allfeats.append(np.array(feats))
+    return allfeats
+
+def save_model(mdl, path):
+    logger.info(f"Saving model to {path}")
+    with open(path, 'wb') as fh:
+        pickle.dump(mdl, fh)
+        
+def load_model(path):
+    logger.info(f"Loading model from {path}")
+    with open(path, 'rb') as fh:
+        return pickle.load(fh)
+
+def train_model(conf):
+    alltps = []
+    for tpf in conf['tps']:
+        tps = extract_feats(tpf)
+    for fpf in conf['fps']:
+        fps = extract_feats(fpf)
+
+    logger.info(f"Loaded {len(tps)} TP and {len(fps)} FPs")
+    feats = tps + fps
+    y = np.array([1.0 for _ in range(len(tps))] + [0.0 for _ in range(len(fps))])
+    clf = RandomForestClassifier(n_estimators=100, max_depth=None, random_state=0, max_features=None)
+    clf.fit(feats, y)
+    return clf
+
+
[…]

### 2022-02-28 scripts/buildclf.py
Update classifier

@@ -65,3 +65,6 @@ def main(conf_file, output):
     conf = yaml.safe_load(open(conf_file).read())
     model = train_model(conf)
     save_model(model, output)
+
+if __name__=="__main__":
+    main(sys.argv[1], sys.argv[2])

### 2022-03-01 call.py
Better handling for low coverage areas

@@ -18,12 +18,6 @@ logger = logging.getLogger(__name__)
 DEVICE = torch.device("cuda") if hasattr(torch, 'cuda') and torch.cuda.is_available() else torch.device("cpu")
 
 
-class LowReadCountException(Exception):
-    """
-    Region of bam file has too few spanning reads for variant detection
-    """
-    pass
-
 
 def gen_suspicious_spots(aln, chrom, start, stop, refseq):
     """

### 2022-03-01 model.py
train script

@@ -103,7 +103,7 @@ class VarTransformer(nn.Module):
         self.embed_dim = nhead * embed_dim_factor
         self.fc1_hidden = 12
 
-        self.fc1 = nn.Linear(feature_count, self.fc1_hidden)
+        self.fc1 = nn.Linear(feature_count + 1, self.fc1_hidden)
         self.fc2 = nn.Linear(read_depth * self.fc1_hidden, self.embed_dim)
 
         # self.pos_encoder = PositionalEncoding(self.embed_dim, p_dropout)
@@ -116,8 +116,8 @@ class VarTransformer(nn.Module):
     def forward(self, src):
 
         # r contains a 1 if the read base matches the reference base, 0 otherwise
-        #r = (src[:, :, :, 0:4] * src[:, :, 0:1, 0:4]).sum(dim=-1) # x[:, :, 0:1..] is the reference seq
-        #src = torch.cat((src, r.unsqueeze(-1)), dim=3).to(self.device)
+        r = (src[:, :, :, 0:4] * src[:, :, 0:1, 0:4]).sum(dim=-1) # x[:, :, 0:1..] is the reference seq
+        src = torch.cat((src, r.unsqueeze(-1)), dim=3).to(self.device)
 
 
         src = self.elu(self.fc1(src))

### 2022-03-01 scripts/run_pregen.sh

@@ -19,11 +19,11 @@ ds2s=/uufs/chpc.utah.edu/common/home/u0379426/src/dnaseq2seq/main.py
 CONF=/uufs/chpc.utah.edu/common/home/u0379426/src/dnaseq2seq/wgs_multindel_conf.yaml
 
 
-DEST=/uufs/chpc.utah.edu/common/home/arup-storage3/u0379426/pregen_wgs_w150_nochr21or22_big/
+DEST=/uufs/chpc.utah.edu/common/home/arup-storage3/u0379426/pregen_wgs_w150_nochr21or22_bigger/
 
 BATCH_SIZE=256
 
 mkdir -p $DEST
 
-$PYTHON $ds2s pregen -c $CONF -d $DEST --threads 16 --batch-size $BATCH_SIZE > $DEST/stdout.log 
+$PYTHON $ds2s pregen -c $CONF -d $DEST --threads 18 --batch-size $BATCH_SIZE > $DEST/stdout.log 
 

### 2022-03-01 scripts/run_train_clone.sh

@@ -35,8 +35,8 @@ LEARNING_RATE=0.00005
 
 CHECKPOINT_FREQ=1
 
-RUN_NAME="wgs_25m_w150_big"
-RUN_NOTES="150 width, more training data"
+RUN_NAME="wgs_25m_w150_big_cont"
+RUN_NOTES="150 width, more training data, continued"
 
 set -x
 
@@ -70,6 +70,7 @@ $PYTHON $ds2s train \
     --checkpoint-freq $CHECKPOINT_FREQ \
     -o ${RUN_NAME}.model \
     --threads 1 \
+    -i /uufs/chpc.utah.edu/common/home/arup-storage3/u0379426/variant_transformer_runs/wgs_25m_w150_big/wgs_25m_w150_big_epoch4.model \
     --max-decomp-batches 4 \
     --wandb-run-name $RUN_NAME \
     --wandb-notes "$RUN_NOTES"

### 2022-03-01 call.py
Fix typo

@@ -340,7 +340,7 @@ def _call_vars_region(aln, model, reference, chrom, start, end, max_read_depth, 
             encoded_with_ref = add_ref_bases(enc_reads, reference, chrom, window_start, window_start + window_size, max_read_depth=max_read_depth)
             batch.append(encoded_with_ref)
             batch_offsets.append(window_start)
-        except LowReadCountException:
+        except bam.LowReadCountException:
             logger.debug(
                 f"Bam window {chrom}:{window_start}-{window_start + window_size} "
                 f"had too few reads for variant calling (< {min_reads})"

### 2022-03-01 model.py

@@ -103,7 +103,7 @@ class VarTransformer(nn.Module):
         self.embed_dim = nhead * embed_dim_factor
         self.fc1_hidden = 12
 
-        self.fc1 = nn.Linear(feature_count, self.fc1_hidden)
+        self.fc1 = nn.Linear(feature_count +1, self.fc1_hidden)
         self.fc2 = nn.Linear(read_depth * self.fc1_hidden, self.embed_dim)
 
         # self.pos_encoder = PositionalEncoding(self.embed_dim, p_dropout)
@@ -116,8 +116,8 @@ class VarTransformer(nn.Module):
     def forward(self, src):
 
         # r contains a 1 if the read base matches the reference base, 0 otherwise
-        #r = (src[:, :, :, 0:4] * src[:, :, 0:1, 0:4]).sum(dim=-1) # x[:, :, 0:1..] is the reference seq
-        #src = torch.cat((src, r.unsqueeze(-1)), dim=3).to(self.device)
+        r = (src[:, :, :, 0:4] * src[:, :, 0:1, 0:4]).sum(dim=-1) # x[:, :, 0:1..] is the reference seq
+        src = torch.cat((src, r.unsqueeze(-1)), dim=3).to(self.device)
 
 
         src = self.elu(self.fc1(src))

### 2022-03-01 scripts/buildclf.py
Add classifier script

@@ -1,3 +1,5 @@
+#!/usr/bin/env python
+
 
 import sys
 import yaml
@@ -9,30 +11,35 @@ import sklearn
 from sklearn.ensemble import RandomForestClassifier
 
 import logging
+import argparse
 
 logger = logging.getLogger(__name__)
 logging.basicConfig(format='[%(asctime)s]  %(name)s  %(levelname)s  %(message)s',
                     datefmt='%m-%d %H:%M:%S',
                     level=logging.INFO)
 
+def var_feats(var):
+    feats = []
+    feats.append(var.qual)
+    feats.append(1 if "PASS" in var.filter else 0)
+    feats.append(1 if "LowCov" in var.filter else 0)
+    feats.append(1 if "SingleCallHet" in var.filter else 0)
+    feats.append(1 if "SingleCallHom" in var.filter else 0)
+    feats.append(min(var.info['QUALS']))
+    feats.append(max(var.info['QUALS']))
+    feats.append(var.info['WIN_VAR_COUNT'][0])
+    feats.append(var.info['WIN_CIS_COUNT'][0])
+    feats.append(var.info['WIN_TRANS_COUNT'][0])
+    feats.append(var.info['STEP_COUNT'][0])
+    feats.append(min(var.info['WIN_OFFSETS']))
+    feats.append(max(var.info['WIN_OFFSETS']))
+    feats.append(var.samples[0]['DP'])
+    return np.array(feats)
+
 def extract_feats(vcf):
     allfeats = []
     for var in pysam.VariantFile(vcf, ignore_truncation=True):
-        feats = []
-        feats.append(var.qual)
-        feats.append(1 if "PASS" in var.filter else 0)
-        feats.append(1 if "LowCov" in var.filter else 0)
-        feats.append(1 if "SingleCallHet" in var.filter else 0)
-        feats.append(1 if "SingleCallHom" in var.filter else 0)
-        feats.append(min(var.info['QUALS']))
-        feats.append(max(var.info['QUALS']))
-        feats.append(var.info['WIN_VAR_COUNT'][0])
-        feats.append(var.info['WIN_CIS_COUNT'][0])
-        feats.append(var.info['WIN_TRANS_COUNT'][0])
-        feats.append(var.info['STEP_COUNT'][0])
-        feats.append(min(var.info['WIN_OFFSETS']))
-        feats.append(max(var.info['WIN_OFFSETS']))
-        allfeats.append(np.array(feats))
[…]

### 2022-03-04 call.py
Add new feature to VCF output, cleanup imports

@@ -1,7 +1,6 @@
 import datetime
 import logging
 from collections import defaultdict
-import random
 
 import torch
 import pysam
@@ -125,10 +124,10 @@ def reconcile_current_window(prev_win, current_win):
 
 def load_model(model_path):
     logger.info(f"Loading model from path {model_path}")
-    attention_heads = 8
-    encoder_layers = 8
-    transformer_dim = 400
-    embed_dim_factor = 100
+    attention_heads = 4
+    encoder_layers = 6
+    transformer_dim = 200
+    embed_dim_factor = 125
     model = VarTransformer(read_depth=100,
                            feature_count=9,
                            out_dim=4,
@@ -341,7 +340,7 @@ def _call_vars_region(aln, model, reference, chrom, start, end, max_read_depth, 
             encoded_with_ref = add_ref_bases(enc_reads, reference, chrom, window_start, window_start + window_size, max_read_depth=max_read_depth)
             batch.append(encoded_with_ref)
             batch_offsets.append(window_start)
-        except LowReadCountException:
+        except bam.LowReadCountException:
             logger.debug(
                 f"Bam window {chrom}:{window_start}-{window_start + window_size} "
                 f"had too few reads for variant calling (< {min_reads})"

### 2022-03-04 loader.py

@@ -19,14 +19,11 @@ import io
 
 import scipy.stats as stats
 import numpy as np
-import pandas as pd
 import torch
 import torch.multiprocessing as mp
 import pysam
 
-import bwasim
 from bam import target_string_to_tensor, encode_with_ref, encode_and_downsample, ensure_dim
-import sim
 import util
 
 

### 2022-03-04 vcf.py

@@ -13,6 +13,7 @@ class Variant:
     hap_model: int = None
     step: int = None
     window_offset: int = None
+    var_index: int = None
 
     def __eq__(self, other):
         return self.ref == other.ref and self.alt == other.alt and self.pos == other.pos
@@ -52,6 +53,7 @@ class VcfVar:
     window_cis_vars: int
     window_trans_vars: int
     window_offset: int
+    var_index: int
     call_count: int
     step_count: int
     genotype: tuple
@@ -76,7 +78,11 @@ def _cigtups(cigstr):
 
 
 def _geomean(probs):
+    """
+    Geometric mean of numbers
+    """
     return np.exp(np.log(probs).mean())
+
 
 def align_sequences(query, target):
     """
@@ -89,6 +95,7 @@ def align_sequences(query, target):
                                                match_score=2,
                                                mismatch_score=-1)
     return aln
+
 
 def _mismatches_to_vars(query, target, offset, probs):
     """
@@ -129,6 +136,7 @@ def _mismatches_to_vars(query, target, offset, probs):
                       qual=_geomean(mismatch_quals),
                       window_offset=mismatchstart - offset)
 
+
 def aln_to_vars(refseq, altseq, offset=0, probs=None):
     """
     Smith-Watterman align the given sequences and return a generator over Variant objects
@@ -138,6 +146,8 @@ def aln_to_vars(refseq, altseq, offset=0, probs=None):
     :param offset: This amount will be added to each variant position
     :return: Generator over variants
     """
+
+    num_vars = 0
     if probs is not None:
         assert len(probs) == len(altseq), f"Probabilities must contain same number of elements as alt sequence"
     else:
@@ -148,10 +158,8 @@ def aln_to_vars(refseq, altseq, offset=0, probs=None):
 
     variant_pos_offset = 0
     if aln.query_begin > 0:
-        # yield Variant(ref='', alt=altseq[0:aln.query_begin], pos=offset, qual=-1)
         q_offset += aln.query_begin # Maybe we don't want this?
     if aln.target_begin > 0:
-        # yield Variant(ref=refseq[0:aln.target_begin], alt='', pos=offset, qual=-1)
         t_offset += aln.target_begin
 
     for cig in _cigtups(aln.cigar):
@@ -161,7 +169,9 @@ def aln_to_vars(refseq, altseq, offset=0, probs=None):
                     altseq[q_offset:q_offset+cig.len],
                     offset + t_offset,
[…]

### 2022-03-04 call.py
Clean up buildclf

@@ -123,10 +123,10 @@ def reconcile_current_window(prev_win, current_win):
 
 def load_model(model_path):
     logger.info(f"Loading model from path {model_path}")
-    attention_heads = 4
-    encoder_layers = 6
-    transformer_dim = 200
-    embed_dim_factor = 125
+    attention_heads = 8
+    encoder_layers = 8
+    transformer_dim = 400
+    embed_dim_factor = 100
     model = VarTransformer(read_depth=100,
                            feature_count=9,
                            out_dim=4,

### 2022-03-04 scripts/buildclf.py

@@ -71,13 +71,12 @@ def train_model(conf):
 def predict(model, vcf, **kwargs):
     model = load_model(model)
     vcf = pysam.VariantFile(vcf, ignore_truncation=True)
-    print(vcf.header)
+    print(vcf.header, end='')
     for var in vcf:
         feats = var_feats(var)
         prediction = model.predict_proba(feats[np.newaxis, ...])
-        print(prediction)
         var.qual = prediction[0, 1]
-        print(var)
+        print(var, end='')
 
 
 def train(conf_file, output, **kwargs):

### 2022-03-04 scripts/run_train_clone.sh

@@ -35,8 +35,8 @@ LEARNING_RATE=0.00005
 
 CHECKPOINT_FREQ=1
 
-RUN_NAME="wgs_25m_w150_bigger"
-RUN_NOTES="150 width, 'bigger' training data"
+RUN_NAME="wgs_25m_w150_bigger_cont"
+RUN_NOTES="150 width, 'bigger' training data, continued"
 
 set -x
 
@@ -70,6 +70,7 @@ $PYTHON $ds2s train \
     --checkpoint-freq $CHECKPOINT_FREQ \
     -o ${RUN_NAME}.model \
     --threads 1 \
+    -i /uufs/chpc.utah.edu/common/home/u0379426/storage/variant_transformer_runs/wgs_25m_w150_bigger/wgs_25m_w150_bigger_epoch2.model \
     --max-decomp-batches 4 \
     --wandb-run-name $RUN_NAME \
     --wandb-notes "$RUN_NOTES"

### 2022-03-04 vcf.py

@@ -428,6 +428,7 @@ def init_vcf(path, sample_name="sample", lowcov=30, cmdline=None):
     # write to new vcf file object
     return pysam.VariantFile(path, "w", header=vcfh)
 
+
 def prob_to_phred(p, max_qual=1000.0):
     """
     Convert a probability in 0..1 to phred scale
@@ -437,6 +438,7 @@ def prob_to_phred(p, max_qual=1000.0):
         return max_qual
     else:
         return min(max_qual, -10 * np.log10(1.0 - p))
+
 
 def create_vcf_rec(var, vcf_file):
     """
@@ -469,6 +471,7 @@ def create_vcf_rec(var, vcf_file):
     if var.duplicate:
         r.info['DUPLICATE'] = ()
     return r
+
 
 def vars_to_vcf(vcf_file, variants):
     """

### 2022-03-04 manuscript/manuscript.tex

@@
 
+Recent years have seen the transformer architecture, originally designed for language modelling, expand to multiple new areas, including vision (Dosovitskiy et al.
+2020, Liu et al.
+2021), video (Arnab et al.
+2021), and others.
+To our knowledge this is the first use of transformers for use of haplotype construction or variant detection in next generation sequencing data.
+
+Modern variant detection tools are extremely accurate, and generally result in very high sensitivity and specificity.
+
+ In his essay "The Bitter Lesson" Richard Sutton remarked that "methods that leverage general computation" often outperform methods with handcrafted statistical or heuristic techniques.
+Our work continues in this tradition, replacing the myriad finely tuned algorithms with a single model, learned end-to-end.
+
  \section{Availability}
@@
  \vspace{8pt}
+ Arnab, Anurag, et al. "Vivit: A video vision transformer." Proceedings of the IEEE/CVF International Conference on Computer Vision. 2021.
+
+ \vspace{8pt}
  Conrad, Donald F., et al. "Origins and functional impact of copy number variation in the human genome." Nature 464.7289 (2010): 704.
@@
 DePristo, Mark A., et al. "A framework for variation discovery and genotyping using next-generation DNA sequencing data." Nature genetics 43.5 (2011): 491-498.
+
+\vspace{8pt}
+Dosovitskiy, Alexey, et al. "An image is worth 16x16 words: Transformers for image recognition at scale." arXiv preprint arXiv:2010.11929 (2020).
 
@@
 Li, Heng, and Richard Durbin. "Fast and accurate short read alignment with Burrows–Wheeler transform." Bioinformatics 25.14 (2009): 1754-1760.
+
+\vspace{8pt}
+Liu, Ze, et al. "Swin transformer: Hierarchical vision transformer using shifted windows." Proceedings of the IEEE/CVF International Conference on Computer Vision. 2021.
 
@@
 \vspace{8pt}
- \subsection{Ablation experiments}
 Vaswani, Ashish, et al. "Attention is all you need." Advances in neural information processing systems 30 (2017).

### 2022-03-05 scripts/buildclf.py
Fix typo in buildclf

@@ -26,7 +26,7 @@ def var_feats(var):
     feats.append(1 if "SingleCallHet" in var.filter else 0)
     feats.append(1 if "SingleCallHom" in var.filter else 0)
     feats.append(len(var.ref))
-    feats.append(max(len(a) for a in var.alts)))
+    feats.append(max(len(a) for a in var.alts))
     feats.append(min(var.info['QUALS']))
     feats.append(max(var.info['QUALS']))
     feats.append(var.info['WIN_VAR_COUNT'][0])

### 2022-03-14 scripts/run_train_clone.sh
tr sc

@@ -35,8 +35,8 @@ LEARNING_RATE=0.00005
 
 CHECKPOINT_FREQ=1
 
-RUN_NAME="wgs_25m_w150_bigger_cont6"
-RUN_NOTES="150 width, 'bigger' training data, continued 6"
+RUN_NAME="wgs_25m_w150_bigger_cont7"
+RUN_NOTES="150 width, 'bigger' training data, continued 7"
 
 set -x
 
@@ -70,7 +70,7 @@ $PYTHON $ds2s train \
     --checkpoint-freq $CHECKPOINT_FREQ \
     -o ${RUN_NAME}.model \
     --threads 1 \
-    -i /uufs/chpc.utah.edu/common/home/u0379426/storage/variant_transformer_runs/wgs_25m_w150_bigger_cont5/wgs_25m_w150_bigger_cont5_epoch1.model \
+    -i /uufs/chpc.utah.edu/common/home/u0379426/storage/variant_transformer_runs/wgs_25m_w150_bigger_cont6/wgs_25m_w150_bigger_cont6_epoch1.model \
     --max-decomp-batches 4 \
     --wandb-run-name $RUN_NAME \
     --wandb-notes "$RUN_NOTES"

### 2022-03-14 train.py
50m params

@@ -329,17 +329,10 @@ def train_epochs(epochs,
                  wandb_notes="",
                  cl_args = {}
 ):
-<<<<<<< HEAD
     attention_heads = 10
-    transformer_dim = 400
+    transformer_dim = 512
     encoder_layers = 10
     embed_dim_factor = 100
-=======
-    attention_heads = 4
-    transformer_dim = 200
-    encoder_layers = 6
-    embed_dim_factor = 125
->>>>>>> master
     model = VarTransformer(read_depth=max_read_depth,
                             feature_count=feats_per_read, 
                             out_dim=4,

### 2022-03-15 manuscript/manuscript.tex

@@
 Here we describe a new approach that replaces the handcrafted statistical methods with a single, end-to-end deep learning model.
-Our model frames variant detection as a sequence-to-sequence modeling task, and employs a transformer-based architecture to translate alignment columns into predicted haplotype sequences.
+Our model frames variant detection as a sequence-to-sequence modeling task, and employs a transformer-based architecture to translate alignment columns into two predicted haplotype sequences.
+Trained on whole genome sequences from Genome-in-a-Bottle, we demonstrate that this approach yields sensitivity and precision similar to other top-performing callers, and that
 \end{abstract}
@@
 
-Training tensors were generated by selecting regions 150bp in width from alignments.
-The encoding process included fetching all reads aligning to the desired region.
-If more reads were found then a specified maximum read depth (100 for all analyses included in this work), reads were randomly downsampled to not exceed the maximum size.
-Reads were sorted by the reference coordinate of the first aligned base.
-For each aligned base in the selected reads nine features were encoded; the first four were the one-hot encoded base call, followed by base quality, two flags indicating if base `consumed' reference base (i.e.
+Training tensors were generated by selecting 150bp genomic regions in a biased manner described below.
+For each region, we fetch all overlapping reads and sort them by first aligned reference position.
+If more than 100 overlapping reads were found we randomly downsampled the set to not exceed the maximum size.
+For each base in the selected reads nine features were encoded; the first four were the one-hot encoded base call, followed by base quality, two flags indicating if base `consumed' reference base (i.e.
 was not an insertion) or consumed a sequence read base (was not a deletion), and additional flags indicating sequence read direction and clipping status.
-No 'gap' tokens or other special handling was performed for insertions or deletions.
+No gap tokens or other special handling was performed for insertions or deletions.
 In addition to sequenced reads, the first row in each encoded region was the reference sequence.
@@
 
+ Things to look at:
+ Difficult regions
+ Complex variants
+ phasing
+
  \section{Discussion}
 
- Convolutions - do they make sense?
+We describe a new approach to the problem of detecting sequence variants in next-generation sequencing (NGS) data based on a single, end-to-end deep learning model.
+Our approach envisions variant detection as a sequence-to-sequence modeling problem, akin to language translation, and leverages the successful transformer architecture to accomplish the task.
+In contrast to other variant detection methods (...), our approach does not rely on handcrafted statistical techniques such as de Bruijn graphs, hidden markov models, or heuristic thresholding and instead learns to construct accurate haplotypes directly from aligned NGS reads.
+
+Our tools moves the deep learning part from just classification (real or false) of a previously detected variant to actual construction of the variants themselves.
+
+Previous uses of deep learning for variant detection have employed Convolutional Neural Networks (CNNs).
+While CNNs have been used successfully in computer vision tasks, genomic data are fundamentally different.
+Specifically, nearby pixels in an image are much more likely to be part of the same structure than pixels that are more distant, and CNNs exploit this implicit structure by repeatedly applying convolutions to a small patch of pixels.
+In contrast, reads aligning to a genomic region do not share this structure.
+Two adjacent "pixels" (base calls) that align to the same genomic coordinate from different reads are just as likely to be from the same structure as any two randomly selected pixels that share the same alignment coordinate.
+In other words, permuting the order of reads in a genomic pileup does not impact the information content, but permuting rows of an image would likely degrade
 
@@
 
+Since the introduction of the first Transformer model (Vaswani et al. 2017) many improvements have been proposed, both in the general case and for vision-specific tasks
+
+Lots of transformer variants out there - swin transformers, fastformer, hierarchical blah blah... good to mention these as future directions.
+
 Modern variant detection tools are extremely accurate, and generally result in very high sensitivity and specificity.
@@
  In his essay "The Bitter Lesson" Richard Sutton remarked that "methods that leverage general computation" often outperform methods with handcrafted statistical or heuristic techniques.
-Our work continues in this tradition, replacing the myriad finely tuned algorithms with a single model, learned end-to-end.
+Our work continues in this tradition, replacing the myriad finely tuned algorithms with a single, fully learned model.
 

### 2022-04-29 manuscript/manuscript.tex
manuscript updates

--- 2022-02-28 manuscript/manuscript.tex
@@
 Accurate reconstruction of the DNA sequence in a sample is a fundamental component of modern genomics analysis.
-Next Generation Sequencing (NGS) workflows typically begin by aligning sequenced fragments to a reference genome, optionally performing some alignment post-processing steps such as identification of duplicate fragments, and then identifying sequence variants relative to the reference genome.
+Next Generation Sequencing (NGS) workflows typically begin by aligning sequenced fragments to a reference genome, optionally performing some alignment post-processing steps such as identification of duplicated fragments, and then identifying sequence variants relative to the reference genome.
 The variant identification step is challenging for many reasons, including technical artifacts in the sequencing process, improper alignments due to sequence homology or low sequence complexity, and the stochastic nature of the library preparation and amplification steps [citation here would be nice].
@@
 
-Our model consists of a transformer-based encoder and simple, fully connected decoder.
+Our model consists of a transformer-based encoder and simple, fully connected decoder that produces two output sequences, one for each predicted haplotype.
 The encoder component is an unmodified transformer with GeLU activations as implemented in PyTorch 1.10 (Paszke et al.
 2019).
+Input tokens are the collection of bases that align to a given reference position across reads, with some modifications described below.
 We add an additional fully connected layer prior to the transformer encoders which embeds the encoded basecalls in $d$ dimensions, where $d=12$ for the analyses here.
-The embedded basecalls are 'flattened' along the read dimension, producing an input with size $dr$ where $r$ is the number of reads.
-Instead of the typical 1-dimensional positional encoding, we employed a 2-dimensional encoding, enabling the model to retain information about which read contained a given sequenced base.
+The embedded basecalls are 'flattened' along the read dimension, producing an input token with size $dr$ where $r$ is the number of reads.
+Instead of the typical 1-dimensional positional encoding, we employed a 2-dimensional encoding, enabling the model to retain information about which read contained a given base across input tokens.
 
@@
 We obtained training data from 17 whole genomes sequenced from 5 Genome-in-a-Bottle cell lines.
-Eight samples were prepped with Illumina Nextera DNA Flex kit and the remained were prepared with the Illumina TruSeq PCR-free kit.
+Eight samples were prepared with Illumina Nextera DNA Flex kit and the remainder were prepared with the Illumina TruSeq PCR-free kit.
 All samples were sequenced on an Illumina NovaSeq 6000 instrument in 2x150 mode, to an approximate read depth of 50.
@@
 
-Training tensors were generated by selecting regions 150bp in width from alignments.
-The encoding process included fetching all reads aligning to the desired region.
-If more reads were found then a specified maximum read depth (100 for all analyses included in this work), reads were randomly downsampled to not exceed the maximum size.
+Training tensors were generated by selecting regions 150bp in width from a list of preselected regions, in a manner described below.
+For each input region, all reads overlapping the window were obtained from the corresponding BAM/ CRAM file.
+If more than a specified maximum number of reads were found (100 for all analyses here), then reads were downsampled to the desired size.
 Reads were sorted by the reference coordinate of the first aligned base.
@@
 
+\subsection{Loss function}
+
+We use a familiar cross-entropy loss function to compare predicted haplotypes to known haplotype sequences for each region, and simply compare each prediction at window index (position) $i$ to the true base at index $i$.
+While we obtain satisfactory results with this approach, two issues were apparent.
+First, predicted haplotype sequences may have inserted or deleted bases relative to the true haplotypes.
+Because incorrect insertions and deletions cause mismatches at every downstream (higher index) position, such mistakes are penalized more heavily when they occur at lower indexes than higher indexes, because a greater number of mismatches are generated.
+Thus the model learns to generate high quality results for the first portion of the window where inaccuracies are punished severely, and is less concerned with accuarcy at high indexes.
+
+A second complexity arises due to ambiguity regarding which haplotype should contain a variant.
+The two haplotypes present in a sample have no inherent order.
+In training data variants are assigned arbitrarily to haplotypes, and this randomness in haplotype assignment means model cannot learn to predict which haplotype should contain a given variant.
+To combat this ambiguity, we calculate loss in both possible configurations - predicted haplotype 0 with training haplotype 0 and 1 with 1, as well as predicted haplotype 0 with training haplotype 1 and 1 with 0.
+We select the configuration with the lowest loss and backpropogate gradients only for that configuration.
+
 \subsection{Variant detection}
@@
 \section{Results}
+
+\subsection{Yes we can really do nontrivial / complex alignments}
+
+Just show some examples here
 
@@
 
+ Ablate reference as read 0
+
+ \subsection{Phasing accuracy}
+
  \section{Discussion}
@@
  Convolutions - do they make sense?
+
+ Loss functions, there's probably something better. The Smith Waterman loss sounds really cool, but didn't lead to high sensitivity in our hands.
 

### 2022-04-30 dnaseq2seq/buildclf.py
try adding gnomad variant normalization stuff

@@ -20,16 +20,34 @@ logging.basicConfig(format='[%(asctime)s]  %(name)s  %(levelname)s  %(message)s'
                     level=logging.INFO)
 
 
+
+def trim_prefix(ref, alt):
+    offset = 0
+    if len(ref) == 1 or len(alt) == 1:
+        return 0, ref, alt
+    for r, a in zip(ref[:-1], alt[:-1]):
+        if r != a:
+            return offset, ref[offset:], alt[offset:]
+        offset += 1
+    return offset, ref[offset:], alt[offset:]
+
+
 def find_var(varfile, chrom, pos, ref, alt):
     """
     Search varfile for VCF record with matching chrom, pos, ref, alt
     If chrom/pos/ref/alt match is found, return that record and allele index of matching alt
     """
-    for var in varfile.fetch(chrom, pos-3, pos+3):
-        if var.ref == ref and var.pos == pos:
-            for i, varalt in enumerate(var.alts):
-                if varalt == alt:
-                    return var, i
+    poffset, trimref, trimalt = trim_prefix(ref, alt)
+    print(f"Trimmed query: {pos + poffset} {trimref}  {trimalt}")
+    pos = pos + poffset
+    for var in varfile.fetch(chrom, pos-5, pos+3):
+        print(f"Comparing to : {var}")
+        for i, varalt in enumerate(var.alts):
+            offset, r, a = trim_prefix(var.ref, varalt)
+            print(f"Comparing to {var.pos + offset}  {r}  {a}")
+            if var.pos + offset == pos and r == trimref and a == trimalt:
+                return var, i
+
 
 
 @lru_cache(maxsize=1000000)
@@ -208,5 +226,7 @@ def main():
 
 
 if __name__ == "__main__":
-    main()
+    varfile = pysam.VariantFile("test.vcf.gz")
+    print(find_var(varfile, "2", 99, "CTTG", "GTA"))
+    #main()
 

### 2022-05-01 manuscript/manuscript.tex

--- 2022-03-15 manuscript/manuscript.tex
@@
 Our model frames variant detection as a sequence-to-sequence modeling task, and employs a transformer-based architecture to translate alignment columns into two predicted haplotype sequences.
-Trained on whole genome sequences from Genome-in-a-Bottle, we demonstrate that this approach yields sensitivity and precision similar to other top-performing callers, and that
+We demonstrate that this method  predicts variant genotypes accurately, phases nearby variants correctly, reconstructs complex variants, and incorporates local sequence context and variation into prediction confidence.
+Trained on whole genome sequences from Genome-in-a-Bottle, we demonstrate that this approach yields slightly lower sensitivity and precision than current top-performing callers,
 \end{abstract}
@@
 2011) examined each alignment column individually to determine genotype.
-Because these tools do not realign reads, they rely on the alignments produced by read mapping tools such as BWA (Li \& Durbin 2009) .
+Because these tools do not realign reads, they rely on the alignments produced by read mapping tools such as BWA (Li \& Durbin 2009).
 Later tools, such as the HaplotypeCaller from the GATK and Strelka 2 (Kim et al.
-2018) incorporated a re-alignment of reads in each region of interest, resulting in a significant improvement in the precision and specificity of variant calls.
+2018) incorporated local re-alignment of reads in each region of interest, resulting in a significant improvement in the precision and specificity of variant calls.
 The HaplotypeCaller tool constructs de Bruijn graphs from k-mers present in the reads in a region, realigns reads to the graphs, and uses a pair Hidden Markov Model (HMM) to determine the most likely path through the graph.
@@
 For each region, we fetch all overlapping reads and sort them by first aligned reference position.
-If more than 100 overlapping reads were found we randomly downsampled the set to not exceed the maximum size.
+If more than 100 overlapping reads were found we randomly downsample the set to not exceed the maximum size.
 For each base in the selected reads nine features were encoded; the first four were the one-hot encoded base call, followed by base quality, two flags indicating if base `consumed' reference base (i.e.
@@
 
-Target / label sequences were produced by obtaining truth variants from the Genome-in-a-Bottle VCF files for each sample and inserting the variants into the reference sequence.
+Target / label haplotype sequences were produced by obtaining truth variants from the Genome-in-a-Bottle VCF files for each sample and inserting the variants into the reference sequence.
 Two sequences were generated for each region, one representing each haplotype.
-In situations where phasing of the variants was ambiguous, the reads in the sequenced sample were examined to determine phase status.
+In regions where phasing of the variants was ambiguous, the reads in the sequenced sample were examined to determine phase status.
 Briefly, all possible genotypes (pairs of haplotype sequences) were generated and reads were aligned via Smith-Waterman to the possible haplotypes, and the highest scoring genotypes selected as the most likely phasing.
-This phasing procedure was only performed if multiple heterozygous variants were present in the region.
+This phasing procedure procedure was only attempted for variants less than 100bp apart, otherwise the region was discarded.
 
@@
 
-Given an alignment file in BAM or CRAM format,
+Given an alignment file in BAM or CRAM format, we first identify regions where a potential variant might exist, and then run forward passes of the model using a sliding window over the region.
+Any genomic position in which at least two reads contain a base that differs from the reference or an indel are flagged as potentially containing a variant.
+Positions closer than 100bp are merged into a single region.
+
+Given a region containing suspected variants, we perform multiple overlapping forward passes of the model with step size $k$.
+On each forward pass the model produces two predicted haplotypes.
+Each haplotype is aligned via Smith-Waterman to the reference sequence, and any mismatching positions are converted to variant calls.
+We record the number of windows in which each variant was detected
 
@@
 Our approach envisions variant detection as a sequence-to-sequence modeling problem, akin to language translation, and leverages the successful transformer architecture to accomplish the task.
-In contrast to other variant detection methods (...), our approach does not rely on handcrafted statistical techniques such as de Bruijn graphs, hidden markov models, or heuristic thresholding and instead learns to construct accurate haplotypes directly from aligned NGS reads.
+In contrast to other variant detection methods (...), our approach does not rely on handcrafted statistical techniques such as de Bruijn graphs, hidden Markov models, or heuristic thresholding and instead learns to construct accurate haplotypes directly from aligned NGS reads.
 

### 2022-05-01 dnaseq2seq/buildclf.py
10m params for calling, add gnomad prefix trimming for gnomad feature

@@ -20,6 +20,14 @@ logging.basicConfig(format='[%(asctime)s]  %(name)s  %(levelname)s  %(message)s'
                     level=logging.INFO)
 
 
+def trim_suffix(ref, alt):
+    r = ref
+    a = alt
+    while r[-1] == a[-1] and min(len(r), len(a)) > 1:
+        r = r[0:-1]
+        a = a[0:-1]
+    return r, a
+
 
 def trim_prefix(ref, alt):
     offset = 0
@@ -37,14 +45,16 @@ def find_var(varfile, chrom, pos, ref, alt):
     Search varfile for VCF record with matching chrom, pos, ref, alt
     If chrom/pos/ref/alt match is found, return that record and allele index of matching alt
     """
+    ref, alt = trim_suffix(ref, alt)
     poffset, trimref, trimalt = trim_prefix(ref, alt)
-    print(f"Trimmed query: {pos + poffset} {trimref}  {trimalt}")
+    #print(f"Trimmed query: {pos + poffset} {trimref}  {trimalt}")
     pos = pos + poffset
     for var in varfile.fetch(chrom, pos-5, pos+3):
-        print(f"Comparing to : {var}")
+        #print(f"Comparing to : {var}")
         for i, varalt in enumerate(var.alts):
-            offset, r, a = trim_prefix(var.ref, varalt)
-            print(f"Comparing to {var.pos + offset}  {r}  {a}")
+            varref, varalt = trim_suffix(var.ref, varalt)
+            offset, r, a = trim_prefix(varref, varalt)
+            #print(f"Comparing to {var.pos + offset}  {r}  {a}")
             if var.pos + offset == pos and r == trimref and a == trimalt:
                 return var, i
 
@@ -53,11 +63,13 @@ def find_var(varfile, chrom, pos, ref, alt):
 @lru_cache(maxsize=1000000)
 def var_af(varfile, chrom, pos, ref, alt):
     result = find_var(varfile, chrom, pos, ref, alt)
+    print(f"Find var result: {result}")
     af = 0.0
     if result is not None:
         var, alt_index = result
         af = var.info['AF'][alt_index]
-        if af is None:
+        print(f"Got AF: {af}")
+        if af is None or "PASS" not in var.filter:
             af = 0.0
         else:
             af = float(af)
@@ -82,6 +94,7 @@ def var_feats(var, var_freq_file):
[…]

### 2022-05-01 dnaseq2seq/call.py

@@ -138,9 +138,9 @@ def reconcile_current_window(prev_win, current_win):
 
 
 def load_model(model_path):
-    attention_heads = 8
-    encoder_layers = 8
-    transformer_dim = 400
+    attention_heads = 6
+    encoder_layers = 6
+    transformer_dim = 250
     embed_dim_factor = 100
     model = VarTransformer(read_depth=100,
                            feature_count=9,

### 2022-05-02 scripts/complexvarfinder.py
Fix complexvarfinder

@@ -36,7 +36,7 @@ def main(vcf, bam, refgenome):
         if len(var.ref) > min_size or any(len(a) > min_size for a in var.alts):
             indel_starts = [s for s in find_indel_starts(aln, var.chrom, var.pos - 20, var.pos + 20)]
             if len(indel_starts) > 3:
-                print(var)
+                print(var, end='')
 
 
 if __name__=="__main__":

### 2022-05-02 manuscript/manuscript.tex

--- 2022-04-29 manuscript/manuscript.tex
@@
 Here we describe a new approach that replaces the handcrafted statistical methods with a single, end-to-end deep learning model.
-Our model frames variant detection as a sequence-to-sequence modeling task, and employs a transformer-based architecture to translate alignment columns into predicted haplotype sequences.
+Our model frames variant detection as a sequence-to-sequence modeling task, and employs a transformer-based architecture to translate alignment columns into two predicted haplotype sequences.
+We demonstrate that this method predicts variant genotypes accurately, phases nearby variants correctly, and has the ability to realign complex and ambiguous read mappings to produce accurate haplotype predictions.
 \end{abstract}
@@
 
-We use a familiar cross-entropy loss function to compare predicted haplotypes to known haplotype sequences for each region, and simply compare each prediction at window index (position) $i$ to the true base at index $i$.
+We use a standard cross-entropy loss function to compare predicted haplotypes to known haplotype sequences for each region, and simply compare each prediction at window index (position) $i$ to the true base at index $i$.
 While we obtain satisfactory results with this approach, two issues were apparent.
 First, predicted haplotype sequences may have inserted or deleted bases relative to the true haplotypes.
-Because incorrect insertions and deletions cause mismatches at every downstream (higher index) position, such mistakes are penalized more heavily when they occur at lower indexes than higher indexes, because a greater number of mismatches are generated.
+Because incorrect insertions and deletions cause mismatches at every downstream (higher index) position, such mistakes are penalized more heavily when they occur at lower indexes than higher indexes.
 Thus the model learns to generate high quality results for the first portion of the window where inaccuracies are punished severely, and is less concerned with accuarcy at high indexes.
@@
 
-\subsection{Yes we can really do nontrivial / complex alignments}
+\subsection{Genotyping}
+ - Are we getting the right variant genotypes
 
-Just show some examples here
+\subsection{Phasing accuracy}
+ - look at nearby variants, see if they are phased correctly compared to GIAB information
+
+\subsection{Complex regions}
+ - Not sure how to define these, but ideally we'd look for variants in regions where the read mapper had
+ lots of overlapping insertion / deletions in reads and/or softclipping
+
+\subsection{Use of local context}
+ - Also ambiguous, but anecdotally when looking at variant calls, it seems apparent that variants that otherwise look real have lower quality scores when there's a lot of junk around.
+That's good, since such variants are more likely To
+ be false positives, and that means the model is to some extent looking at local context to modify prediction accuracy. Which is cool.
+
+\subsection{Variant detection accuracy}
+
+ - Tabulate Hap.py results across a few common stratifications, compare to other callers
 

### 2022-05-09 scripts/complexvarfinder.py
Update complexvarfinder script, pregen script

@@ -9,6 +9,13 @@ Find indels greater than 10bp where the aligned reads do not agree on the positi
 
 REFGENOME="/uufs/chpc.utah.edu/common/home/u0379426/arup-storage3/Reference/Data/B37/GATKBundle/2.8/human_g1k_v37_decoy_phiXAdaptr.fasta"
 
+
+def softclipped_bases(read):
+    if read.cigartuples is None:
+        return 0
+    return sum(c for op, c in read.cigartuples if op in {4,5})
+
+
 def find_indel_starts(aln, chrom, start, stop):
     for col in aln.pileup(chrom, start=start, stop=stop, stepper='nofilter'):
         # The pileup returned by pysam actually starts long before the first start position, but we only want to
@@ -24,6 +31,16 @@ def find_indel_starts(aln, chrom, start, stop):
                     yield col.reference_pos
                     break
 
+def count_softclipped_bases(aln, chrom, start, stop):
+    tot_reads = 0
+    reads_with_big_softclips = 0
+    for read in aln.fetch(chrom, start, stop):
+        tot_reads += 1
+        sc = softclipped_bases(read)
+        if sc > 10:
+            reads_with_big_softclips += 1
+    return reads_with_big_softclips, tot_reads
+
 
 def main(vcf, bam):
     min_size = 4
@@ -33,8 +50,9 @@ def main(vcf, bam):
     print(vcf.header, end='')
     for var in vcf:
         if len(var.ref) > min_size or any(len(a) > min_size for a in var.alts):
+            bigsoftclips, totreads = count_softclipped_bases(aln, var.chrom, var.pos - 20, var.pos + 20)
             indel_starts = [s for s in find_indel_starts(aln, var.chrom, var.pos - 20, var.pos + 20)]
-            if len(indel_starts) > 3:
+            if len(indel_starts) > 4 or bigsoftclips / totreads > 0.20:
                 print(var, end='')
 
 

### 2022-05-09 scripts/run_pregen.sh

@@ -18,11 +18,11 @@ ds2s=/uufs/chpc.utah.edu/common/home/u0379426/src/dnaseq2seq/dnaseq2seq/bin/main
 #CONF=/uufs/chpc.utah.edu/common/home/u0379426/src/dnaseq2seq/multindel_novaav_conf.yaml
 CONF=/uufs/chpc.utah.edu/common/home/u0379426/src/dnaseq2seq/wgs_multindel_conf.yaml
 
-DEST=/uufs/chpc.utah.edu/common/home/arup-storage3/u0379426/wgs_pregen_huge
+DEST=/uufs/chpc.utah.edu/common/home/arup-storage3/u0379426/wgs_pregen_no_lc
 
-BATCH_SIZE=512
+BATCH_SIZE=256
 
 mkdir -p $DEST
 
-$PYTHON $ds2s pregen -c $CONF -d $DEST --threads 48 --batch-size $BATCH_SIZE > $DEST/stdout.log 
+$PYTHON $ds2s pregen -c $CONF -d $DEST --threads 24 --batch-size $BATCH_SIZE > $DEST/stdout.log 
 

### 2022-05-10 scripts/run_pregen.sh
pregen scr

@@ -18,7 +18,7 @@ ds2s=/uufs/chpc.utah.edu/common/home/u0379426/src/dnaseq2seq/dnaseq2seq/bin/main
 #CONF=/uufs/chpc.utah.edu/common/home/u0379426/src/dnaseq2seq/multindel_novaav_conf.yaml
 CONF=/uufs/chpc.utah.edu/common/home/u0379426/src/dnaseq2seq/wgs_multindel_conf.yaml
 
-DEST=/uufs/chpc.utah.edu/common/home/arup-storage3/u0379426/wgs_pregen_no_lc
+DEST=/uufs/chpc.utah.edu/common/home/arup-storage3/u0379426/wgs_pregen_snvs
 
 BATCH_SIZE=256
 

### 2022-05-10 manuscript/manuscript.tex

@@
 We demonstrate that this method predicts variant genotypes accurately, phases nearby variants correctly, and has the ability to realign complex and ambiguous read mappings to produce accurate haplotype predictions.
+
 \end{abstract}
@@
 Training tensors were generated by selecting regions 150bp in width from a list of preselected regions, in a manner described below.
-For each input region, all reads overlapping the window were obtained from the corresponding BAM/ CRAM file.
+For each region, all reads overlapping the window were obtained from the corresponding BAM/ CRAM file.
 If more than a specified maximum number of reads were found (100 for all analyses here), then reads were downsampled to the desired size.
@@
  \section{Discussion}
+
+We describe a new approach to the problem of detecting sequence variants in next-generation sequencing (NGS) data based on a single, end-to-end deep learning model.
+Our approach envisions variant detection as a sequence-to-sequence modeling problem, akin to language translation, and leverages the successful transformer architecture to accomplish the task.
+In contrast to other variant detection methods (...), our approach does not rely on handcrafted statistical techniques such as de Bruijn graphs, hidden Markov models, or heuristic thresholding and instead learns to construct accurate haplotypes directly from aligned NGS reads.
 

### 2022-05-10 dnaseq2seq/loader.py
randomize pregenloader

@@ -154,6 +154,7 @@ class PregenLoader:
             self.pathpairs = pathpairs
         else:
             self.pathpairs = util.find_files(self.datadir, self.src_prefix, self.tgt_prefix, self.vaftgt_prefix)
+            self.pathpairs = random.shuffle(self.pathpairs)
         self.threads = threads
         self.max_decomped = max_decomped_batches # Max number of decompressed items to store at once - increasing this uses more memory, but allows increased parallelization
         logger.info(f"Creating PreGen data loader with {self.threads} threads")

### 2022-05-10 scripts/assignclasses.py

@@ -42,7 +42,7 @@ vcf = pysam.VariantFile(sys.argv[1])
 #sys.stderr.write("Loading forest from " + sys.argv[2])
 #forest = buildforest(sys.argv[2])
 
-for line in open(sys.argv[3]):
+for line in open(sys.argv[2]):
     toks = line.split("\t")
     start = int(toks[1])
     end = int(toks[2])
@@ -72,8 +72,6 @@ for line in open(sys.argv[3]):
         label = "ins"
     elif snv_count:
         label = "snv"
-    elif interval_count:
-        label = "tn-flag"
     else:
         label = "tn"
 

### 2022-05-10 scripts/run_pregen.sh

@@ -17,8 +17,9 @@ ds2s=/uufs/chpc.utah.edu/common/home/u0379426/src/dnaseq2seq/dnaseq2seq/bin/main
 
 #CONF=/uufs/chpc.utah.edu/common/home/u0379426/src/dnaseq2seq/multindel_novaav_conf.yaml
 CONF=/uufs/chpc.utah.edu/common/home/u0379426/src/dnaseq2seq/wgs_multindel_conf.yaml
+#CONF=/uufs/chpc.utah.edu/common/home/u0379426/src/dnaseq2seq/wgs_bigvars_conf.yaml
 
-DEST=/uufs/chpc.utah.edu/common/home/arup-storage3/u0379426/wgs_pregen_snvs
+DEST=/uufs/chpc.utah.edu/common/home/arup-storage3/u0379426/wgs_pregen_bigvars
 
 BATCH_SIZE=256
 

### 2022-05-12 manuscript/manuscript.tex
manuscript changes

@@
 
-Variant discovery tools use a variety of statistical techniques to identify variants.
+All variant discovery tools must address two challenges.
+First, they must identify candidate alleles in a region.
+Then, given a set of allele candidates, they must assess the likelihood of each candidate (or pairs of candidates in the diploid case), to determine the most likely alleles.
 Early callers, such as samtools / mpileup (Li et al.
 2009) and the UnifiedGenotyper tool from the Genome Analysis ToolKit (GATK, De Pristo et al.
-2011) examined each alignment column individually to determine genotype.
-Because these tools do not realign reads, they rely on the alignments produced by read mapping tools such as BWA (Li \& Durbin 2009).
-Later tools, such as the HaplotypeCaller from the GATK and Strelka 2 (Kim et al.
-2018) incorporated local re-alignment of reads in each region of interest, resulting in a significant improvement in the precision and specificity of variant calls.
-The HaplotypeCaller tool constructs de Bruijn graphs from k-mers present in the reads in a region, realigns reads to the graphs, and uses a pair Hidden Markov Model (HMM) to determine the most likely path through the graph.
+2011) rely on the read aligner to generate candidate allele candidates, and utilize ad-hoc heuristics and thresholds to determine likely alleles.
+Later tools (e.g.
+Poplin et al.
+2018, Kim et al.
+2018, Cooke et al.
+2021) incorporated local re-alignment of reads in each region of interest, resulting in a significant improvement in the precision and specificity of variant calls.
+For instance, the HaplotypeCaller tool identifies candidate haplotypes by constructing de Bruijn graphs from k-mers present in the reads, and assesses haplotype likelihoods with a pair Hidden Markov Model (HMM).
 
-More recent tools have incorporated elements of deep learning to further improve accuracy.
+More recent tools have incorporated elements of deep learning into the allele likelihood calculation.
 DeepVariant (Poplin et al.
@@
 2020), designed to work on hybrid short- and long-read datasets, employs a mixture-of-experts approach with separate 1-dimensional convolutions across the read and allele dimensions of the input.
-While these tools deliver accurate results, they rely heavily on hand-crafted statistical algorithms and heuristics to generate and assess candidate alleles.
 
-Here, we show that none of the bespoke statistical apparatus developed in earlier tools is required for accurate variant identification, and that a robust variant detection model can be learned directly from raw NGS data.
-To achieve this, we recast variant detection as a sequence-to-sequence translation problem.
-The series of bases aligning to a given reference position is a single input token, and the sequence of these alignment columns constitutes the input to the model.
-The output of the model is the two predicted haplotype sequences in the sample.
+Here, we explore a new approach to variant detection that replaces the bespoke statistical techniques used in previous tools with a single, end-to-end machine learning model.
+We recast variant detection as a sequence-to-sequence translation problem in which the input sequence is the series of bases aligning to a single genomic location, and the output sequences are the two predicted haplotypes.
 The sequence-to-sequence translation is performed by a transformer-based deep neural network (Vaswani et al.
 2017).
-Transformer models have achieved state-of-the-art results in many sequence-to-sequence problems,
+We demonstrate that this approach identifies real genomic variants with sensitivity equivalent to current top-performing callers, albeit with slightly lower specificity.
 
@@
 
+For all analyses described here, we obtain training data only from the human autosomes 1-20, and hold out chromosomes 21 and 22 for model evaluation.
+
 \subsection{Loss function}
@@
 Given an alignment file in BAM or CRAM format, we first identify regions where a potential variant might exist, and then run forward passes of the model using a sliding window over the region.
-Any genomic position in which at least two reads contain a base that differs from the reference or an indel are flagged as potentially containing a variant.
+Any genomic position in which at least three reads contain a base that differs from the reference or an indel are flagged as potentially containing a variant.
 Positions closer than 100bp are merged into a single region.
@@
 \subsection{Complex regions}
- - Not sure how to define these, but ideally we'd look for variants in regions where the read mapper had
- lots of overlapping insertion / deletions in reads and/or softclipping
+ - Not sure how to define these, but ideally we'd look for variants in regions where the read mapper had lots of overlapping insertion / deletions in reads and/or softclipping
 
@@
 \vspace{8pt}
+Cooke, Daniel P., David C. Wedge, and Gerton Lunter. "A unified haplotype-based method for accurate and comprehensive variant calling." Nature biotechnology 39.7 (2021): 885-892.
+
+\vspace{8pt}
 DePristo, Mark A., et al. "A framework for variation discovery and genotyping using next-generation DNA sequencing data." Nature genetics 43.5 (2011): 491-498.
@@
 \vspace{8pt}
-Poplin, Ryan, et al. "A universal SNP and small-indel variant caller using deep neural networks." Nature biotechnology 36.10 (2018): 983-987.
-
-\vspace{8pt}
 Paszke, A., et al.
@@
 \vspace{8pt}
+Poplin, Ryan, et al. "A universal SNP and small-indel variant caller using deep neural networks." Nature biotechnology 36.10 (2018): 983-987.
+
+\vspace{8pt}
+Poplin, Ryan, et al. "Scaling accurate genetic variant discovery to tens of thousands of samples." BioRxiv (2018): 201178.
+
+\vspace{8pt}
 Ramachandran, Anand, et al. "HELLO: A hybrid variant calling approach." bioRxiv (2020).

### 2022-05-20 manuscript/manuscript.tex
Various manuscript updates

@@
 2020), designed to work on hybrid short- and long-read datasets, employs a mixture-of-experts approach with separate 1-dimensional convolutions across the read and allele dimensions of the input.
+
+Maybe brief explanation of how transformers work?
 
@@
 Given a region containing suspected variants, we perform multiple overlapping forward passes of the model with step size $k$.
-On each forward pass the model produces two predicted haplotypes.
+On each forward pass the model produces two predicted haplotypes by taking the base with the high predicted probability at each position.
 Each haplotype is aligned via Smith-Waterman to the reference sequence, and any mismatching positions are converted to variant calls.
-We record the number of windows in which each variant was detected
+For each variant we record the number of windows in which each variant was detected, the number of variants in cis and trans, and the mean probability of the variant bases, and the position of the variant within each window.
+
+\subsection{Variant quality calculation and false positive removal}
+
+The above variant detection procedure has two shortcomings.
+First, if every detected variant is emitted, then specificity is poor.
+Second, variant quality scores are poorly calibrated and do not closely reflect the probability that a variant is real.
+To address these shortcomings we introduce a post-hoc random forest classifier that we train to discriminate true and false-positive variant calls (see table X for features).
+The resulting score is used as a final variant quality score that can be used for tuning the tradeoff between sensitivity and specificity.
 
@@
 
-\subsection{Genotyping}
- - Are we getting the right variant genotypes
+\subsection{Model size and training data experiments}
 
-\subsection{Phasing accuracy}
- - look at nearby variants, see if they are phased correctly compared to GIAB information
+We explore the effects of model size and training data size on validation loss and variant detection accuracy. We evaluate three model sizes and two sets of training data.
+
+\begin{center}
+	\begin{tabular}{ |c|c|c|c|c| }
+	 Model name & Encoder layers & Attention heads & Hidden dimension & Embedding dimension \\ 
+	\hline
+	 5m & 4 & 4 & 200 & 600 \\ 
+	 10m & 6 & 6 & 250 & 400 \\ 
+	 25m & 8 & 8 & 400 & 800 \\ 
+	 \hline
+	\end{tabular}
+\end{center}
+
+\subsection{Genotyping \& Phasing}
+
+To check the accuracy of predicted genotypes and phasing of variants we compared calls to the known genotype and, when available, phase from Genome-in-a-Bottle (GIAB).
+Approximately 99.5\% of homozygous  and 99.0\% of heterozygous calls matched the true genotype, indicating that our model learns to construct both haplotypes accurately.
+For phasing accuracy, we identified pairs of heterozygous variants less than 100 base pairs apart and compared their phasing predictions with the known phasing information from GIAB.
+The phasing information in GIAB samples is only available for HG001, HG002 and HG005 cell lines, so our analysis is restricted to only those samples.
+Approximately 99.4\% of predicted variant pairs had phase that matched the true phase from GIAB.
+
+Accurate phase prediction requires that information regarding which reads support a given haplotype is propogated among input tokens (genomic positions).
+Our model's relatively high phasing accuracy suggests that
 
 \subsection{Complex regions}
- - Not sure how to define these, but ideally we'd look for variants in regions where the read mapper had lots of overlapping insertion / deletions in reads and/or softclipping
+ While many genomic variants are trivial to identify by examination of mapped reads, some regions require a more sophisticated reassembly of reads to identify the true haplotype.
+We examine the ability of our model to perform this more complicated reconstruction by identifying real variants in regions containing multiple conflicting read mappings in addition to extensive read soft-clipping.
+ To identify such regions we first obtain all indel variants from Genome-in-a-Bottle greater than X bp.
+We then examine the reads in the alignment file and flag any region that contains more than Y distinct indel start sites or at least Z reads with more than 10bp softclipped.
+We then
 
@@
 
-\subsection{Model size experiments}
-
- small (5M) medium (10M), large (25M) pretty big (50M)
+\subsection{Ablation experiments}
 
@@
 
- \subsection{Phasing accuracy}
-
 \section{Discussion}
@@
 We describe a new approach to the problem of detecting sequence variants in next-generation sequencing (NGS) data based on a single, end-to-end deep learning model.
-Our approach envisions variant detection as a sequence-to-sequence modeling problem, akin to language translation, and leverages the successful transformer architecture to accomplish the task.
+Our approach envisions variant detection as a sequence-to-sequence modeling problem, akin to language translation, and leverages the successful transformer architecture to accomplish this task.
 In contrast to other variant detection methods (...), our approach does not rely on handcrafted statistical techniques such as de Bruijn graphs, hidden Markov models, or heuristic thresholding and instead learns to construct accurate haplotypes directly from aligned NGS reads.
 
- Convolutions - do they make sense?
+Our analysis demonstrates that the model learns many features of modern, state of the art variant callers.
+For instance, our model learns to reconstruct large, complex variants in the presence of ambiguous mapping and extensive soft clipping.
+In addition, the model learns to employ local context and read-specific features to predict phase
 
  Loss functions, there's probably something better. The Smith Waterman loss sounds really cool, but didn't lead to high sensitivity in our hands.
+
+For this work we have employed an unmodified transformer model with a very simple decoder consisting of just two fully connected layers, and we leave to future work the exploration of the many transformer variations proposed in recent years (vision transformers, swin transformers, fast-former, etc.)
 
@@
 \section{References}
-
- \vspace{8pt}
- Conrad, Donald F., et al. "Origins and functional impact of copy number variation in the human genome." Nature 464.7289 (2010): 704.
 

### 2022-05-23 manuscript/manuscript.tex
Try some BAM feats for classifier

@@
 All variant discovery tools must address two challenges.
-First, they must identify candidate alleles in a region.
-Then, given a set of allele candidates, they must assess the likelihood of each candidate (or pairs of candidates in the diploid case), to determine the most likely alleles.
+First, callers must identify potential  alleles in a region.
+Then, given a set of allele candidates, they must assess the likelihood of each candidate (or pairs of candidates in the diploid case), to determine which should be included in the caller output.
 Early callers, such as samtools / mpileup (Li et al.
 2009) and the UnifiedGenotyper tool from the Genome Analysis ToolKit (GATK, De Pristo et al.
-2011) rely on the read aligner to generate candidate allele candidates, and utilize ad-hoc heuristics and thresholds to determine likely alleles.
+2011) rely on the read aligner to generate candidate allele candidates, and utilize ad-hoc heuristics and thresholds to determine likely alleles (Nielsen et al.
+2011).
 Later tools (e.g.
@@
 2018, Cooke et al.
-2021) incorporated local re-alignment of reads in each region of interest, resulting in a significant improvement in the precision and specificity of variant calls.
+2021) incorporated local re-assembly of reads in each region of interest, resulting in a significant improvement in the precision and specificity of variant calls.
 For instance, the HaplotypeCaller tool identifies candidate haplotypes by constructing de Bruijn graphs from k-mers present in the reads, and assesses haplotype likelihoods with a pair Hidden Markov Model (HMM).
@@
 
-Maybe brief explanation of how transformers work?
-
-Here, we explore a new approach to variant detection that replaces the bespoke statistical techniques used in previous tools with a single, end-to-end machine learning model.
+While replacing the statistical and heuristic components of the allele likelihood calculation with a deep learning model has resulted in improved specificity of variant callers, current tools continue to rely on complex, handcrafted statistical machinery to identify possible alleles.
+Here, we describe a new approach that replaces both the candidate allele generation and the likelihood calculation with a single, learned model.
 We recast variant detection as a sequence-to-sequence translation problem in which the input sequence is the series of bases aligning to a single genomic location, and the output sequences are the two predicted haplotypes.
@@
 \vspace{8pt}
+Nielsen, Rasmus, et al. "Genotype and SNP calling from next-generation sequencing data." Nature Reviews Genetics 12.6 (2011): 443-451.
+
+\vspace{8pt}
 Paszke, A., et al.

### 2022-05-23 dnaseq2seq/buildclf.py

@@ -1,10 +1,10 @@
 #!/usr/bin/env python
 
 
-import sys
+from enum import Enum
 import yaml
 import numpy as np
-import scipy.stats as stats
+import bisect
 import pysam
 import pickle
 import sklearn
@@ -19,6 +19,29 @@ logging.basicConfig(format='[%(asctime)s]  %(name)s  %(levelname)s  %(message)s'
                     datefmt='%m-%d %H:%M:%S',
                     level=logging.INFO)
 
+
+SUPPORTS_REF=0
+SUPPORTS_ALT=1
+SUPPORTS_OTHER=2
+
+
+class CigarOperator(Enum):
+    BAM_CMATCH = 0  # 'M'
+    BAM_CINS = 1  # 'I'
+    BAM_CDEL = 2  # 'D'
+    BAM_CREF_SKIP = 3  # 'N'
+    BAM_CSOFT_CLIP = 4  # 'S'
+    BAM_CHARD_CLIP = 5  # 'H'
+    BAM_CPAD = 6  # 'P'
+    BAM_CEQUAL = 7  # '='
+    BAM_CDIFF = 8  # 'X'
+    BAM_CBACK = 9  # 'B'
+
+class OffsetOutOfBoundsException(Exception):
+    """
+    Raised when computing read position offset if we end up with an invalid value
+    """
+    pass
 
 def trim_suffix(ref, alt):
     r = ref
@@ -38,6 +61,16 @@ def trim_prefix(ref, alt):
             return offset, ref[offset:], alt[offset:]
         offset += 1
     return offset, ref[offset:], alt[offset:]
+
+
+def full_prefix_trim(ref, alt):
+    """
+    Trim prefix, leaving empty alleles if necessary
+    """
+    for i, (r, a) in enumerate(zip(ref + "$", alt + "$")):
+        if r != a:
+            break
+    return i, ref[i:], alt[i:]
 
 
 def find_var(varfile, chrom, pos, ref, alt):
@@ -70,9 +103,129 @@ def var_af(varfile, chrom, pos, ref, alt):
             af = float(af)
     return af
 
+def get_query_pos(read, ref_pos):
+    """
+    Returns the offset in the read sequence corresponding to the given reference position. Follows the
+    rules given by bisect.bisect_left, so if multiple read reference positions correspond to the ref_pos,
+    then the index of the leftmost is returned.
+    :param read: pysam.AlignedSegment
+    :param ref_pos: Reference genomic coordinate
[…]

### 2022-05-23 dnaseq2seq/buildclf.py
Update header

@@ -263,7 +263,10 @@ def feat_names():
             "max_win_offset",
             "dp",
             "af",
-            "het"
+            "het",
+            "vaf",
+            "highmq_vaf",
+            "altreads",
         ]
 
 

### 2022-05-23 dnaseq2seq/buildclf.py
Fix conf reading

@@ -297,7 +297,7 @@ def load_model(path):
         return pickle.load(fh)
 
 
-def train_model(conf, threads, var_freq_file, feat_csv=None, labels_csv=None):
+def train_model(conf, threads, var_freq_file, feat_csv=None, labels_csv=None, reference_filename=None):
     alltps = []
     allfps = []
     var_freq_file = pysam.VariantFile(var_freq_file)
@@ -308,10 +308,12 @@ def train_model(conf, threads, var_freq_file, feat_csv=None, labels_csv=None):
     else:
         feat_fh = None
 
-    for tpf in conf['tps']:
-        alltps.extend(extract_feats(tpf, var_freq_file, feat_fh))
-    for fpf in conf['fps']:
-        allfps.extend(extract_feats(fpf, var_freq_file, feat_fh))
+    for sample in conf.keys():
+        aln = pysam.AlignmentFile(sample['bam'], reference_filename=reference_filename)
+        if 'tps' in sample:
+            alltps.extend(extract_feats(sample['tps'], aln, var_freq_file, feat_fh))
+        if 'fps' in sample:
+            allfps.extend(extract_feats(sample['fps'], aln, var_freq_file, feat_fh))
 
     if feat_fh:
         feat_fh.close()
@@ -344,7 +346,7 @@ def predict(model, vcf, **kwargs):
         print(var, end='')
 
 
-def predict_one_record(loaded_model, var_rec, var_freq_file, **kwargs):
+def predict_one_record(loaded_model, var_rec, aln, var_freq_file, **kwargs):
     """
     given a loaded model object and a pysam variant record, return classifier quality
     :param loaded_model: loaded model object for classifier
@@ -352,7 +354,7 @@ def predict_one_record(loaded_model, var_rec, var_freq_file, **kwargs):
     :param kwargs:
     :return: classifier quality
     """
-    feats = var_feats(var_rec, var_freq_file)
+    feats = var_feats(var_rec, aln, var_freq_file)
     prediction = loaded_model.predict_proba(feats[np.newaxis, ...])
     return prediction[0, 1]
 
@@ -364,7 +366,8 @@ def train(conf, output, **kwargs):
             threads=kwargs.get('threads'), 
             var_freq_file=kwargs.get('freq_file'),
             feat_csv=kwargs.get('feat_csv'),
-            labels_csv=kwargs.get('labels_csv'))
[…]

### 2022-05-23 dnaseq2seq/buildclf.py
Fix up conf, div by zero

@@ -218,14 +218,16 @@ def bamfeats(var, aln):
 def var_feats(var, aln, var_freq_file):
     feats = []
     bamreads, pos_ref, pos_alt, neg_ref, neg_alt, highmq_ref, highmq_alt = bamfeats(var, aln)
-    if bamreads > 0:
+    if pos_ref + neg_ref + pos_alt + neg_alt > 0:
         vaf = (pos_alt + neg_alt) / (pos_ref + neg_ref + pos_alt + neg_alt)
     else:
         vaf = 0.0
+
     if (highmq_ref + highmq_alt) > 0:
         mq_vaf = (highmq_alt) / (highmq_alt + highmq_ref)
     else:
         mq_vaf = 0.0
+  
     feats.append(var.qual)
     feats.append(len(var.ref))
     feats.append(max(len(a) for a in var.alts))
@@ -309,11 +311,11 @@ def train_model(conf, threads, var_freq_file, feat_csv=None, labels_csv=None, re
         feat_fh = None
 
     for sample in conf.keys():
-        aln = pysam.AlignmentFile(sample['bam'], reference_filename=reference_filename)
-        if 'tps' in sample:
-            alltps.extend(extract_feats(sample['tps'], aln, var_freq_file, feat_fh))
-        if 'fps' in sample:
-            allfps.extend(extract_feats(sample['fps'], aln, var_freq_file, feat_fh))
+        aln = pysam.AlignmentFile(conf[sample]['bam'], reference_filename=reference_filename)
+        if 'tps' in conf[sample]:
+            alltps.extend(extract_feats(conf[sample]['tps'], aln, var_freq_file, feat_fh))
+        if 'fps' in conf[sample]:
+            allfps.extend(extract_feats(conf[sample]['fps'], aln, var_freq_file, feat_fh))
 
     if feat_fh:
         feat_fh.close()

### 2022-05-28 manuscript/manuscript.tex
Manuscript updates

--- 2022-05-20 manuscript/manuscript.tex
@@
 
+While transformers have not been explored for variant detection, a model similar to ours was proposed by Baid et al.
+(2021) for generation of accurate consensus reads from PacBio data.
+The Baid model (`DeepConsensus') used a network architecture similar to ours, but included an extra token representing a gap between the aligned sequences, and a novel loss function that accounted for gaps between the predicted and true sequence in a region.
+In contrast we use a simple cross-entropy loss function, and do not introduce gaps in the aligned sequences compared to the reference.
+
 \section{Methods}
@@
 The decoder consists of a single fully-connected layer followed by two additional fully connected components that each generate a single haplotype.
-Predictions are discrete probability distributions over the four possible bases at each predicted output position.
+Predictions are discrete probability distributions over the four possible bases at each predicted output position, produced with a softmax function over the raw outputs.
 The decoder is of fixed size, and each predicted haplotype contains the same number of predicted bases as the number of positions $g$ in the input.
@@
 We additionally included `true negative' regions where no known variant was present.
-The total number of regions generated per sample was 325,000, yielding 5.53M total training regions across our 17 samples.
+For all analyses described here, we obtain training data only from the human autosomes 1-20, and hold out chromosomes 21 and 22 for model evaluation.
+We examine three different sizes of training data sets, with 5.5M, 10.4M, and 20.4M examples.
 
-For all analyses described here, we obtain training data only from the human autosomes 1-20, and hold out chromosomes 21 and 22 for model evaluation.
+\subsection{Training procedure}
+
+Models were implemented in Pytorch (v1.10, Paszke et al 2019). Models were trained for 25 epochs unless noted otherwise, using the Adam optimizer and a learning rate of $5e-5$.
 
@@
 Because incorrect insertions and deletions cause mismatches at every downstream (higher index) position, such mistakes are penalized more heavily when they occur at lower indexes than higher indexes.
-Thus the model learns to generate high quality results for the first portion of the window where inaccuracies are punished severely, and is less concerned with accuarcy at high indexes.
+Thus the model learns to generate high quality results for the first portion of the window where inaccuracies are punished severely, and is less concerned with accuracy at high indexes.
 
@@
 
-Given an alignment file in BAM or CRAM format, we first identify regions where a potential variant might exist, and then run forward passes of the model using a sliding window over the region.
+Given an alignment file in BAM or CRAM format, we first identify regions where a potential variant might exist.
 Any genomic position in which at least three reads contain a base that differs from the reference or an indel are flagged as potentially containing a variant.
@@
 Each haplotype is aligned via Smith-Waterman to the reference sequence, and any mismatching positions are converted to variant calls.
-For each variant we record the number of windows in which each variant was detected, the number of variants in cis and trans, and the mean probability of the variant bases, and the position of the variant within each window.
+For each variant we record the number of windows in which each variant was detected, the number of variants in cis and trans, and the mean probability of the variant bases, and the distance of the variant from the window start coordinate.
 
-\subsection{Variant quality calculation and false positive removal}
+\subsection{Variant quality calculation}
 
 The above variant detection procedure has two shortcomings.
-First, if every detected variant is emitted, then specificity is poor.
-Second, variant quality scores are poorly calibrated and do not closely reflect the probability that a variant is real.
+First, if every potential variant is emitted, then specificity is poor.
+Second, the model does not innately produce well calibrated variant quality scores.
 To address these shortcomings we introduce a post-hoc random forest classifier that we train to discriminate true and false-positive variant calls (see table X for features).
-The resulting score is used as a final variant quality score that can be used for tuning the tradeoff between sensitivity and specificity.
+The resulting score is used as a final variant quality score suitable for tuning the tradeoff between sensitivity and specificity.
+
+The random forest classifier is trained by calling variants on selected regions from autosomes 1-4 totalling approximately 120Mb.
+True and false positive variant calls were identified with the \textit{vcfeval} tool (Cleary et al.
+2015).
+We used the scikit-learn implementation of the random forest with 100 trees, a maximum tree depth of 25, and 'balanced' class weighting option.
 
@@
 \subsection{Complex regions}
- While many genomic variants are trivial to identify by examination of mapped reads, some regions require a more sophisticated reassembly of reads to identify the true haplotype.
-We examine the ability of our model to perform this more complicated reconstruction by identifying real variants in regions containing multiple conflicting read mappings in addition to extensive read soft-clipping.
- To identify such regions we first obtain all indel variants from Genome-in-a-Bottle greater than X bp.
-We then examine the reads in the alignment file and flag any region that contains more than Y distinct indel start sites or at least Z reads with more than 10bp softclipped.
-We then
+ While many genomic variants are trivial to identify by examination of mapped reads, some regions require a more sophisticated reassembly to identify the true haplotype.
+We examine the ability of our model to perform this more complicated reconstruction by identifying real variants in regions containing multiple conflicting read mappings or extensive read soft-clipping.
+
+ To identify variants requiring complex reconstruction, we first obtain all indel variants from Genome-in-a-Bottle greater than 4 bp from the validation regions on chromosomes 21 and 22.
+For each variant, we examine the reads in the alignment file that overlap the variant position, and flag any with with more than 4 distinct indel start sites or at least 20\% reads with more than 10bp softclipped as requiring 'complex' reconstruction.
+These variants are moved to a separate truth set.
+After running our variant calling procedure on the alignment file, we compute the fraction of variants from the truth set that were identified by our model.
+This procedure was repeated for all 17 samples in our data set.
+
+ Our model is able to reconstruct the correct haplotypes for ~90.8\% (mid 50\% 86.4-95.2) of these variants, somewhat lower than HaplotypeCaller accuracy of ~95.8\% (mid 50\% 91.0-100.0).
+Although model performance does not match current state-of-the-art callers, it nonetheless is able to reconstruct the true haplotype in most cases.
+Overall, these results indicate that the model learns a reconstruction procedure more nuanced than a simple column-by-column examination of reads,
 
@@
 In contrast to other variant detection methods (...), our approach does not rely on handcrafted statistical techniques such as de Bruijn graphs, hidden Markov models, or heuristic thresholding and instead learns to construct accurate haplotypes directly from aligned NGS reads.
+On short read WGS data, our approach demonstrates performance similar to current top-performing callers, with somewhat higher sensitivity but lower precision for insertion/deletion variants ()
 
-Our analysis demonstrates that the model learns many features of modern, state of the art variant callers.
+Our analysis demonstrates that the model learns many capabilities of modern, state of the art variant callers.
 For instance, our model learns to reconstruct large, complex variants in the presence of ambiguous mapping and extensive soft clipping.
-In addition, the model learns to employ local context and read-specific features to predict phase
+While our model's sensitivity is ~5\% less than HaplotypeCaller for variants in these complex regions, we nonetheless are able to reconstruct the correct haplotypes in more than 90\% of cases.
+In addition, the model learns to employ local context and read-specific features to predict phase and genotype accurately, and pairs of variants are phased correctly in ~99\% of cases where variants are within 100bp.
 
- Loss functions, there's probably something better. The Smith Waterman loss sounds really cool, but didn't lead to high sensitivity in our hands.
+All models in this work were trained using a standard cross-entropy loss function.
+Cross-entropy loss is probably not optimal for variant detection, because insertion and deletion errors are more severely punished than SNVs, and the loss has a strong dependence on the relative position of the variant in the input sequence.
+During development we experimented with the differentiable Smith-Waterman loss function as described by Petti et al.
+(2021), but this lead to reduced sensitivity in our experiments.
 
-For this work we have employed an unmodified transformer model with a very simple decoder consisting of just two fully connected layers, and we leave to future work the exploration of the many transformer variations proposed in recent years (vision transformers, swin transformers, fast-former, etc.)
+For this work we have employed an unmodified transformer model with a very simple decoder consisting of just two fully connected layers, and we leave to future work exploration of the many transformer variations proposed in recent years (e.g.
+Dosovitskiy 2018, Fedus et al.
+2021, Liu et al 2021, Wu et al.
+2021).
+Our simple decoder only predicts haplotypes that are equal in length to the number of input tokens (genomic positions), but a natural improvement would be to add a more sophisticated decoder capable of producing a variable length sequence to handle larger insertion and deletion events.
+In addition, newer transformer models that avoid the $O(n^2)$ scaling with the number of input tokens may be able to handle longer input sequence lengths, and therefore may be able to detect larger variants.
 
@@
 \vspace{8pt}
+Baid, Gunjan, et al. "DeepConsensus: Gap-Aware Sequence Transformers for Sequence Correction." bioRxiv (2021).
+
+\vspace{8pt}
+Cleary, John G., et al. "Comparing variant call files for performance benchmarking of next-generation sequencing variant calling pipelines." BioRxiv (2015): 023754.
+
+\vspace{8pt}
 Cooke, Daniel P., David C. Wedge, and Gerton Lunter. "A unified haplotype-based method for accurate and comprehensive variant calling." Nature biotechnology 39.7 (2021): 885-892.
@@
 DePristo, Mark A., et al. "A framework for variation discovery and genotyping using next-generation DNA sequencing data." Nature genetics 43.5 (2011): 491-498.
+
+\vspace{8pt}
+Dosovitskiy, Alexey, et al. "An image is worth 16x16 words: Transformers for image recognition at scale." arXiv preprint arXiv:2010.11929 (2020).
+
+\vspace{8pt}
+Fedus, William, Barret Zoph, and Noam Shazeer. "Switch transformers: Scaling to trillion parameter models with simple and efficient sparsity." arXiv preprint arXiv:2101.03961 (2021).
 
@@
 \vspace{8pt}
+Liu, Ze, et al. "Swin transformer: Hierarchical vision transformer using shifted windows." Proceedings of the IEEE/CVF International Conference on Computer Vision. 2021.
+
+\vspace{8pt}
 Marco-Sola S., Sammeth M., Guigó R., Ribeca P. "The GEM mapper: fast, accurate and versatile alignment by filtration". Nat Methods. (2012);9(12):1185-1188. doi:10.1038/nmeth.2221
@@
 Retrieved from http://papers.neurips.cc/paper/9015-pytorch-an-imperative-style-high-performance-deep-learning-library.pdf
+
+\vspace{8pt}
+Petti, Samantha, et al. "End-to-end learning of multiple sequence alignments with differentiable Smith-Waterman." BioRxiv (2021).
 
@@
 
+\vspace{8pt}
+Wu, Chuhan, et al. "Fastformer: Additive attention can be all you need." arXiv preprint arXiv:2108.09084 (2021).
+
 \end{document}

### 2022-05-28 manuscript/manuscript.tex
manuscript

--- 2022-05-23 manuscript/manuscript.tex
@@
 Next Generation Sequencing (NGS) workflows typically begin by aligning sequenced fragments to a reference genome, optionally performing some alignment post-processing steps such as identification of duplicated fragments, and then identifying sequence variants relative to the reference genome.
-The variant identification step is challenging for many reasons, including technical artifacts in the sequencing process, improper alignments due to sequence homology or low sequence complexity, and the stochastic nature of the library preparation and amplification steps [citation here would be nice].
+The variant identification step is challenging for many reasons, including technical artifacts in the sequencing process, improper alignments due to sequence homology or low sequence complexity, and the stochastic nature of the library preparation and amplification steps (Li 2014).
 
@@
 2017).
-We demonstrate that this approach identifies real genomic variants with sensitivity equivalent to current top-performing callers, albeit with slightly lower specificity.
+We demonstrate that this approach identifies real genomic variants with sensitivity similar to current top-performing callers, albeit with lower specificity.
 
@@
 For this special row we inserted a base quality of 100 for every position and did not set any of the other flags.
-Resulting tensors had dimension $[g, r, 9]$, where $g$ represents genomic position, $r$ represents read, and 9 is the number of features.
+Resulting tensors had dimension $[150, 100, 9]$, where the indices are for the genomic position, read, and feature respectively.
 
@@
 Regions of the reference genome overlapping the high-confidence regions from Genome-in-a-Bottle were subdivided into 150bp nonoverlapping windows, and these regions labelled according to the presence of variants.
-Separate labels were generated for regions conataining a single SNV, deletion, or insertion, as well as regions containing multiple insertion-deletions, or those containing variants intersecting  low-complexity regions.
+Separate labels were generated for regions conataining a single SNV, deletion, or insertion, as well as regions containing multiple insertion-deletions, or those containing variants intersecting  low complexity regions.
+Low complexity regions were obtained from Genome-in-a-Bottle stratification BED files.
 We additionally included `true negative' regions where no known variant was present.
@@
 We use a standard cross-entropy loss function to compare predicted haplotypes to known haplotype sequences for each region, and simply compare each prediction at window index (position) $i$ to the true base at index $i$.
-While we obtain satisfactory results with this approach, two issues were apparent.
-First, predicted haplotype sequences may have inserted or deleted bases relative to the true haplotypes.
-Because incorrect insertions and deletions cause mismatches at every downstream (higher index) position, such mistakes are penalized more heavily when they occur at lower indexes than higher indexes.
-Thus the model learns to generate high quality results for the first portion of the window where inaccuracies are punished severely, and is less concerned with accuarcy at high indexes.
-
-A second complexity arises due to ambiguity regarding which haplotype should contain a variant.
+A complexity arises due to ambiguity regarding which haplotype should contain a variant.
 The two haplotypes present in a sample have no inherent order.
-In training data variants are assigned arbitrarily to haplotypes, and this randomness in haplotype assignment means model cannot learn to predict which haplotype should contain a given variant.
+In training data variants are assigned arbitrarily to haplotypes, and this randomness in haplotype assignment means the model cannot learn to predict which haplotype should contain a given variant.
 To combat this ambiguity, we calculate loss in both possible configurations - predicted haplotype 0 with training haplotype 0 and 1 with 1, as well as predicted haplotype 0 with training haplotype 1 and 1 with 0.
@@
 
-Given a region containing suspected variants, we perform multiple overlapping forward passes of the model with step size $k$.
+Given a region containing suspected variants, we perform multiple overlapping forward passes of the model with step size $k$, where $k=50$ for the results reported here.
 On each forward pass the model produces two predicted haplotypes by taking the base with the high predicted probability at each position.
@@
 For instance, our model learns to reconstruct large, complex variants in the presence of ambiguous mapping and extensive soft clipping.
-In addition, the model learns to employ local context and read-specific features to predict phase
+In addition, the model learns to employ local context and read-specific features to predict phase and genotype accurately.
 
- Loss functions, there's probably something better. The Smith Waterman loss sounds really cool, but didn't lead to high sensitivity in our hands.
+In this work we have used a standard cross-entropy loss function that simply compares each prediction at base $i$ to the label at base $i$.
+While training with this loss yields accurate results, it has two significant shortcomings.
+First, the impact of an insertion / deletion (indel) error depends on its position in the region, with indels at lower indices leading to a greater number of mismatches and high loss compared to indels at higher indices.
+Secondly, indel errors ar punished more severely than similar SNV errors.
+We hypothesize the second fact explains the lower performance of our model on SNV calls compared to indels.
+Alternative, alignment-based loss functions such as those described by Baid et al.
+(2021) and Petti et al.
+(2021) may provide a path forward, however the Petti implementation lead to lower sensitivity in our experiments.
 
@@
 \vspace{8pt}
-Li, Heng, et al. "The sequence alignment/map format and SAMtools." Bioinformatics 25.16 (2009): 2078-2079.
+Li, Heng. "Toward better understanding of artifacts in variant calling from high-coverage samples." Bioinformatics 30.20 (2014): 2843-2851.
+
+\vspace{8pt}
+Li, Heng, et al. "The sequence alignment/map format and SAMtools." Bioinformatics 25.16 (2009): 
+2078-2079.
 

### 2022-05-28 manuscript/manuscript.tex

--- 2022-05-28 manuscript/manuscript.tex
@@
 Next Generation Sequencing (NGS) workflows typically begin by aligning sequenced fragments to a reference genome, optionally performing some alignment post-processing steps such as identification of duplicated fragments, and then identifying sequence variants relative to the reference genome.
-The variant identification step is challenging for many reasons, including technical artifacts in the sequencing process, improper alignments due to sequence homology or low sequence complexity, and the stochastic nature of the library preparation and amplification steps [citation here would be nice].
+The variant identification step is challenging for many reasons, including technical artifacts in the sequencing process, improper alignments due to sequence homology or low sequence complexity, and the stochastic nature of the library preparation and amplification steps (Li 2014).
 
 All variant discovery tools must address two challenges.
-First, they must identify candidate alleles in a region.
-Then, given a set of allele candidates, they must assess the likelihood of each candidate (or pairs of candidates in the diploid case), to determine the most likely alleles.
+First, callers must identify potential  alleles in a region.
+Then, given a set of allele candidates, they must assess the likelihood of each candidate (or pairs of candidates in the diploid case), to determine which should be included in the caller output.
 Early callers, such as samtools / mpileup (Li et al.
 2009) and the UnifiedGenotyper tool from the Genome Analysis ToolKit (GATK, De Pristo et al.
-2011) rely on the read aligner to generate candidate allele candidates, and utilize ad-hoc heuristics and thresholds to determine likely alleles.
+2011) rely on the read aligner to generate candidate allele candidates, and utilize ad-hoc heuristics and thresholds to determine likely alleles (Nielsen et al.
+2011).
 Later tools (e.g.
@@
 We use a standard cross-entropy loss function to compare predicted haplotypes to known haplotype sequences for each region, and simply compare each prediction at window index (position) $i$ to the true base at index $i$.
-While we obtain satisfactory results with this approach, two issues were apparent.
-First, predicted haplotype sequences may have inserted or deleted bases relative to the true haplotypes.
-Because incorrect insertions and deletions cause mismatches at every downstream (higher index) position, such mistakes are penalized more heavily when they occur at lower indexes than higher indexes.
-Thus the model learns to generate high quality results for the first portion of the window where inaccuracies are punished severely, and is less concerned with accuracy at high indexes.
-
-A second complexity arises due to ambiguity regarding which haplotype should contain a variant.
+A complexity arises due to ambiguity regarding which haplotype should contain a variant.
 The two haplotypes present in a sample have no inherent order.
-In training data variants are assigned arbitrarily to haplotypes, and this randomness in haplotype assignment means model cannot learn to predict which haplotype should contain a given variant.
+In training data variants are assigned arbitrarily to haplotypes, and this randomness in haplotype assignment means the model cannot learn to predict which haplotype should contain a given variant.
 To combat this ambiguity, we calculate loss in both possible configurations - predicted haplotype 0 with training haplotype 0 and 1 with 1, as well as predicted haplotype 0 with training haplotype 1 and 1 with 0.
@@
 
-Given a region containing suspected variants, we perform multiple overlapping forward passes of the model with step size $k$.
+Given a region containing suspected variants, we perform multiple overlapping forward passes of the model with step size $k$, where $k=50$ for the results reported here.
 On each forward pass the model produces two predicted haplotypes by taking the base with the high predicted probability at each position.
 Each haplotype is aligned via Smith-Waterman to the reference sequence, and any mismatching positions are converted to variant calls.
-For each variant we record the number of windows in which each variant was detected, the number of variants in cis and trans, and the mean probability of the variant bases, and the distance of the variant from the window start coordinate.
+For each variant we record the number of windows in which each variant was detected, the number of variants in cis and trans, and the mean probability of the variant bases, and the position of the variant within each window.
 
@@
 In contrast to other variant detection methods (...), our approach does not rely on handcrafted statistical techniques such as de Bruijn graphs, hidden Markov models, or heuristic thresholding and instead learns to construct accurate haplotypes directly from aligned NGS reads.
-On short read WGS data, our approach demonstrates performance similar to current top-performing callers, with somewhat higher sensitivity but lower precision for insertion/deletion variants ()
 
-Our analysis demonstrates that the model learns many capabilities of modern, state of the art variant callers.
+Our analysis demonstrates that the model learns many features of modern, state of the art variant callers.
 For instance, our model learns to reconstruct large, complex variants in the presence of ambiguous mapping and extensive soft clipping.
-While our model's sensitivity is ~5\% less than HaplotypeCaller for variants in these complex regions, we nonetheless are able to reconstruct the correct haplotypes in more than 90\% of cases.
 In addition, the model learns to employ local context and read-specific features to predict phase and genotype accurately, and pairs of variants are phased correctly in ~99\% of cases where variants are within 100bp.
 
-All models in this work were trained using a standard cross-entropy loss function.
-Cross-entropy loss is probably not optimal for variant detection, because insertion and deletion errors are more severely punished than SNVs, and the loss has a strong dependence on the relative position of the variant in the input sequence.
-During development we experimented with the differentiable Smith-Waterman loss function as described by Petti et al.
-(2021), but this lead to reduced sensitivity in our experiments.
+In this work we have used a standard cross-entropy loss function that simply compares each prediction at base $i$ to the label at base $i$.
+While training with this loss yields accurate results, it has two significant shortcomings.
+First, the impact of an insertion / deletion (indel) error depends on its position in the region, with indels at lower indices leading to a greater number of mismatches and high loss compared to indels at higher indices.
+Secondly, indel errors ar punished more severely than similar SNV errors.
+We hypothesize the second fact explains the lower performance of our model on SNV calls compared to indels.
+Alternative, alignment-based loss functions such as those described by Baid et al.
+(2021) and Petti et al.
+(2021) may provide a path forward, however the Petti implementation lead to lower sensitivity in our experiments.
 
@@
 Marco-Sola S., Sammeth M., Guigó R., Ribeca P. "The GEM mapper: fast, accurate and versatile alignment by filtration". Nat Methods. (2012);9(12):1185-1188. doi:10.1038/nmeth.2221
+
+\vspace{8pt}
+Nielsen, Rasmus, et al. "Genotype and SNP calling from next-generation sequencing data." Nature Reviews Genetics 12.6 (2011): 443-451.
 

### 2022-05-28 dnaseq2seq/buildclf.py
Calling fixes for more features

@@ -312,10 +312,18 @@ def train_model(conf, threads, var_freq_file, feat_csv=None, labels_csv=None, re
 
     for sample in conf.keys():
         aln = pysam.AlignmentFile(conf[sample]['bam'], reference_filename=reference_filename)
-        if 'tps' in conf[sample]:
-            alltps.extend(extract_feats(conf[sample]['tps'], aln, var_freq_file, feat_fh))
-        if 'fps' in conf[sample]:
-            allfps.extend(extract_feats(conf[sample]['fps'], aln, var_freq_file, feat_fh))
+        tps = conf[sample].get('tps')
+        if type(tps) == str:
+            alltps.extend(extract_feats(tps, aln, var_freq_file, feat_fh))
+        elif type(tps) == list:
+            for tp in tps:
+                alltps.extend(extract_feats(tp, aln, var_freq_file, feat_fh))
+        fps = conf[sample].get('fps')
+        if type(fps) == str:
+            allfps.extend(extract_feats(fps, aln, var_freq_file, feat_fh))
+        elif type(fps) == list:
+            for fp in fps:
+                allfps.extend(extract_feats(fp, aln, var_freq_file, feat_fh))
 
     if feat_fh:
         feat_fh.close()

### 2022-05-28 dnaseq2seq/call.py

@@ -530,7 +530,7 @@ def vars_hap_to_records(
     for rec in vcf_records:
         if classifier_model:
             rec.info["RAW_QUAL"] = rec.qual
-            rec.qual = buildclf.predict_one_record(classifier_model, rec, var_freq_file)
+            rec.qual = buildclf.predict_one_record(classifier_model, rec, aln, var_freq_file)
         vcf_file.write(rec)
 
     vcf_file.close()

### 2022-06-04 manuscript/manuscript.tex
Update readme, manuscript

@@
 \author[1]{Luobin Yang}
+\author[1]{Eric Frederickson}
+\author[1]{Katherine Noble}
+\author[1]{Joshua Coleman}
+\author[1]{Hunter Best}
 
@@
 
- - Tabulate Hap.py results across a few common stratifications, compare to other callers
+To characterize variant detection accuracy, we called variants on validation regions from chromosomes 21 and 22 for each of the 17 samples, and compared results to DeepVariant (v1.3.0, Poplin et al.
+2018), HaplotypeCaller (v.
+4.1.4.1), and Strelka (v.
+2.9.10, Kim et al.
+2018).
+Calls were obtained with the 10M model trained for 25 epochs on the 'bigger' training set (~$10^6$ training examples).
+We selected filtering criteria for each caller by reviewing ROC curves produced by \textit{vcfeval} and selecting values close to the $F_1$ maximizing value across samples.
+Jovian calls were filtered at 0.4, HaplotypeCaller at 60 (phred-scaled), Strelka at 4 (phred-scaled), and DeepVariant at 0.
+Sensitivity, precision, and related calculations were computed with the hap.py tool (Krushe et al.
+2019) using Genome-in-a-Bottle (v.4.2.1) as the truth set of variants.
 
-\subsection{Ablation experiments}
-
- Ablate the 2D positional encoding.
-
- Ablate reference as read 0
+Across all regions, Jovian's sensitivity and precision for indels is slightly higher than that for HaplotypeCaller, and slightly lower than DeepVariant or Strelka2.
+For SNV calls Jovian has middling sensitivity and the lowest precision of the callers.
 
@@
 (2021) may provide a path forward, however the Petti implementation lead to lower sensitivity in our experiments.
+
+While our model has comparable sensitivity to the other callers we examined, our specificity for SNV calls is lower.
+Strategies to improve SNV specificity include adding additional training data with false positive SNV calls, adding new features to the quality-estimation classifier, or possibly running an ensemble of models.
 
@@
 Kim, Sangtae, et al. "Strelka2: fast and accurate calling of germline and somatic variants." Nature methods 15.8 (2018): 591-594.
+
+\vspace{8pt}
+Krusche, Peter, et al. "Best practices for benchmarking germline small-variant calls in human genomes." Nature biotechnology 37.5 (2019): 555-560.
 

### 2022-06-04 manuscript/manuscript.tex
Manuscript

@@
 
-Accurate reconstruction of the DNA sequence in a sample is a fundamental component of modern genomics analysis.
-Next Generation Sequencing (NGS) workflows typically begin by aligning sequenced fragments to a reference genome, optionally performing some alignment post-processing steps such as identification of duplicated fragments, and then identifying sequence variants relative to the reference genome.
-The variant identification step is challenging for many reasons, including technical artifacts in the sequencing process, improper alignments due to sequence homology or low sequence complexity, and the stochastic nature of the library preparation and amplification steps (Li 2014).
+Accurate detection of DNA sequence variants in a sample from is a fundamental component of modern genomics analysis. Tools that aim to
 
@@
 
-To check the accuracy of predicted genotypes and phasing of variants we compared calls to the known genotype and, when available, phase from Genome-in-a-Bottle (GIAB).
+To check the accuracy of predicted genotypes and phasing of variants we compared calls to the known genotype and, when available, phase from Genome-in-a-Bottle (GIAB) variant data.
 Approximately 99.5\% of homozygous  and 99.0\% of heterozygous calls matched the true genotype, indicating that our model learns to construct both haplotypes accurately.
 For phasing accuracy, we identified pairs of heterozygous variants less than 100 base pairs apart and compared their phasing predictions with the known phasing information from GIAB.
-The phasing information in GIAB samples is only available for HG001, HG002 and HG005 cell lines, so our analysis is restricted to only those samples.
-Approximately 99.4\% of predicted variant pairs had phase that matched the true phase from GIAB.
+The phasing information in GIAB samples is only available for HG001, HG002 and HG005 cell lines, so analysis was restricted to those samples.
+Phase predictions matched GIAB phase for approximately 96.9\% of SNV pairs and 99.3\% of indel pairs.
 
@@
 
- To identify variants requiring complex reconstruction, we first obtain all indel variants from Genome-in-a-Bottle greater than 4 bp from the validation regions on chromosomes 21 and 22.
-For each variant, we examine the reads in the alignment file that overlap the variant position, and flag any with with more than 4 distinct indel start sites or at least 20\% reads with more than 10bp softclipped as requiring 'complex' reconstruction.
+ To identify variants requiring complex reconstruction, we first obtained all indel variants from Genome-in-a-Bottle greater than 4 bp from the validation regions on chromosomes 21 and 22.
+For each variant, we examine the reads in the alignment file that overlap the variant position, and flag any with with more than 4 distinct indel start sites or at least 20\% reads with more than 10bp soft-clipped as requiring 'complex' reconstruction.
 These variants are moved to a separate truth set.
@@
 
- Our model is able to reconstruct the correct haplotypes for ~90.8\% (mid 50\% 86.4-95.2) of these variants, somewhat lower than HaplotypeCaller accuracy of ~95.8\% (mid 50\% 91.0-100.0).
+ Our model was able to reconstruct the correct haplotypes for ~90.8\% (mid 50\% 86.4-95.2) of these variants, somewhat lower than HaplotypeCaller accuracy of ~95.8\% (mid 50\% 91.0-100.0).
 Although model performance does not match current state-of-the-art callers, it nonetheless is able to reconstruct the true haplotype in most cases.
@@
 2018), HaplotypeCaller (v.
-4.1.4.1), and Strelka (v.
+4.1.4.1), and Strelka2 (v.
 2.9.10, Kim et al.
@@
 Jovian calls were filtered at 0.4, HaplotypeCaller at 60 (phred-scaled), Strelka at 4 (phred-scaled), and DeepVariant at 0.
-Sensitivity, precision, and related calculations were computed with the hap.py tool (Krushe et al.
-2019) using Genome-in-a-Bottle (v.4.2.1) as the truth set of variants.
+Sensitivity, precision, and related calculations were computed with the \textit{hap.py} tool (Krushe et al.
+2019) using Genome-in-a-Bottle (v.4.2.1) as the benchmark variant set.
 
 Across all regions, Jovian's sensitivity and precision for indels is slightly higher than that for HaplotypeCaller, and slightly lower than DeepVariant or Strelka2.
-For SNV calls Jovian has middling sensitivity and the lowest precision of the callers.
+For SNV calls Jovian has similar sensitivity to other callers but the lowest precision of the callers ()
 
@@
 
-We describe a new approach to the problem of detecting sequence variants in next-generation sequencing (NGS) data based on a single, end-to-end deep learning model.
+We describe a new approach to the problem of detecting sequence variants in next-generation sequencing (NGS).
 Our approach envisions variant detection as a sequence-to-sequence modeling problem, akin to language translation, and leverages the successful transformer architecture to accomplish this task.
-In contrast to other variant detection methods (...), our approach does not rely on handcrafted statistical techniques such as de Bruijn graphs, hidden Markov models, or heuristic thresholding and instead learns to construct accurate haplotypes directly from aligned NGS reads.
+The sequence-to-sequence approach allows a single deep learning model to be used for candidate allele generation and haplotype construction.
+In contrast to other variant detection methods (...), our approach does not rely on any handcrafted statistical techniques such as de Bruijn graphs, hidden Markov models, or heuristic thresholding and instead learns to construct accurate haplotypes directly from aligned NGS reads.
 
 Our analysis demonstrates that the model learns many features of modern, state of the art variant callers.
-For instance, our model learns to reconstruct large, complex variants in the presence of ambiguous mapping and extensive soft clipping.
-In addition, the model learns to employ local context and read-specific features to predict phase and genotype accurately, and pairs of variants are phased correctly in ~99\% of cases where variants are within 100bp.
+For instance, our model learns to reconstruct large, complex variants in the presence of ambiguous mapping and extensive soft clipping with a sensitivity greater than 90\%.
+In addition, the model learns to employ local context and read-specific features to predict phase and genotype accurately, and pairs of variants are phased correctly in ~97-99\% of cases where variants are within 100bp.
+The model also learns to consider the surrounding read context when predicting the presence of a SNV.
+For instance, a potential SNV in a region with few reads containing reference mismatches is much more likely to be called than a similar variant in a region where many reads contain mismatches, indicating that the model has learned to be cautious of potential SNVs in `dirty' regions where many reads contain errors or are misaligned.
+
+While our model has comparable sensitivity to the other callers we examined, our specificity for SNV calls is lower.
+Many avenues existing for further improving performance, including adding additional training data with false positive SNV calls, adding new features to the quality-estimation classifier, or possibly running an ensemble of models.
 
@@
 (2021) and Petti et al.
-(2021) may provide a path forward, however the Petti implementation lead to lower sensitivity in our experiments.
-
-While our model has comparable sensitivity to the other callers we examined, our specificity for SNV calls is lower.
-Strategies to improve SNV specificity include adding additional training data with false positive SNV calls, adding new features to the quality-estimation classifier, or possibly running an ensemble of models.
+(2021) may provide a path forward, however the Petti implementation lead to lower sensitivity and much longer training times in our experiments.
 
@@
 Dosovitskiy 2018, Fedus et al.
-2021, Liu et al 2021, Wu et al.
+2021, Liu et al.
+2021, Wu et al.
 2021).
-Our simple decoder only predicts haplotypes that are equal in length to the number of input tokens (genomic positions), but a natural improvement would be to add a more sophisticated decoder capable of producing a variable length sequence to handle larger insertion and deletion events.
+Our simple decoder has a fixed length equal to the number of input tokens (genomic positions), but a natural improvement would be to add a more sophisticated decoder capable of producing a variable length sequence to handle larger insertion and deletion events.
 In addition, newer transformer models that avoid the $O(n^2)$ scaling with the number of input tokens may be able to handle longer input sequence lengths, and therefore may be able to detect larger variants.
+
+In his essay `The Bitter Lesson' (Sutton 2019), noted researcher Richard Sutton commented that methods that leverage computation and search are often more successful in the long term when compared to methods that directly model human knowledge.
+Top performing chess engines, for instance, do not directly contain rules or strategies developed by humans and instead play millions of games to discover a successful approach.
+In the context of genomic variant discovery, approaches that enable machines to \textit{learn} the correct method for assembling reads are likely to yield better results.
 
@@
 \vspace{8pt}
+Sutton, R. "The Bitter Lesson", http://incompleteideas.net/IncIdeas/BitterLesson.html (2019).
+
+\vspace{8pt}
 Vaswani, Ashish, et al. "Attention is all you need." Advances in neural information processing systems 30 (2017).

### 2022-06-04 manuscript/manuscript.tex
manuscript updates

--- 2022-05-28 manuscript/manuscript.tex
@@
 \begin{abstract}
-	Detection of germline variants in next-generation sequencing data is an essential element of modern genomics.
+	Detection of germline variants in next-generation sequencing data is an important element of modern genomics.
 Variant detection tools typically rely on a variety of statistical algorithms and heuristic techniques to identify variants.
@@
 Next Generation Sequencing (NGS) workflows typically begin by aligning sequenced fragments to a reference genome, optionally performing some alignment post-processing steps such as identification of duplicated fragments, and then identifying sequence variants relative to the reference genome.
-The variant identification step is challenging for many reasons, including technical artifacts in the sequencing process, improper alignments due to sequence homology or low sequence complexity, and the stochastic nature of the library preparation and amplification steps (Li 2014).
+The variant identification step is challenging for many reasons, including technical artifacts in the sequencing process, improper alignments due to sequence homology or low sequence complexity, and the stochastic nature of the library preparation and amplification steps [citation here would be nice].
 
 All variant discovery tools must address two challenges.
-First, callers must identify potential  alleles in a region.
-Then, given a set of allele candidates, they must assess the likelihood of each candidate (or pairs of candidates in the diploid case), to determine which should be included in the caller output.
+First, they must identify candidate alleles in a region.
+Then, given a set of allele candidates, they must assess the likelihood of each candidate (or pairs of candidates in the diploid case), to determine the most likely alleles.
 Early callers, such as samtools / mpileup (Li et al.
 2009) and the UnifiedGenotyper tool from the Genome Analysis ToolKit (GATK, De Pristo et al.
-2011) rely on the read aligner to generate candidate allele candidates, and utilize ad-hoc heuristics and thresholds to determine likely alleles (Nielsen et al.
-2011).
+2011) rely on the read aligner to generate candidate allele candidates, and utilize ad-hoc heuristics and thresholds to determine likely alleles.
 Later tools (e.g.
@@
 2018, Cooke et al.
-2021) incorporated local re-assembly of reads in each region of interest, resulting in a significant improvement in the precision and specificity of variant calls.
+2021) incorporated local re-alignment of reads in each region of interest, resulting in a significant improvement in the precision and specificity of variant calls.
 For instance, the HaplotypeCaller tool identifies candidate haplotypes by constructing de Bruijn graphs from k-mers present in the reads, and assesses haplotype likelihoods with a pair Hidden Markov Model (HMM).
@@
 
-While replacing the statistical and heuristic components of the allele likelihood calculation with a deep learning model has resulted in improved specificity of variant callers, current tools continue to rely on complex, handcrafted statistical machinery to identify possible alleles.
-Here, we describe a new approach that replaces both the candidate allele generation and the likelihood calculation with a single, learned model.
+Maybe brief explanation of how transformers work?
+
+Here, we explore a new approach to variant detection that replaces the bespoke statistical techniques used in previous tools with a single, end-to-end machine learning model.
 We recast variant detection as a sequence-to-sequence translation problem in which the input sequence is the series of bases aligning to a single genomic location, and the output sequences are the two predicted haplotypes.
@@
 2017).
-We demonstrate that this approach identifies real genomic variants with sensitivity similar to current top-performing callers, albeit with lower specificity.
+We demonstrate that this approach identifies real genomic variants with sensitivity equivalent to current top-performing callers, albeit with slightly lower specificity.
 
@@
 
-\subsection{Model architecture}
+\subsection{Model architecture \& read encoding}
 
@@
 We add an additional fully connected layer prior to the transformer encoders which embeds the encoded basecalls in $d$ dimensions, where $d=12$ for the analyses here.
-The embedded basecalls are 'flattened' along the read dimension, producing an input token with size $dr$ where $r$ is the number of reads.
+The embedded basecalls are 'flattened' along the read dimension, producing an input token with size $dr$ where $r$ is the number of reads ($r=100$ for all experiments proposed here).
 Instead of the typical 1-dimensional positional encoding, we employed a 2-dimensional encoding, enabling the model to retain information about which read contained a given base across input tokens.
@@
 The decoder is of fixed size, and each predicted haplotype contains the same number of predicted bases as the number of positions $g$ in the input.
+
+Input tensors are generated by selecting a 150bp region from a set of aligned reads in BAM / CRAM format.
+For each region, all reads overlapping the window were obtained from the corresponding BAM/ CRAM file.
+Regions containing more than the maximum number of reads (100 for all analyses here) were downsampled to the maximum size.
+Reads are then sorted by the reference coordinate of the first aligned base.
+For each aligned base in the selected reads nine features are encoded; the first four were the one-hot encoded base call, followed by base quality, two flags indicating if base `consumed' reference base (i.e.
+was not an insertion) or consumed a sequence read base (was not a deletion), and additional flags indicating sequence read direction and clipping status.
+No 'gap' tokens or other special handling was performed for insertions or deletions.
+In addition to sequenced reads, the first row in each encoded region was the reference sequence.
+For this special row we inserted a base quality of 100 for every position and did not set any of the other flags.
+Resulting tensors had dimension $[g, r, 9]$, where $g$ represents genomic position, $r$ represents read, and 9 is the number of features.
+Positions in the input not corresponding to an aligned base were all set to zero.
 
@@
 
-Training tensors were generated by selecting regions 150bp in width from a list of preselected regions, in a manner described below.
-For each region, all reads overlapping the window were obtained from the corresponding BAM/ CRAM file.
-If more than a specified maximum number of reads were found (100 for all analyses here), then reads were downsampled to the desired size.
-Reads were sorted by the reference coordinate of the first aligned base.
-For each aligned base in the selected reads nine features were encoded; the first four were the one-hot encoded base call, followed by base quality, two flags indicating if base `consumed' reference base (i.e.
-was not an insertion) or consumed a sequence read base (was not a deletion), and additional flags indicating sequence read direction and clipping status.
-No 'gap' tokens or other special handling was performed for insertions or deletions.
-In addition to sequenced reads, the first row in each encoded region was the reference sequence.
-For this special row we inserted a base quality of 100 for every position and did not set any of the other flags.
-Resulting tensors had dimension $[150, 100, 9]$, where the indices are for the genomic position, read, and feature respectively.
-
-Target / label haplotype sequences were produced by obtaining truth variants from the Genome-in-a-Bottle VCF files for each sample and inserting the variants into the reference sequence.
-Two sequences were generated for each region, one representing each haplotype.
-In regions where phasing of the variants was ambiguous, the reads in the sequenced sample were examined to determine phase status.
-Briefly, all possible genotypes (pairs of haplotype sequences) were generated and reads were aligned via Smith-Waterman to the possible haplotypes, and the highest scoring genotypes selected as the most likely phasing.
-This phasing procedure procedure was only attempted for variants less than 100bp apart, otherwise the region was discarded.
-
 To select regions to include for training data, we developed a scheme to sample regions in a biased manner, prioritizing regions containing variants and, especially, regions with multiple or complex variants.
@@
 
-\subsection{Training procedure}
+Target / label haplotype sequences were produced by obtaining truth variants from the Genome-in-a-Bottle VCF files for each sample and inserting the variants into the reference sequence.
+Two sequences were generated for each region, one representing each haplotype.
+In regions where phasing of the variants was ambiguous, the reads in the sequenced sample were examined to determine phase status.
+Briefly, all possible genotypes (pairs of haplotype sequences) were generated and reads were aligned via Smith-Waterman to the possible haplotypes, and the highest scoring genotypes selected as the most likely phasing.
+This phasing procedure was only attempted for variants less than 100bp apart, otherwise the region was discarded.
+
+\subsection{Training procedure \& loss function}
 
@@
 
-\subsection{Loss function}
+We use a standard cross-entropy loss function to compare predicted haplotypes to known haplotype sequences for each region, and simply compare the prediction at window index (position) $i$ to the true base at index $i$, where $i \in 1..g$.
+While we obtain satisfactory results with this approach, two issues were apparent.
+First, predicted haplotype sequences may have inserted or deleted bases relative to the true haplotypes.
+Because incorrect insertions and deletions cause mismatches at every downstream (higher index) position, such mistakes are penalized more heavily when they occur at lower indexes than higher indexes.
+Thus the model learns to generate high quality results for the first portion of the window where inaccuracies are punished severely, and is less concerned with accuracy at high indexes.
 
-We use a standard cross-entropy loss function to compare predicted haplotypes to known haplotype sequences for each region, and simply compare each prediction at window index (position) $i$ to the true base at index $i$.
-A complexity arises due to ambiguity regarding which haplotype should contain a variant.
+A second complexity arises due to ambiguity regarding which haplotype should contain a variant.
 The two haplotypes present in a sample have no inherent order.
-In training data variants are assigned arbitrarily to haplotypes, and this randomness in haplotype assignment means the model cannot learn to predict which haplotype should contain a given variant.
+In training data variants are assigned arbitrarily to haplotypes, and this randomness in haplotype assignment means model cannot learn to predict which haplotype should contain a given variant.
 To combat this ambiguity, we calculate loss in both possible configurations - predicted haplotype 0 with training haplotype 0 and 1 with 1, as well as predicted haplotype 0 with training haplotype 1 and 1 with 0.
@@
 
-Given a region containing suspected variants, we perform multiple overlapping forward passes of the model with step size $k$, where $k=50$ for the results reported here.
+Given a region containing suspected variants, we perform multiple overlapping forward passes of the model with step size $k$.
 On each forward pass the model produces two predicted haplotypes by taking the base with the high predicted probability at each position.
 Each haplotype is aligned via Smith-Waterman to the reference sequence, and any mismatching positions are converted to variant calls.
-For each variant we record the number of windows in which each variant was detected, the number of variants in cis and trans, and the mean probability of the variant bases, and the position of the variant within each window.
+For each variant we record the number of windows in which each variant was detected, the number of variants in cis and trans, and the mean probability of the variant bases, and the distance of the variant from the window start coordinate.
 
@@
 In contrast to other variant detection methods (...), our approach does not rely on handcrafted statistical techniques such as de Bruijn graphs, hidden Markov models, or heuristic thresholding and instead learns to construct accurate haplotypes directly from aligned NGS reads.
+On short read WGS data, our approach demonstrates performance similar to current top-performing callers, with somewhat higher sensitivity but lower precision for insertion/deletion variants ()
 
-Our analysis demonstrates that the model learns many features of modern, state of the art variant callers.
+Our analysis demonstrates that the model learns many capabilities of modern, state of the art variant callers.
 For instance, our model learns to reconstruct large, complex variants in the presence of ambiguous mapping and extensive soft clipping.
+While our model's sensitivity is ~5\% less than HaplotypeCaller for variants in these complex regions, we nonetheless are able to reconstruct the correct haplotypes in more than 90\% of cases.
 In addition, the model learns to employ local context and read-specific features to predict phase and genotype accurately, and pairs of variants are phased correctly in ~99\% of cases where variants are within 100bp.
 
-In this work we have used a standard cross-entropy loss function that simply compares each prediction at base $i$ to the label at base $i$.
-While training with this loss yields accurate results, it has two significant shortcomings.
-First, the impact of an insertion / deletion (indel) error depends on its position in the region, with indels at lower indices leading to a greater number of mismatches and high loss compared to indels at higher indices.
-Secondly, indel errors ar punished more severely than similar SNV errors.
-We hypothesize the second fact explains the lower performance of our model on SNV calls compared to indels.
-Alternative, alignment-based loss functions such as those described by Baid et al.
-(2021) and Petti et al.
-(2021) may provide a path forward, however the Petti implementation lead to lower sensitivity in our experiments.
+All models in this work were trained using a standard cross-entropy loss function.
+Cross-entropy loss is probably not optimal for variant detection, because insertion and deletion errors are more severely punished than SNVs, and the loss has a strong dependence on the relative position of the variant in the input sequence.
+During development we experimented with the differentiable Smith-Waterman loss function as described by Petti et al.
+(2021), but this lead to reduced sensitivity in our experiments.
 
@@
 \vspace{8pt}
-Li, Heng. "Toward better understanding of artifacts in variant calling from high-coverage samples." Bioinformatics 30.20 (2014): 2843-2851.
-
-\vspace{8pt}
-Li, Heng, et al. "The sequence alignment/map format and SAMtools." Bioinformatics 25.16 (2009): 
-2078-2079.
+Li, Heng, et al. "The sequence alignment/map format and SAMtools." Bioinformatics 25.16 (2009): 2078-2079.
 
@@
 Marco-Sola S., Sammeth M., Guigó R., Ribeca P. "The GEM mapper: fast, accurate and versatile alignment by filtration". Nat Methods. (2012);9(12):1185-1188. doi:10.1038/nmeth.2221
-
-\vspace{8pt}
-Nielsen, Rasmus, et al. "Genotype and SNP calling from next-generation sequencing data." Nature Reviews Genetics 12.6 (2011): 443-451.
 

### 2022-06-04 manuscript/manuscript.tex

--- 2022-06-04 manuscript/manuscript.tex
@@
 
-\subsection{Loss function}
-
-We use a standard cross-entropy loss function to compare predicted haplotypes to known haplotype sequences for each region, and simply compare each prediction at window index (position) $i$ to the true base at index $i$.
-A complexity arises due to ambiguity regarding which haplotype should contain a variant.
-The two haplotypes present in a sample have no inherent order.
-In training data variants are assigned arbitrarily to haplotypes, and this randomness in haplotype assignment means the model cannot learn to predict which haplotype should contain a given variant.
-To combat this ambiguity, we calculate loss in both possible configurations - predicted haplotype 0 with training haplotype 0 and 1 with 1, as well as predicted haplotype 0 with training haplotype 1 and 1 with 0.
-We select the configuration with the lowest loss and backpropogate gradients only for that configuration.
+We use a standard cross-entropy loss function to compare predicted haplotypes to known haplotype sequences for each region, and simply compare the prediction at window index (position) $i$ to the true base at index $i$, where $i \in 1..g$.
+While we obtain satisfactory results with this approach, two issues were apparent.
+First, predicted haplotype sequences may have inserted or deleted bases relative to the true haplotypes.
+Because incorrect insertions and deletions cause mismatches at every downstream (higher index) position, such mistakes are penalized more heavily when they occur at lower indexes than higher indexes.
+Thus the model learns to generate high quality results for the first portion of the window where inaccuracies are punished severely, and is less concerned with accuracy at high indexes.
 

### 2022-06-08 manuscript/manuscript.tex
manuscript updates

@@
 
+\begin{figure}[htp]
+	\includegraphics[scale=0.5]{./jovian_architecture_diagram.png}
+	\caption{ Overview of network architecture, numbers in parentheses indicate size of tensors  }
+	\label{fig:architecture}
+\end{figure}
+
 \subsection{NGS data}
@@
 
-Models were implemented in Pytorch (v1.10, Paszke et al 2019). Models were trained for 25 epochs unless noted otherwise, using the Adam optimizer and a learning rate of $5e-5$.
+Models were implemented in Pytorch (v1.10, Paszke et al 2019). Models were trained for 25 epochs unless noted otherwise, using the Adam optimizer and a learning rate of $5e-5$ and batch size of 512.
 
-We use a standard cross-entropy loss function to compare predicted haplotypes to known haplotype sequences for each region, and simply compare the prediction at window index (position) $i$ to the true base at index $i$, where $i \in 1..g$.
-While we obtain satisfactory results with this approach, two issues were apparent.
-First, predicted haplotype sequences may have inserted or deleted bases relative to the true haplotypes.
-Because incorrect insertions and deletions cause mismatches at every downstream (higher index) position, such mistakes are penalized more heavily when they occur at lower indexes than higher indexes.
-Thus the model learns to generate high quality results for the first portion of the window where inaccuracies are punished severely, and is less concerned with accuracy at high indexes.
+For a given haplotype we use a standard cross-entropy loss function to compare the prediction to the target sequence.
+However, two haplotypes exist for each region and they have no inherent order.
+For instance, a single heterozygous variant can appear on either haplotype, and the model cannot learn predict which because the assigment is random.
+To make the predictions opaque to haplotype order, we compute the loss in both haplotype configurations - predicted haplotype 0 with target haplotype 0 and 1 with 1, as well as predicted haplotype 1 with target haplotype 0, and 0 with 1, and backpropogate gradients only for the configuration with the lowest loss.
 

### 2022-06-10 dnaseq2seq/call.py
10m params for calling

@@ -116,9 +116,9 @@ def reconcile_current_window(prev_win, current_win):
 
 
 def load_model(model_path):
-    attention_heads = 8
-    encoder_layers = 8
-    transformer_dim = 400
+    attention_heads = 6
+    encoder_layers = 6
+    transformer_dim = 250
     embed_dim_factor = 100
     model = VarTransformer(read_depth=100,
                            feature_count=9,

### 2022-06-10 manuscript/manuscript.tex
Working prototype

--- 2022-06-04 manuscript/manuscript.tex
@@
 \begin{abstract}
-	Detection of germline variants in next-generation sequencing data is an important element of modern genomics.
+	Detection of germline variants in next-generation sequencing data is an essential element of modern genomics.
 Variant detection tools typically rely on a variety of statistical algorithms and heuristic techniques to identify variants.
@@
 
-While transformers have not been explored for variant detection, a model similar to ours was proposed by Baid et al.
-(2021) for generation of accurate consensus reads from PacBio data.
-The Baid model (`DeepConsensus') used a network architecture similar to ours, but included an extra token representing a gap between the aligned sequences, and a novel loss function that accounted for gaps between the predicted and true sequence in a region.
-In contrast we use a simple cross-entropy loss function, and do not introduce gaps in the aligned sequences compared to the reference.
-
 \section{Methods}
 
-\subsection{Model architecture \& read encoding}
+\subsection{Model architecture}
 
@@
 We add an additional fully connected layer prior to the transformer encoders which embeds the encoded basecalls in $d$ dimensions, where $d=12$ for the analyses here.
-The embedded basecalls are 'flattened' along the read dimension, producing an input token with size $dr$ where $r$ is the number of reads ($r=100$ for all experiments proposed here).
+The embedded basecalls are 'flattened' along the read dimension, producing an input token with size $dr$ where $r$ is the number of reads.
 Instead of the typical 1-dimensional positional encoding, we employed a 2-dimensional encoding, enabling the model to retain information about which read contained a given base across input tokens.
@@
 The decoder consists of a single fully-connected layer followed by two additional fully connected components that each generate a single haplotype.
-Predictions are discrete probability distributions over the four possible bases at each predicted output position, produced with a softmax function over the raw outputs.
+Predictions are discrete probability distributions over the four possible bases at each predicted output position.
 The decoder is of fixed size, and each predicted haplotype contains the same number of predicted bases as the number of positions $g$ in the input.
-
-Input tensors are generated by selecting a 150bp region from a set of aligned reads in BAM / CRAM format.
-For each region, all reads overlapping the window were obtained from the corresponding BAM/ CRAM file.
-Regions containing more than the maximum number of reads (100 for all analyses here) were downsampled to the maximum size.
-Reads are then sorted by the reference coordinate of the first aligned base.
-For each aligned base in the selected reads nine features are encoded; the first four were the one-hot encoded base call, followed by base quality, two flags indicating if base `consumed' reference base (i.e.
-was not an insertion) or consumed a sequence read base (was not a deletion), and additional flags indicating sequence read direction and clipping status.
-No 'gap' tokens or other special handling was performed for insertions or deletions.
-In addition to sequenced reads, the first row in each encoded region was the reference sequence.
-For this special row we inserted a base quality of 100 for every position and did not set any of the other flags.
-Resulting tensors had dimension $[g, r, 9]$, where $g$ represents genomic position, $r$ represents read, and 9 is the number of features.
-Positions in the input not corresponding to an aligned base were all set to zero.
 
@@
 
-To select regions to include for training data, we developed a scheme to sample regions in a biased manner, prioritizing regions containing variants and, especially, regions with multiple or complex variants.
-Regions of the reference genome overlapping the high-confidence regions from Genome-in-a-Bottle were subdivided into 150bp nonoverlapping windows, and these regions labelled according to the presence of variants.
-Separate labels were generated for regions conataining a single SNV, deletion, or insertion, as well as regions containing multiple insertion-deletions, or those containing variants intersecting low-complexity regions.
-We additionally included `true negative' regions where no known variant was present.
-For all analyses described here, we obtain training data only from the human autosomes 1-20, and hold out chromosomes 21 and 22 for model evaluation.
-We examine three different sizes of training data sets, with 5.5M, 10.4M, and 20.4M examples.
+Training tensors were generated by selecting regions 150bp in width from a list of preselected regions, in a manner described below.
+For each region, all reads overlapping the window were obtained from the corresponding BAM/ CRAM file.
+If more than a specified maximum number of reads were found (100 for all analyses here), then reads were downsampled to the desired size.
+Reads were sorted by the reference coordinate of the first aligned base.
+For each aligned base in the selected reads nine features were encoded; the first four were the one-hot encoded base call, followed by base quality, two flags indicating if base `consumed' reference base (i.e.
+was not an insertion) or consumed a sequence read base (was not a deletion), and additional flags indicating sequence read direction and clipping status.
+No 'gap' tokens or other special handling was performed for insertions or deletions.
+In addition to sequenced reads, the first row in each encoded region was the reference sequence.
+For this special row we inserted a base quality of 100 for every position and did not set any of the other flags.
+Resulting tensors had dimension $[g, r, 9]$, where $g$ represents genomic position, $r$ represents read, and 9 is the number of features.
 
@@
 Briefly, all possible genotypes (pairs of haplotype sequences) were generated and reads were aligned via Smith-Waterman to the possible haplotypes, and the highest scoring genotypes selected as the most likely phasing.
-This phasing procedure was only attempted for variants less than 100bp apart, otherwise the region was discarded.
+This phasing procedure procedure was only attempted for variants less than 100bp apart, otherwise the region was discarded.
 
-\subsection{Training procedure \& loss function}
+To select regions to include for training data, we developed a scheme to sample regions in a biased manner, prioritizing regions containing variants and, especially, regions with multiple or complex variants.
+Regions of the reference genome overlapping the high-confidence regions from Genome-in-a-Bottle were subdivided into 150bp nonoverlapping windows, and these regions labelled according to the presence of variants.
+Separate labels were generated for regions conataining a single SNV, deletion, or insertion, as well as regions containing multiple insertion-deletions, or those containing variants intersecting  low-complexity regions.
+We additionally included `true negative' regions where no known variant was present.
+The total number of regions generated per sample was 325,000, yielding 5.53M total training regions across our 17 samples.
 
-Models were implemented in Pytorch (v1.10, Paszke et al 2019). Models were trained for 25 epochs unless noted otherwise, using the Adam optimizer and a learning rate of $5e-5$.
+For all analyses described here, we obtain training data only from the human autosomes 1-20, and hold out chromosomes 21 and 22 for model evaluation.
 
-We use a standard cross-entropy loss function to compare predicted haplotypes to known haplotype sequences for each region, and simply compare the prediction at window index (position) $i$ to the true base at index $i$, where $i \in 1..g$.
+\subsection{Loss function}
+
+We use a standard cross-entropy loss function to compare predicted haplotypes to known haplotype sequences for each region, and simply compare each prediction at window index (position) $i$ to the true base at index $i$.
 While we obtain satisfactory results with this approach, two issues were apparent.
@@
 Because incorrect insertions and deletions cause mismatches at every downstream (higher index) position, such mistakes are penalized more heavily when they occur at lower indexes than higher indexes.
-Thus the model learns to generate high quality results for the first portion of the window where inaccuracies are punished severely, and is less concerned with accuracy at high indexes.
+Thus the model learns to generate high quality results for the first portion of the window where inaccuracies are punished severely, and is less concerned with accuarcy at high indexes.
 
@@
 
-Given an alignment file in BAM or CRAM format, we first identify regions where a potential variant might exist.
+Given an alignment file in BAM or CRAM format, we first identify regions where a potential variant might exist, and then run forward passes of the model using a sliding window over the region.
 Any genomic position in which at least three reads contain a base that differs from the reference or an indel are flagged as potentially containing a variant.
@@
 Each haplotype is aligned via Smith-Waterman to the reference sequence, and any mismatching positions are converted to variant calls.
-For each variant we record the number of windows in which each variant was detected, the number of variants in cis and trans, and the mean probability of the variant bases, and the distance of the variant from the window start coordinate.
+For each variant we record the number of windows in which each variant was detected, the number of variants in cis and trans, and the mean probability of the variant bases, and the position of the variant within each window.
 
-\subsection{Variant quality calculation}
+\subsection{Variant quality calculation and false positive removal}
 
 The above variant detection procedure has two shortcomings.
-First, if every potential variant is emitted, then specificity is poor.
-Second, the model does not innately produce well calibrated variant quality scores.
+First, if every detected variant is emitted, then specificity is poor.
+Second, variant quality scores are poorly calibrated and do not closely reflect the probability that a variant is real.
 To address these shortcomings we introduce a post-hoc random forest classifier that we train to discriminate true and false-positive variant calls (see table X for features).
-The resulting score is used as a final variant quality score suitable for tuning the tradeoff between sensitivity and specificity.
-
-The random forest classifier is trained by calling variants on selected regions from autosomes 1-4 totalling approximately 120Mb.
-True and false positive variant calls were identified with the \textit{vcfeval} tool (Cleary et al.
-2015).
-We used the scikit-learn implementation of the random forest with 100 trees, a maximum tree depth of 25, and 'balanced' class weighting option.
+The resulting score is used as a final variant quality score that can be used for tuning the tradeoff between sensitivity and specificity.
 
@@
 \subsection{Complex regions}
- While many genomic variants are trivial to identify by examination of mapped reads, some regions require a more sophisticated reassembly to identify the true haplotype.
-We examine the ability of our model to perform this more complicated reconstruction by identifying real variants in regions containing multiple conflicting read mappings or extensive read soft-clipping.
-
- To identify variants requiring complex reconstruction, we first obtain all indel variants from Genome-in-a-Bottle greater than 4 bp from the validation regions on chromosomes 21 and 22.
-For each variant, we examine the reads in the alignment file that overlap the variant position, and flag any with with more than 4 distinct indel start sites or at least 20\% reads with more than 10bp softclipped as requiring 'complex' reconstruction.
-These variants are moved to a separate truth set.
-After running our variant calling procedure on the alignment file, we compute the fraction of variants from the truth set that were identified by our model.
-This procedure was repeated for all 17 samples in our data set.
-
- Our model is able to reconstruct the correct haplotypes for ~90.8\% (mid 50\% 86.4-95.2) of these variants, somewhat lower than HaplotypeCaller accuracy of ~95.8\% (mid 50\% 91.0-100.0).
-Although model performance does not match current state-of-the-art callers, it nonetheless is able to reconstruct the true haplotype in most cases.
-Overall, these results indicate that the model learns a reconstruction procedure more nuanced than a simple column-by-column examination of reads,
+ While many genomic variants are trivial to identify by examination of mapped reads, some regions require a more sophisticated reassembly of reads to identify the true haplotype.
+We examine the ability of our model to perform this more complicated reconstruction by identifying real variants in regions containing multiple conflicting read mappings in addition to extensive read soft-clipping.
+ To identify such regions we first obtain all indel variants from Genome-in-a-Bottle greater than X bp.
+We then examine the reads in the alignment file and flag any region that contains more than Y distinct indel start sites or at least Z reads with more than 10bp softclipped.
+We then
 
@@
 In contrast to other variant detection methods (...), our approach does not rely on handcrafted statistical techniques such as de Bruijn graphs, hidden Markov models, or heuristic thresholding and instead learns to construct accurate haplotypes directly from aligned NGS reads.
-On short read WGS data, our approach demonstrates performance similar to current top-performing callers, with somewhat higher sensitivity but lower precision for insertion/deletion variants ()
 
-Our analysis demonstrates that the model learns many capabilities of modern, state of the art variant callers.
+Our analysis demonstrates that the model learns many features of modern, state of the art variant callers.
 For instance, our model learns to reconstruct large, complex variants in the presence of ambiguous mapping and extensive soft clipping.
-While our model's sensitivity is ~5\% less than HaplotypeCaller for variants in these complex regions, we nonetheless are able to reconstruct the correct haplotypes in more than 90\% of cases.
-In addition, the model learns to employ local context and read-specific features to predict phase and genotype accurately, and pairs of variants are phased correctly in ~99\% of cases where variants are within 100bp.
+In addition, the model learns to employ local context and read-specific features to predict phase
 
-All models in this work were trained using a standard cross-entropy loss function.
-Cross-entropy loss is probably not optimal for variant detection, because insertion and deletion errors are more severely punished than SNVs, and the loss has a strong dependence on the relative position of the variant in the input sequence.
-During development we experimented with the differentiable Smith-Waterman loss function as described by Petti et al.
-(2021), but this lead to reduced sensitivity in our experiments.
+ Loss functions, there's probably something better. The Smith Waterman loss sounds really cool, but didn't lead to high sensitivity in our hands.
 
-For this work we have employed an unmodified transformer model with a very simple decoder consisting of just two fully connected layers, and we leave to future work exploration of the many transformer variations proposed in recent years (e.g.
-Dosovitskiy 2018, Fedus et al.
-2021, Liu et al 2021, Wu et al.
-2021).
-Our simple decoder only predicts haplotypes that are equal in length to the number of input tokens (genomic positions), but a natural improvement would be to add a more sophisticated decoder capable of producing a variable length sequence to handle larger insertion and deletion events.
-In addition, newer transformer models that avoid the $O(n^2)$ scaling with the number of input tokens may be able to handle longer input sequence lengths, and therefore may be able to detect larger variants.
+For this work we have employed an unmodified transformer model with a very simple decoder consisting of just two fully connected layers, and we leave to future work the exploration of the many transformer variations proposed in recent years (vision transformers, swin transformers, fast-former, etc.)
+
+Limitations
+ - Low SNV precision
+ - not using strand bias?
+ -
 
@@
 \vspace{8pt}
-Baid, Gunjan, et al. "DeepConsensus: Gap-Aware Sequence Transformers for Sequence Correction." bioRxiv (2021).
-
-\vspace{8pt}
-Cleary, John G., et al. "Comparing variant call files for performance benchmarking of next-generation sequencing variant calling pipelines." BioRxiv (2015): 023754.
-
-\vspace{8pt}
 Cooke, Daniel P., David C. Wedge, and Gerton Lunter. "A unified haplotype-based method for accurate and comprehensive variant calling." Nature biotechnology 39.7 (2021): 885-892.
@@
 DePristo, Mark A., et al. "A framework for variation discovery and genotyping using next-generation DNA sequencing data." Nature genetics 43.5 (2011): 491-498.
-
-\vspace{8pt}
-Dosovitskiy, Alexey, et al. "An image is worth 16x16 words: Transformers for image recognition at scale." arXiv preprint arXiv:2010.11929 (2020).
-
-\vspace{8pt}
-Fedus, William, Barret Zoph, and Noam Shazeer. "Switch transformers: Scaling to trillion parameter models with simple and efficient sparsity." arXiv preprint arXiv:2101.03961 (2021).
 
@@
 \vspace{8pt}
-Liu, Ze, et al. "Swin transformer: Hierarchical vision transformer using shifted windows." Proceedings of the IEEE/CVF International Conference on Computer Vision. 2021.
+Marco-Sola S., Sammeth M., Guigó R., Ribeca P. "The GEM mapper: fast, accurate and versatile alignment by filtration". Nat Methods. (2012);9(12):1185-1188. doi:10.1038/nmeth.2221
 
 \vspace{8pt}
-Marco-Sola S., Sammeth M., Guigó R., Ribeca P. "The GEM mapper: fast, accurate and versatile alignment by filtration". Nat Methods. (2012);9(12):1185-1188. doi:10.1038/nmeth.2221
+Nielsen, Rasmus, et al. "Genotype and SNP calling from next-generation sequencing data." Nature Reviews Genetics 12.6 (2011): 443-451.
 
@@
 Retrieved from http://papers.neurips.cc/paper/9015-pytorch-an-imperative-style-high-performance-deep-learning-library.pdf
-
-\vspace{8pt}
-Petti, Samantha, et al. "End-to-end learning of multiple sequence alignments with differentiable Smith-Waterman." BioRxiv (2021).
 
@@
 
-\vspace{8pt}
-Wu, Chuhan, et al. "Fastformer: Additive attention can be all you need." arXiv preprint arXiv:2108.09084 (2021).
-
 \end{document}

### 2022-06-10 dnaseq2seq/bin/main.py

@@ -297,7 +297,10 @@ def main():
     args = parser.parse_args()
     args.cmdline = " ".join(sys.argv[1:])
     args.cl_args = vars(args).copy()  # command line copy for logging
-    args.func(**vars(args))
+    if len(vars(args)) == 0 or getattr(args, 'func') is None:
+        print(args.usage())
+    else:
+        args.func(**vars(args))
 
 
 if __name__ == "__main__":

### 2022-06-10 dnaseq2seq/call.py

@@ -1,3 +1,4 @@
+import itertools
 import os
 import time
 import math
@@ -717,4 +718,69 @@ def _call_vars_region(
 
     logger.debug(f"{cpname}: Enc time total: {enctime_total.total_seconds()}  calltime total: {calltime_total.total_seconds()}")
     return chrom, window_idx, hap0_passing, hap1_passing
+
+
+
+
+def vars_dont_overlap(v0, v1):
+    return v0.pos + len(v0.ref) < v1.pos or v1.pos + len(v1.ref) < v0.pos
+
+
+def merge_genotypes(genos):
+    """
+    Genos is a list of 2-tuples representing haplotypes called in overlapping windows. This function
+    attempts to merge the haplotypes in a way that makes sense, given the possibility of conflicting information
+
+    """
+    allvars = sorted(list(v for g in genos for v in g[0]) + list(v for g in genos for v in g[1]), key=lambda v: v.pos)
+    allkeys = set()
+    varsnodups = []
+    for v in allvars:
+        if v.key not in allkeys:
+            allkeys.add(v.key)
+            varsnodups.append(v)
+
+    results = [[], []]
+
+    prev_het = None
+    prev_het_index = None
+    for p in varsnodups:
+        homcount = 0
+        hetcount = 0
+        for g in genos:
+            a = p.key in [v.key for v in g[0]]
+            b = p.key in [v.key for v in g[1]]
+            if a and b:
+                homcount += 1
+            elif a or b:
+                hetcount += 1
+        if homcount > hetcount:
+            results[0].append(p)
+            results[1].append(p)
+        elif prev_het is None:
+            results[0].append(p)
+            prev_het = p
+            prev_het_index = 0
+        else:
+            # There was a previous het variant, so figure out where this new one should go
+            # determine if p should be in cis or trans with prev_het
+            cis = 0
+            trans = 0
+            for g in genos:
+                g0keys = [v.key for v in g[0]]
+                g1keys = [v.key for v in g[1]]
+                if (p.key in g0keys and prev_het.key in g0keys) or (p.key in g1keys and prev_het.key in g1keys):
+                    cis += 1
+                elif (p.key in g0keys and prev_het.key in g1keys) or (p.key in g1keys and prev_het.key in g0keys):
+                    trans += 1
+            if trans >= cis: # If there's a tie, assume trans. This covers the case where cis==0 and trans==0, because trans is safer
+                results[1 - prev_het_index].append(p)
+                prev_het = p
[…]

### 2022-06-10 dnaseq2seq/test/test_call.py

@@ -1,7 +1,7 @@
 
 import pytest
 
-from dnaseq2seq import call
+from dnaseq2seq import call, vcf
 
 @pytest.mark.parametrize(["n_items", "n_chunks"],
                          [(1,1), (5,1), (7,2)])
@@ -14,6 +14,109 @@ def test_split_chunks(n_items, n_chunks):
     for start, end in results:
         assert start == prev_end
         prev_end = end
+
+
+def test_gen_combos():
+    for n in range(2, 5):
+        allresults = list(p for p in call.gen_combos(n))
+        assert len(allresults) == len(set(allresults)), "Repeated entries in permutation results"
+        assert len(allresults) == 2**n, "Incorrect number of permutations"
+
+
+def test_gen_idx():
+    for c in call.gen_idx(3):
+        print(c)
+
+def test_score_conf():
+    gconf = [
+        ([vcf.Variant(pos=10, ref='A', alt='T', qual=1), vcf.Variant(pos=20, ref='G', alt='C', qual=1)],
+         []),
+        ([vcf.Variant(pos=10, ref='A', alt='T', qual=1), vcf.Variant(pos=20, ref='G', alt='C', qual=1)],
+         []),
+        ([],
+         [vcf.Variant(pos=10, ref='A', alt='T', qual=1), vcf.Variant(pos=20, ref='G', alt='C', qual=1)]),
+    ]
+
+    gconf2 = [
+        ([vcf.Variant(pos=10, ref='A', alt='T', qual=1), vcf.Variant(pos=20, ref='G', alt='C', qual=1)],
+         []),
+        ([vcf.Variant(pos=10, ref='A', alt='T', qual=1), vcf.Variant(pos=20, ref='G', alt='C', qual=1)],
+         []),
+        ([vcf.Variant(pos=10, ref='A', alt='T', qual=1), vcf.Variant(pos=20, ref='G', alt='C', qual=1)],
+         []),
+    ]
+    score = call.score_conf(gconf)
+    score2 = call.score_conf(gconf2)
+
+    assert score2 < score
+
+
+def test_merge_genotypes():
+    # gconf = [
[…]

### 2022-06-10 dnaseq2seq/vcf.py

@@ -24,6 +24,11 @@ class Variant:
     def __gt__(self, other):
         return self.pos > other.pos
 
+    @property
+    def key(self):
+        return f"{self.pos}-{self.ref}-{self.alt}"
+
+
 @dataclass
 class Cigar:
     op: str

### 2022-06-11 dnaseq2seq/buildclf.py
Turn off strandbiars feat

@@ -250,7 +250,7 @@ def var_feats(var, aln, var_freq_file):
     feats.append(vaf)
     feats.append(mq_vaf)
     feats.append(pos_alt + neg_alt)
-    feats.append(strandbias_stat)
+    #feats.append(strandbias_stat)
     return np.array(feats)
 
 

### 2022-06-14 dnaseq2seq/buildclf.py
Remove stepcount vcf entry, fix log messages

@@ -310,13 +310,14 @@ def train_model(conf, threads, var_freq_file, feat_csv=None, labels_csv=None, re
     allfps = []
     var_freq_file = pysam.VariantFile(var_freq_file)
     if feat_csv:
-        logger.info("Writing feature dump to {feat_csv}")
+        logger.info(f"Writing feature dump to {feat_csv}")
         feat_fh = open(feat_csv, "w")
         feat_fh.write("varstr," + ",".join(feat_names()) + "\n")
     else:
         feat_fh = None
 
     for sample in conf.keys():
+        logger.info(f"Processing sample {sample}")
         aln = pysam.AlignmentFile(conf[sample]['bam'], reference_filename=reference_filename)
         tps = conf[sample].get('tps')
         if type(tps) == str:
@@ -330,6 +331,7 @@ def train_model(conf, threads, var_freq_file, feat_csv=None, labels_csv=None, re
         elif type(fps) == list:
             for fp in fps:
                 allfps.extend(extract_feats(fp, aln, var_freq_file, feat_fh))
+        logger.info(f"TPs: {len(alltps)} FPs: {len(allfps)}")
 
     if feat_fh:
         feat_fh.close()

### 2022-06-14 dnaseq2seq/vcf.py

@@ -306,7 +306,6 @@ def vcf_vars(vars_hap0, vars_hap1, chrom, window_idx, aln, reference, mindepth=3
         vcfvars_hap0[var].window_cis_vars += vcfvars_hap0[var].window_trans_vars  # cis and trans are now cis
         vcfvars_hap0[var].window_trans_vars = 0  # moved all vars to cis
         vcfvars_hap0[var].call_count += vcfvars_hap1[var].call_count  # combine call counts in both haplotypes
-        vcfvars_hap0[var].step_count = len(set(vcfvars_hap0[var].step_indexes))  # combine call counts in both haplotypes
         vcfvars_hap0[var].genotype = (1, 1)
         vcfvars_hap0[var].het = False
         vcfvars_hap0[var].window_offset = sorted(set(vcfvars_hap0[var].window_offset + vcfvars_hap1[var].window_offset))

### 2022-06-15 manuscript/manuscript.tex
Manuscript updates

--- 2022-06-04 manuscript/manuscript.tex
@@
 Our model frames variant detection as a sequence-to-sequence modeling task, and employs a transformer-based architecture to translate alignment columns into two predicted haplotype sequences.
-We demonstrate that this method predicts variant genotypes accurately, phases nearby variants correctly, and has the ability to realign complex and ambiguous read mappings to produce accurate haplotype predictions.
+We demonstrate that this method predicts variant genotypes and phase accurately, has the ability to realign complex and ambiguous read mappings to produce accurate haplotype predictions, and uses the local surrounding read context to modify predictions about a given position.
 
@@
 
-Accurate detection of DNA sequence variants in a sample from is a fundamental component of modern genomics analysis. Tools that aim to
+Accurate inference of the DNA sequence variants in a sample is a fundamental component of modern genomics analysis.
+Tools that aim to identify variants must overcome several challenges to produce an accurate set of variant calls.
+For instance, the sequenced reads themselves may contain errors in the form of miscalled bases or PCR-induced repeat length errors, or the tool used to align reads to the reference genome may produce incorrect mappings (e.g.
+Li et al.
+2014).
 
@@
 Given a region containing suspected variants, we perform multiple overlapping forward passes of the model with step size $k$, where $k=50$ for the results reported here.
-On each forward pass the model produces two predicted haplotypes by taking the base with the high predicted probability at each position.
+On each forward pass the model predicts two haplotypes by taking the base with the high predicted probability at each position.
 Each haplotype is aligned via Smith-Waterman to the reference sequence, and any mismatching positions are converted to variant calls.
+After variants are collected across multiple windows, results are merged using a method that attempts to minimize the number of conflicting calls between windows.
 For each variant we record the number of windows in which each variant was detected, the number of variants in cis and trans, and the mean probability of the variant bases, and the position of the variant within each window.
@@
 
+\subsection{Variant detection accuracy}
+
+To characterize variant detection accuracy, we called variants on validation regions from chromosomes 21 and 22 for each of the 17 samples, and compared results to DeepVariant (v1.3.0, Poplin et al.
+2018), HaplotypeCaller (v.
+4.1.4.1), and Strelka2 (v.
+2.9.10, Kim et al.
+2018).
+Calls were obtained with the 10M model trained for 25 epochs on the 'bigger' training set (~$10^6$ training examples).
+We selected filtering criteria for each caller by reviewing ROC curves produced by \textit{vcfeval} and selecting values close to the $F_1$ maximizing value across samples.
+Jovian calls were filtered at 0.4, HaplotypeCaller at 60 (phred-scaled), Strelka at 4 (phred-scaled), and DeepVariant at 0.
+Sensitivity, precision, and related calculations were computed with the \textit{hap.py} tool (Krushe et al.
+2019) using Genome-in-a-Bottle (v.4.2.1) as the benchmark variant set.
+
+Across all regions, Jovian's sensitivity and precision for indels is slightly higher than that for HaplotypeCaller, and slightly lower than DeepVariant or Strelka2.
+For SNV calls Jovian has similar sensitivity to other callers but the lowest precision of the callers ()
+
 \subsection{Complex regions}
@@
 
-\subsection{Variant detection accuracy}
+\subsection{Ablation of 2D positional encoding}
 
-To characterize variant detection accuracy, we called variants on validation regions from chromosomes 21 and 22 for each of the 17 samples, and compared results to DeepVariant (v1.3.0, Poplin et al.
-2018), HaplotypeCaller (v.
-4.1.4.1), and Strelka2 (v.
-2.9.10, Kim et al.
-2018).
-Calls were obtained with the 10M model trained for 25 epochs on the 'bigger' training set (~$10^6$ training examples).
-We selected filtering criteria for each caller by reviewing ROC curves produced by \textit{vcfeval} and selecting values close to the $F_1$ maximizing value across samples.
-Jovian calls were filtered at 0.4, HaplotypeCaller at 60 (phred-scaled), Strelka at 4 (phred-scaled), and DeepVariant at 0.
-Sensitivity, precision, and related calculations were computed with the \textit{hap.py} tool (Krushe et al.
-2019) using Genome-in-a-Bottle (v.4.2.1) as the benchmark variant set.
-
-Across all regions, Jovian's sensitivity and precision for indels is slightly higher than that for HaplotypeCaller, and slightly lower than DeepVariant or Strelka2.
-For SNV calls Jovian has similar sensitivity to other callers but the lowest precision of the callers ()
+To assess the impact of the two-dimensional positional encoding, we replaced the 2D encoding module with traditional 1D encoding, and left other model parameters unmodified.
+We conducted two training runs, one with the 10M model and smaller training data size, and the other with the 25M model and medium training data size.
+Both training runs were conducted for 25 epochs.
+When compared to their equivalent models with 2D encoding, the 1D runs were significantly compromised.
+In particular, precision for most variant classes was markedly lower than the 2D runs, and sensitivity was also reduced but to a lesser degree.
 
@@
 We describe a new approach to the problem of detecting sequence variants in next-generation sequencing (NGS).
-Our approach envisions variant detection as a sequence-to-sequence modeling problem, akin to language translation, and leverages the successful transformer architecture to accomplish this task.
+Our approach envisions variant detection as a sequence-to-sequence modeling problem, akin to language translation, and leverages the successful transformer architecture to perform the translation.
 The sequence-to-sequence approach allows a single deep learning model to be used for candidate allele generation and haplotype construction.

### 2022-06-21 dnaseq2seq/call.py
10m call params

@@ -119,16 +119,16 @@ def reconcile_current_window(prev_win, current_win):
 def load_model(model_path):
     
     # 25m
-    attention_heads = 8
-    encoder_layers = 8
-    transformer_dim = 400
-    embed_dim_factor = 100
+    #attention_heads = 8
+    #encoder_layers = 8
+    #transformer_dim = 400
+    #embed_dim_factor = 100
     
     # 10m
-    #attention_heads = 6
-    #encoder_layers = 6
-    #transformer_dim = 250
-    #embed_dim_factor = 100
+    attention_heads = 6
+    encoder_layers = 6
+    transformer_dim = 250
+    embed_dim_factor = 100
     model = VarTransformer(read_depth=100,
                            feature_count=9,
                            out_dim=4,

### 2022-06-22 manuscript/manuscript.tex
Manuscript updates

@@
 \begin{abstract}
-	Detection of germline variants in next-generation sequencing data is an important element of modern genomics.
+	Detection of germline variants in next-generation sequencing data is an essential element of modern genomics.
 Variant detection tools typically rely on a variety of statistical algorithms and heuristic techniques to identify variants.
-Here we describe a new approach that replaces the handcrafted statistical methods with a single, end-to-end deep learning model.
-Our model frames variant detection as a sequence-to-sequence modeling task, and employs a transformer-based architecture to translate alignment columns into two predicted haplotype sequences.
-We demonstrate that this method predicts variant genotypes and phase accurately, has the ability to realign complex and ambiguous read mappings to produce accurate haplotype predictions, and uses the local surrounding read context to modify predictions about a given position.
+Here we describe a new approach that replaces handcrafted statistical methods with a single, end-to-end deep learning model.
+Our model frames variant detection as a sequence-to-sequence modeling task, akin to language translation, and employs a transformer-based architecture to translate alignment columns into two predicted haplotype sequences.
+We demonstrate that this method predicts variant genotypes and phase accurately, has the ability to realign complex and ambiguous read mappings to produce accurate haplotype predictions, and uses the local surrounding read context to inform predictions about a given position.
 
@@
 
-Accurate inference of the DNA sequence variants in a sample is a fundamental component of modern genomics analysis.
+Accurate inference of the DNA sequence variants in a sample is a fundamental component of genomics analysis.
 Tools that aim to identify variants must overcome several challenges to produce an accurate set of variant calls.
-For instance, the sequenced reads themselves may contain errors in the form of miscalled bases or PCR-induced repeat length errors, or the tool used to align reads to the reference genome may produce incorrect mappings (e.g.
-Li et al.
-2014).
+For instance, the sequenced reads themselves may contain errors in the form of miscalled bases or PCR-induced repeat length errors, or the tools used to align reads to the reference genome may produce incorrect mappings (e.g.
+Nielson et al.
+2011, Li et al.
+2014, Goldfeder et al.
+2016).
+Identifying these errors and differentiating them from real genomic variation is challenging, and modern variant detection tools employ a variety of statistical and heuristic techniques to achieve high precision and sensitivity.
 
 All variant discovery tools must address two challenges.
-First, callers must identify potential alleles in a region.
-Then, given a set of allele candidates, they must assess the likelihood of each candidate (or pairs of candidates in the diploid case), to determine which should be included in the caller output.
+First, callers must identify the potential alleles in a region.
+Then, given a set of potential alleles, they must assess the likelihood of each candidate (or pairs of candidates in the diploid case), to determine which should be included in the caller output.
 Early callers, such as samtools / mpileup (Li et al.
@@
 
-More recent tools have incorporated elements of deep learning into the allele likelihood calculation.
+More recent tools have incorporated elements of deep learning into the allele likelihood calculation or allele generation steps.
 DeepVariant (Poplin et al.
-2018) builds on the HaplotypeCaller approach, but adds a Convolutional Neural Net (CNN) to classify variants as true or false positive detections.
+2018) uses a statistical method based on the HaplotypeCaller approach to identify candidates, but adds a Convolutional Neural Net (CNN) to classify variants as true or false positive detections.
 HELLO (Ramachandran et al.
 2020), designed to work on hybrid short- and long-read datasets, employs a mixture-of-experts approach with separate 1-dimensional convolutions across the read and allele dimensions of the input.
+Clair (Luo et al.
+2019, Luo et al.
+2020), uses a novel multitask deep learning approach to predict several properties of a potential variant at a given site, including zygosity and allele length.
 
-While replacing the statistical and heuristic components of the allele likelihood calculation with a deep learning model has resulted in improved specificity of variant callers, current tools continue to rely on complex, handcrafted statistical machinery to identify possible alleles.
 Here, we describe a new approach that replaces both the candidate allele generation and the likelihood calculation with a single, learned model.
 We recast variant detection as a sequence-to-sequence translation problem in which the input sequence is the series of bases aligning to a single genomic location, and the output sequences are the two predicted haplotypes.
-The sequence-to-sequence translation is performed by a transformer-based deep neural network (Vaswani et al.
+The sequence-to-sequence translation is learned by a transformer-based deep neural network (Vaswani et al.
 2017).
-We demonstrate that this approach identifies real genomic variants with sensitivity similar to current top-performing callers, albeit with lower specificity.
-
-While transformers have not been explored for variant detection, a model similar to ours was proposed by Baid et al.
-(2021) for generation of accurate consensus reads from PacBio data.
-The Baid model (`DeepConsensus') used a network architecture similar to ours, but included an extra token representing a gap between the aligned sequences, and a novel loss function that accounted for gaps between the predicted and true sequence in a region.
-In contrast we use a simple cross-entropy loss function, and do not introduce gaps in the aligned sequences compared to the reference.
+This approach does not require any of the statistical machinery and heuristic thresholds utilized by most modern callers, and instead learns how to reconstruct simple and complex alleles, with phase and genotype, directly from aligned NGS reads.
 
@@
 2018).
-Calls were obtained with the 10M model trained for 25 epochs on the 'bigger' training set (~$10^6$ training examples).
+For this comparison we used 25M model trained for 25 epochs on the 'bigger' training set (~$10^6$ training examples).
 We selected filtering criteria for each caller by reviewing ROC curves produced by \textit{vcfeval} and selecting values close to the $F_1$ maximizing value across samples.
@@
 \subsection{Complex regions}
-While many genomic variants are trivial to identify by examination of mapped reads, some regions require a more sophisticated reassembly to identify the true haplotype.
+While many genomic variants are trivial to identify by examination of mapped reads, some regions require a more sophisticated reassembly to identify the true haplotypes.
 We examine the ability of our model to perform this more complicated reconstruction by identifying real variants in regions containing multiple conflicting read mappings or extensive read soft-clipping.
@@
 
+\begin{figure}[htp]
+	\includegraphics[scale=0.325]{./encoding_ablation.png}
+	\caption{ Precision for each variant class for 1-dimensional and 2-dimensional positional encodings, for two different model sizes }
+	\label{fig:encoding_ablation}
+\end{figure}
+
 To assess the impact of the two-dimensional positional encoding, we replaced the 2D encoding module with traditional 1D encoding, and left other model parameters unmodified.
@@
 Both training runs were conducted for 25 epochs.
-When compared to their equivalent models with 2D encoding, the 1D runs were significantly compromised.
-In particular, precision for most variant classes was markedly lower than the 2D runs, and sensitivity was also reduced but to a lesser degree.
+When compared to their equivalent models with 2D encoding, the 1D encoding model showed substantially reduced precision, in particular for the SNV class (Figure \ref{fig:encoding_ablation}).
+Sensitivity was also reduced but to a lesser degree.
+
+Two-dimensional encoding allows the model to track which read contains a given base across genomic positions.
+We hypothesize that read-level tracking is important for precision because mismapped or poorly aligned reads generate many reference mismatches, most of which are likely to be errors.
+The model must learn to identify and ignore such reads, and 2D encoding facilitates this because the model can more easily track read-level features across input tokens.
 
@@
 
-We describe a new approach to the problem of detecting sequence variants in next-generation sequencing (NGS).
+We describe a new approach to the problem of detecting sequence variants in next-generation sequencing (NGS) data.
 Our approach envisions variant detection as a sequence-to-sequence modeling problem, akin to language translation, and leverages the successful transformer architecture to perform the translation.
 The sequence-to-sequence approach allows a single deep learning model to be used for candidate allele generation and haplotype construction.
-In contrast to other variant detection methods (...), our approach does not rely on any handcrafted statistical techniques such as de Bruijn graphs, hidden Markov models, or heuristic thresholding and instead learns to construct accurate haplotypes directly from aligned NGS reads.
+In contrast to other variant detection methods (e.g.
+Poplin et al.
+2018,  Kim et al.
+2018), our approach does not rely on any handcrafted statistical techniques such as de Bruijn graphs, hidden Markov models, or heuristic thresholding and instead learns to construct accurate haplotypes directly from aligned NGS reads.
 
@@
 
-While our model has comparable sensitivity to the other callers we examined, our specificity for SNV calls is lower.
-Many avenues existing for further improving performance, including adding additional training data with false positive SNV calls, adding new features to the quality-estimation classifier, or possibly running an ensemble of models.
+Overall variant detection accuracy is similar to other modern callers.
+At our chosen quality cutoffs (selected to be near the $F_1$-maximizing value for each caller), Jovian's performance on insertion-deletion variants outperforms HaplotypeCaller, but falls short of DeepVariant and Strelka2 (\ref{fig:sensitivity.png}).
+The differences are partially due to Jovian's relatively poor sensitivity for indels larger than 15bp (mean 75.8\% for deletions, 71.2\% for insertions) and low sensitivity in tandem repeat regions (mean sensitivity 84.2\%).
+For SNVs Jovian demonstrated the highest sensitivity among all callers (mean 99.0\%), and precision similar to HaplotypeCaller but lower than DeepVariant and Strelka2 (Jovian mean precision 99.55\%).
+
+While transformers have not been explored for variant detection, a model similar to ours was proposed by Baid et al.
+(2021) for generation of accurate consensus reads from PacBio data.
+The Baid model (`DeepConsensus') used a network architecture similar to ours, but included an extra token representing a gap between the aligned sequences, and a novel loss function that accounted for gaps between the predicted and true sequence in a region.
+In contrast we use a simple cross-entropy loss function, and do not introduce gaps in the aligned sequences compared to the reference.
 
@@
 Secondly, indel errors ar punished more severely than similar SNV errors.
-We hypothesize the second fact explains the lower performance of our model on SNV calls compared to indels.
+We hypothesize the second fact contributes to the lower performance of our model on SNV calls compared to indels.
 Alternative, alignment-based loss functions such as those described by Baid et al.
@@
 \vspace{8pt}
+Goldfeder, Rachel L., et al. "Medical implications of technical accuracy in genome sequencing." Genome medicine 8.1 (2016): 1-12.
+
+\vspace{8pt}
 Kim, Sangtae, et al. "Strelka2: fast and accurate calling of germline and somatic variants." Nature methods 15.8 (2018): 591-594.
@@
 Liu, Ze, et al. "Swin transformer: Hierarchical vision transformer using shifted windows." Proceedings of the IEEE/CVF International Conference on Computer Vision. 2021.
+
+\vspace{8pt}
+Luo, Ruibang, et al. "A multi-task convolutional deep neural network for variant calling in single molecule sequencing." Nature communications 10.1 (2019): 1-11.
+
+\vspace{8pt}
+Luo, Ruibang, et al. "Exploring the limit of using a deep neural network on pileup data for germline variant calling." Nature Machine Intelligence 2.4 (2020): 220-227.
 

### 2022-06-22 dnaseq2seq/util.py

@@ -31,6 +31,7 @@ def var_type(variant):
     print(f"Whoa, unknown variant type: {variant}")
     return 'unknown'
 
+
 def concat_metafile(sample_metafile, dest_metafh):
     """
     Concate the given sample metadata file to destination metadata file. 

### 2022-06-28 manuscript/manuscript.tex
Added short variant context section and plot

--- 2022-06-15 manuscript/manuscript.tex
@@
 
+Model input, derived from aligned reads, includes information about nearby non-reference bases and other context information such as base qualities.
+We were interested to see how model variant calling was informed by or correlated to aligned read information near a prospective variant.
+Prospective SNVs with a VAF much lower than the roughly 50\% expected for a nominal heterozygous variant are less likely to be true variants and are less likely to be called than those with higher VAF.
+This model decision process may involve the use of local context information.
+For example, the model may detect many low VAF prospective variants in a local region.
+In many cases this can be related to a difficult to sequence region that presents numerous false positive prospective variants.
+Through training, the model may learn to reduce variant call probability in such regions.
+Base quality scores, a local context parameter that can also indicate incorrect non-reference base calls, may also play an important roll in informing the model that a prospective variant call is incorrect.
+
+To explore the transformer models use of these local context data for prospective low VAF variants, we evaluated low VAF variant call probability relative to the total number of nearby low VAF prospective variants.
+Low versus High base quality was included to determine if it was also correlated to call probability and possibly used in the model decision process.
+Figure XXX shows three different low VAF ranges for prospective variants and what fraction of these variants were called by the model compared to the number of other nearby low VAF variants and also compared to base quality score.
+The results show a strong trend between low local prospective variant population and higher probability of being called by the model.
+Base quality scores, on the other hand, were not strongly linked to call probability or to the local population of prospective variants.
+This indicates that to some extent the model uses local non-reference base information to determine variant calls.
+
 \subsection{Ablation of 2D positional encoding}

### 2022-06-29 dnaseq2seq/call.py
25m params for calling

@@ -119,16 +119,16 @@ def reconcile_current_window(prev_win, current_win):
 def load_model(model_path):
     
     # 25m
-    #attention_heads = 8
-    #encoder_layers = 8
-    #transformer_dim = 400
-    #embed_dim_factor = 100
+    attention_heads = 8
+    encoder_layers = 8
+    transformer_dim = 400
+    embed_dim_factor = 100
     
     # 10m
-    attention_heads = 6
-    encoder_layers = 6
-    transformer_dim = 250
-    embed_dim_factor = 100
+    #attention_heads = 6
+    #encoder_layers = 6
+    #transformer_dim = 250
+    #embed_dim_factor = 100
     model = VarTransformer(read_depth=100,
                            feature_count=9,
                            out_dim=4,

### 2022-06-30 manuscript/manuscript.tex

@@
 We were interested to see how model variant calling was informed by or correlated to aligned read information near a prospective variant.
-Prospective SNVs with a VAF much lower than the roughly 50\% expected for a nominal heterozygous variant are less likely to be true variants and are less likely to be called than those with higher VAF.
+Prospective SNVs with a variant allele fraction (VAF) much lower than the roughly 50\% expected for a nominal heterozygous variant are less likely to be true variants and are less likely to be called than those with higher VAF.
 This model decision process may involve the use of local context information.
@@
 Through training, the model may learn to reduce variant call probability in such regions.
-Base quality scores, a local context parameter that can also indicate incorrect non-reference base calls, may also play an important roll in informing the model that a prospective variant call is incorrect.
+Base quality scores (BQ), a local context parameter that can also indicate incorrect non-reference base calls, may also play an important roll in informing the model that a prospective variant call is incorrect.
 
-To explore the transformer models use of these local context data for prospective low VAF variants, we evaluated low VAF variant call probability relative to the total number of nearby low VAF prospective variants.
-Low versus High base quality was included to determine if it was also correlated to call probability and possibly used in the model decision process.
-Figure XXX shows three different low VAF ranges for prospective variants and what fraction of these variants were called by the model compared to the number of other nearby low VAF variants and also compared to base quality score.
+To explore the transformer model's use of these local context data for prospective low VAF variants, we evaluated low VAF variant call probability relative to the total number of nearby low VAF prospective variants within a centered 41 base window.
+The data set included the approximately 96,000 low VAF (0.1 <= VAF <= 0.3) prospective variants from one sample on chromosome 21.
+Low BQ (< 30) versus high BQ (>=30) was included to determine if it was also correlated to call probability and possibly used in the model decision process.
+Figure XXX shows three different low VAF ranges for prospective variants and the fraction of these variants called by the model compared to the number of other nearby low VAF variants and also compared to base quality score.
 The results show a strong trend between low local prospective variant population and higher probability of being called by the model.

### 2022-07-01 manuscript/manuscript.tex
manuscript updates

--- 2022-06-22 manuscript/manuscript.tex
@@
 \graphicspath{ {./images/} }
-\linespread{1.15}
+\linespread{1.25}
 \textwidth=380pt
@@
 
+vim
+
 \section{Introduction}
@@
 For instance, the sequenced reads themselves may contain errors in the form of miscalled bases or PCR-induced repeat length errors, or the tools used to align reads to the reference genome may produce incorrect mappings (e.g.
-Nielson et al.
-2011, Li et al.
+Li et al.
 2014, Goldfeder et al.
 2016).
-Identifying these errors and differentiating them from real genomic variation is challenging, and modern variant detection tools employ a variety of statistical and heuristic techniques to achieve high precision and sensitivity.
+Identifying these errors and differentiating them from real genomic variation is challenging, and modern variant detection tools employ a variety of statistical and heuristic techniques to achieve high precision and sensitivity (De Pristo et al.
+2011, Nielson et al.
+2011).
 
@@
 2018, Cooke et al.
-2021) incorporated local re-assembly of reads in each region of interest, resulting in a significant improvement in the precision and specificity of variant calls.
+2021) incorporated local re-assembly of reads in each region of interest, resulting in a significant improvement in the sensitivity and precision of variant calls.
 For instance, the HaplotypeCaller tool identifies candidate haplotypes by constructing de Bruijn graphs from k-mers present in the reads, and assesses haplotype likelihoods with a pair Hidden Markov Model (HMM).
@@
 2017).
-This approach does not require any of the statistical machinery and heuristic thresholds utilized by most modern callers, and instead learns how to reconstruct simple and complex alleles, with phase and genotype, directly from aligned NGS reads.
+This approach does not require any of the statistical machinery and heuristic thresholds utilized by most modern callers, and instead uses a deep learning model to reconstruct complete haplotypes directly from aligned NGS reads.
 
@@
 
-Our model consists of a transformer-based encoder and simple, fully connected decoder that produces two output sequences, one for each predicted haplotype.
-The encoder component is an unmodified transformer with GeLU activations as implemented in PyTorch 1.10 (Paszke et al.
+Our model consists of a transformer-based encoder and a simple, fixed-length decoder that produces two output sequences, one for each predicted haplotype.
+The encoder component is an unmodified transformer with GeLU activations as implemented in PyTorch 1.10 ((Vaswani et al.
+2017, Paszke et al.
 2019).
@@
 We add an additional fully connected layer prior to the transformer encoders which embeds the encoded basecalls in $d$ dimensions, where $d=12$ for the analyses here.
-The embedded basecalls are 'flattened' along the read dimension, producing an input token with size $dr$ where $r$ is the number of reads ($r=100$ for all experiments proposed here).
+The embedded basecalls are `flattened' along the read dimension, producing an input token with size $dr$ where $r$ is the number of reads ($r=100$ for all experiments proposed here).
 Instead of the typical 1-dimensional positional encoding, we employed a 2-dimensional encoding, enabling the model to retain information about which read contained a given base across input tokens.
@@
 Predictions are discrete probability distributions over the four possible bases at each predicted output position, produced with a softmax function over the raw outputs.
-The decoder is of fixed size, and each predicted haplotype contains the same number of predicted bases as the number of positions $g$ in the input.
-
-Input tensors are generated by selecting a 150bp region from a set of aligned reads in BAM / CRAM format.
-For each region, all reads overlapping the window were obtained from the corresponding BAM/ CRAM file.
-Regions containing more than the maximum number of reads (100 for all analyses here) were downsampled to the maximum size.
-Reads are then sorted by the reference coordinate of the first aligned base.
-For each aligned base in the selected reads nine features are encoded; the first four were the one-hot encoded base call, followed by base quality, two flags indicating if base `consumed' reference base (i.e.
-was not an insertion) or consumed a sequence read base (was not a deletion), and additional flags indicating sequence read direction and clipping status.
-No 'gap' tokens or other special handling was performed for insertions or deletions.
-In addition to sequenced reads, the first row in each encoded region was the reference sequence.
-For this special row we inserted a base quality of 100 for every position and did not set any of the other flags.
-Resulting tensors had dimension $[g, r, 9]$, where $g$ represents genomic position, $r$ represents read, and 9 is the number of features.
-Positions in the input not corresponding to an aligned base were all set to zero.
+The decoder is of fixed length and produces two haplotype sequences equal in length to the number of input tokens $g$.
+The full model architecture is shown in \ref{fig:architecture}.
 
@@
 
+Input tensors are generated by selecting a 150bp region from a set of aligned reads in BAM / CRAM format.
+For each region, all reads overlapping the window were obtained from the corresponding alignment file.
+Regions containing more than the maximum number of reads (100 for all analyses here) were downsampled to the maximum size.
+Reads are then sorted by the reference coordinate of the first aligned base.
+For each aligned base in the selected reads nine features are encoded; the first four are the one-hot encoded base call, followed by base quality, two flags indicating if base `consumed' reference base (i.e.
+was not an insertion) or consumed a sequence read base (was not a deletion), and additional flags indicating sequence read direction and clipping status.
+No gap tokens or other special handling was performed for insertions or deletions.
+In addition to sequenced reads, the first row in each encoded region was the reference sequence.
+For this special row we inserted a base quality of 100 for every position and did not set any of the other flags.
+Resulting tensors had dimension $[g, r, 9]$, where $g$ represents genomic position, $r$ represents read, and 9 is the number of features.
+Positions in the input not corresponding to an aligned base were all set to zero.
+
 \subsection{NGS data}
 
-We obtained training data from 17 whole genomes sequenced from 5 Genome-in-a-Bottle cell lines.
+Training data was obtained from 17 whole genomes sequenced from 5 Genome-in-a-Bottle cell lines.
 Eight samples were prepared with Illumina Nextera DNA Flex kit and the remainder were prepared with the Illumina TruSeq PCR-free kit.
@@
 For all analyses described here, we obtain training data only from the human autosomes 1-20, and hold out chromosomes 21 and 22 for model evaluation.
-We examine three different sizes of training data sets, with 5.5M, 10.4M, and 20.4M examples.
+We examine three different sizes of training data sets with 5.1M, 10.2M and 20.4M examples.
 
@@
 In regions where phasing of the variants was ambiguous, the reads in the sequenced sample were examined to determine phase status.
-Briefly, all possible genotypes (pairs of haplotype sequences) were generated and reads were aligned via Smith-Waterman to the possible haplotypes, and the highest scoring genotypes selected as the most likely phasing.
+Briefly, all possible genotypes (pairs of haplotypes) were generated and reads were aligned via Smith-Waterman to the possible haplotypes, and the highest scoring genotypes selected as the most likely phasing.
 This phasing procedure was only attempted for variants less than 100bp apart, otherwise the region was discarded.
@@
 
-Models were implemented in Pytorch (v1.10, Paszke et al 2019). Models were trained for 25 epochs unless noted otherwise, using the Adam optimizer and a learning rate of $5e-5$ and batch size of 512.
+Models were trained for 30 epochs unless noted otherwise, using the Adam optimizer and a learning rate of $5e-5$ and batch size of 512. Models were implemented in PyTorch (v1.10, Paszke et al 2019).
 
@@
 
-Given a region containing suspected variants, we perform multiple overlapping forward passes of the model with step size $k$, where $k=50$ for the results reported here.
-On each forward pass the model predicts two haplotypes by taking the base with the high predicted probability at each position.
+For a single region containing suspected variants, we perform multiple overlapping forward passes of the model with step size $k$, where $k=50$ for the results reported here.
+On each forward pass the model predicts two haplotypes by taking the base with the highest predicted probability at each position.
 Each haplotype is aligned via Smith-Waterman to the reference sequence, and any mismatching positions are converted to variant calls.
 After variants are collected across multiple windows, results are merged using a method that attempts to minimize the number of conflicting calls between windows.
-For each variant we record the number of windows in which each variant was detected, the number of variants in cis and trans, and the mean probability of the variant bases, and the position of the variant within each window.
+For each variant we record the number of windows in which each variant was detected, the number of variants in cis and trans, the mean probability of the variant bases, and the position of the variant within each window.
 
@@
 \begin{center}
-	\begin{tabular}{ |c|c|c|c|c| }
+	\begin{tabular}{ lllll }
 	 Model name & Encoder layers & Attention heads & Hidden dimension & Embedding dimension \\ 
@@
 
-\subsection{Genotyping \& Phasing}
-
-To check the accuracy of predicted genotypes and phasing of variants we compared calls to the known genotype and, when available, phase from Genome-in-a-Bottle (GIAB) variant data.
-Approximately 99.5\% of homozygous  and 99.0\% of heterozygous calls matched the true genotype, indicating that our model learns to construct both haplotypes accurately.
-For phasing accuracy, we identified pairs of heterozygous variants less than 100 base pairs apart and compared their phasing predictions with the known phasing information from GIAB.
-The phasing information in GIAB samples is only available for HG001, HG002 and HG005 cell lines, so analysis was restricted to those samples.
-Phase predictions matched GIAB phase for approximately 96.9\% of SNV pairs and 99.3\% of indel pairs.
-
-Accurate phase prediction requires that information regarding which reads support a given haplotype is propogated among input tokens (genomic positions).
-Our model's relatively high phasing accuracy suggests that
-
 \subsection{Variant detection accuracy}
@@
 2018).
-For this comparison we used 25M model trained for 25 epochs on the 'bigger' training set (~$10^6$ training examples).
+For this comparison we used 25M model trained for 25 epochs on the 'huge' training set (~$20^6$ training examples).
 We selected filtering criteria for each caller by reviewing ROC curves produced by \textit{vcfeval} and selecting values close to the $F_1$ maximizing value across samples.
-Jovian calls were filtered at 0.4, HaplotypeCaller at 60 (phred-scaled), Strelka at 4 (phred-scaled), and DeepVariant at 0.
+Jovian calls were filtered at 0.25, HaplotypeCaller at 60 (phred-scaled), Strelka at 4 (phred-scaled), and DeepVariant at 0.
 Sensitivity, precision, and related calculations were computed with the \textit{hap.py} tool (Krushe et al.
@@
 
-Across all regions, Jovian's sensitivity and precision for indels is slightly higher than that for HaplotypeCaller, and slightly lower than DeepVariant or Strelka2.
-For SNV calls Jovian has similar sensitivity to other callers but the lowest precision of the callers ()
+\begin{figure}[htp]
+	\includegraphics[scale=0.5]{./sensitivity-precision.png}
+	\caption{ Sensitivity and precision of Jovian variant calls (10M parameter model, large training size) compared to other callers }
+	\label{fig:sens-ppv}
+\end{figure}
+
+Across all regions, Jovian's sensitivity and precision for indels is slightly higher than that for HaplotypeCaller, and slightly lower than DeepVariant or Strelka2 (\ref{fig:sens-ppv}).
+For SNV calls Jovian has the highest sensitivity (99.1\%) and precision similar to HaplotypeCaller (99.5\%), but precision is lower than DeepVariant and Strelka2 (99.8\% and 99.9\%, respectively).
+
+Across indel size classes Jovian has high sensitivity for deletions less than about 15bp, but relatively low sensitivity for larger deletions (~75\%).
+For insertions Jovian has sensitivity similar to other callers at smaller sizes, but again reduced sensitivity for the largest class (~68\%).
+Precision follows a similar pattern, but with relatively high precision for smaller indel sizes and reduced precision at larger sizes.
+
+\begin{figure}[htp]
+	\includegraphics[scale=0.5]{./accuracy_indel_size.png}
+	\caption{ Sensitivity and precision of Jovian indel calls by size class (10M parameter model, large training size) compared to other callers }
+	\label{fig:indelsize-sens-ppv}
+\end{figure}
+
+\subsection{Genotyping \& Phasing}
+
+To check the accuracy of predicted genotypes and phasing of variants we compared calls in the chromosome 21 and 22 evaluation regions to the known genotype and phase from Genome-in-a-Bottle (GIAB) variant data.
+Approximately 99.5\% of homozygous and 99.0\% of heterozygous calls matched the true genotype, indicating that our model learns to construct both genotypes accurately.
+For phasing accuracy, we identified pairs of heterozygous variants less than 100 base pairs apart and compared their phasing predictions with the known phasing information from GIAB.
+The phasing information in GIAB samples is only available for HG001, HG002 and HG005 cell lines, so analysis was restricted to those samples.
+Phase predictions matched GIAB phase for approximately 96.9\% of SNV pairs and 99.3\% of indel pairs.
 
@@
 
-Our model was able to reconstruct the correct haplotypes for ~90.8\% (mid 50\% 86.4-95.2) of these variants, somewhat lower than HaplotypeCaller accuracy of ~95.8\% (mid 50\% 91.0-100.0).
-Although model performance does not match current state-of-the-art callers, it nonetheless is able to reconstruct the true haplotype in most cases.
-Overall, these results indicate that the model learns a reconstruction procedure more nuanced than a simple column-by-column examination of reads,
+Our model was able to reconstruct the correct haplotypes for ~89.5\% (mid 50\% 82.5-94.8) of these variants, somewhat lower than other callers (\ref{table:complex-sensitivity}).
+These results indicate that the model learns a reconstruction procedure more nuanced than a simple column-by-column examination of reads.
+Instead, the model learns to synthesize information across tokens to produce complete haplotypes with accuracy similar to (but slightly less than) the top performing current methods.
+
+\begin{center}
+	\begin{tabular}{ ccc }
+	 Caller & Sensitivity & Mid 50\%  \\ 
+	\hline
+	 Jovian(10M) & 89.5\% & 82.5 - 94.8 \\ 
+	 HaplotypeCaller & 95.8\% & 91.0 - 100.0  \\ 
+	 DeepVariant & 91.7\% & 83.5 - 97.1  \\ 
+	 Strelka & 93.1\% & 90.9 - 98.4 \\
+	 \hline
+	\end{tabular}
+	\label{table:complex-sensitivity}
+\end{center}
 
 \subsection{Use of local context}
- - Also ambiguous, but anecdotally when looking at variant calls, it seems apparent that variants that otherwise look real have lower quality scores when there's a lot of junk around.
-That's good, since such variants are more likely To
- be false positives, and that means the model is to some extent looking at local context to modify prediction accuracy. Which is cool.
+
+We were interested to see how model variant calling was informed by contextual information from nearby alignment columns.
+In particular, the model must learn to discriminate real variation from noise at low Variant Allele Frequency (VAF, the fraction of reads supporting a given allele candidate), and this model decision process may involve the use of local context information.
+For example, the model may be less likely to call a SNV in a region containing many reference mismatches, because the presence of multiple mismatches suggests poor sequencing quality or other technical artifacts.
+However, the call probability is likely to be modified by other factors, including VAF and the base call quality associated with the prospective variant.
+
+To examine the influence of the number of nearby mismatches, VAF, and base quality on the likelihood that a prospective SNV was called, we first collected genomic positions where 10 - 30\% of reads supported a non-reference basecall.
+We then tallied the number of genomic positions at which 10\% - 20\% of reads supported a non-reference call in a window 20bp upstream and downstream of the candidate.
+We also stratified the candidate position into low base quality and high base quality groups, and smaller VAF windows of 10-20\%, 15-25\% and 20-30\%.
+
+\begin{figure}[htp]
+	\includegraphics[scale=0.25]{./variant_context_BQ.png}
+	\caption{ Effect of number of nearby reference mismatches on the likelihood that a poorly supported candidate variant will be called. }
+	\label{fig:variant_context}
+\end{figure}
+
+The probability that a prospective variant is called is strongly associated with the number of nearby alignment columns containing mismatches (Figure \ref{fig:variant_context}).
+For instance, prospective variants with VAF between 20-30\% and high base quality were called in over 60\% of examined cases when there were no nearby mismatches, but called in less than 30\% of cases when a single other column contains a mismatch, and in less than 10\% of cases when 5 or more nearby positions contain a mismatch.
+A similar, but less pronounced trend exists when the prospective variant is at lower VAF (left two panels, figure \ref{fig:variant_context}).
+When the prospective variant has poor base quality scores, the association disappears and the likelihood that the variant is called remains uniformly low.
+Together, these results indicate that the models learns a nuanced SNV detection procedure that incorporates multiple elements, including base quality, VAF, and information from local surrounding bases to discriminate actual genetic variation from technical artifacts and sequencing errors.
 
@@
 	\includegraphics[scale=0.325]{./encoding_ablation.png}
-	\caption{ Precision for each variant class for 1-dimensional and 2-dimensional positional encodings, for two different model sizes }
+	\caption{ Precision of variant calls for each variant class for 1-dimensional and 2-dimensional positional encodings, for two different model sizes }
 	\label{fig:encoding_ablation}
@@
 Our analysis demonstrates that the model learns many features of modern, state of the art variant callers.
-For instance, our model learns to reconstruct large, complex variants in the presence of ambiguous mapping and extensive soft clipping with a sensitivity greater than 90\%.
+For instance, our model learns to reconstruct large, complex variants in the presence of ambiguous mapping and extensive soft clipping with a sensitivity near 90\%.
 In addition, the model learns to employ local context and read-specific features to predict phase and genotype accurately, and pairs of variants are phased correctly in ~97-99\% of cases where variants are within 100bp.
@@
 Overall variant detection accuracy is similar to other modern callers.
-At our chosen quality cutoffs (selected to be near the $F_1$-maximizing value for each caller), Jovian's performance on insertion-deletion variants outperforms HaplotypeCaller, but falls short of DeepVariant and Strelka2 (\ref{fig:sensitivity.png}).
-The differences are partially due to Jovian's relatively poor sensitivity for indels larger than 15bp (mean 75.8\% for deletions, 71.2\% for insertions) and low sensitivity in tandem repeat regions (mean sensitivity 84.2\%).
-For SNVs Jovian demonstrated the highest sensitivity among all callers (mean 99.0\%), and precision similar to HaplotypeCaller but lower than DeepVariant and Strelka2 (Jovian mean precision 99.55\%).
+At our chosen quality cutoffs (selected to be near the $F_1$-maximizing value for each caller), Jovian's performance on insertion-deletion variants outperforms HaplotypeCaller, but falls short of DeepVariant and Strelka2 (\ref{fig:sens-ppv}).
+The differences are partially due to Jovian's relatively poor sensitivity for indels larger than 15bp (mean 75.6\% for deletions, 68.8\% for insertions) and low sensitivity in tandem repeat regions (mean sensitivity 84.0\%).
+For SNVs Jovian demonstrated the highest sensitivity among all callers (mean 99.1\%), and precision similar to HaplotypeCaller but lower than DeepVariant and Strelka2 (Jovian mean precision 99.5\%).
 
-While transformers have not been explored for variant detection, a model similar to ours was proposed by Baid et al.
+To our knowledge, the transformer architecture has not been previously explored for variant detection, but a model similar to ours was proposed by Baid et al.
 (2021) for generation of accurate consensus reads from PacBio data.
 The Baid model (`DeepConsensus') used a network architecture similar to ours, but included an extra token representing a gap between the aligned sequences, and a novel loss function that accounted for gaps between the predicted and true sequence in a region.
-In contrast we use a simple cross-entropy loss function, and do not introduce gaps in the aligned sequences compared to the reference.
+In contrast we use a simple cross-entropy loss function, do not introduce gaps in the aligned sequences compared to the reference, and use 2-D positional encoding.
 
 In this work we have used a standard cross-entropy loss function that simply compares each prediction at base $i$ to the label at base $i$.
-While training with this loss yields accurate results, it has two significant shortcomings.
+While training with this loss yields accurate results, it has two shortcomings.
 First, the impact of an insertion / deletion (indel) error depends on its position in the region, with indels at lower indices leading to a greater number of mismatches and high loss compared to indels at higher indices.
-Secondly, indel errors ar punished more severely than similar SNV errors.
+Secondly, indel errors are punished more severely than similar SNV errors.
 We hypothesize the second fact contributes to the lower performance of our model on SNV calls compared to indels.

### 2022-08-05 manuscript/manuscript.tex
bq figure

@@
 
-\title{Variant calling with transformers}
+\title{Sequence modeling enables direct inference of germline haplotypes from short read sequencing data}
 \author[1]{Brendan O'Fallon}
@@
 \begin{abstract}
-	Detection of germline variants in next-generation sequencing data is an essential element of modern genomics.
-Variant detection tools typically rely on a variety of statistical algorithms and heuristic techniques to identify variants.
-Here we describe a new approach that replaces handcrafted statistical methods with a single, end-to-end deep learning model.
+	Detection of germline variants in next-generation sequencing data is an essential component of modern genomics analysis.
+Variant detection tools typically rely on statistical algorithms such as de Bruijn graphs, Hidden Markov Models and regression models, often coupled with heuristic techniques and thresholds to identify variants.
+Here we describe a new approach that replaces these handcrafted statistical methods with a single, end-to-end deep learning that directly infers germline haplotypes from short read pileups.
 Our model frames variant detection as a sequence-to-sequence modeling task, akin to language translation, and employs a transformer-based architecture to translate alignment columns into two predicted haplotype sequences.
-We demonstrate that this method predicts variant genotypes and phase accurately, has the ability to realign complex and ambiguous read mappings to produce accurate haplotype predictions, and uses the local surrounding read context to inform predictions about a given position.
-
+After training with 17 whole genome sequences from Genome-in-a-Bottle cell lines, we demonstrate that this method predicts variant genotypes and phase accurately, has the ability to realign complex and ambiguous read mappings to produce accurate haplotype predictions, and uses the local read context to inform predictions about a given position.
+We also demonstrate that a 2-dimensional positional encoding significantly improved precision of the detected variants.
+Compared to other callers, sensitivity and precision is higher than GATK HaplotypeCaller, but lower than DeepVariant and Strelka2.
 \end{abstract}
-
-vim
 
@@
 
-Accurate inference of the DNA sequence variants in a sample is a fundamental component of genomics analysis.
+Accurate inference of the DNA sequence variants in a sample is a fundamental element of genomics analysis.
 Tools that aim to identify variants must overcome several challenges to produce an accurate set of variant calls.
@@
 2016).
-Identifying these errors and differentiating them from real genomic variation is challenging, and modern variant detection tools employ a variety of statistical and heuristic techniques to achieve high precision and sensitivity (De Pristo et al.
+In addition, large or complex sequence variants may be obscured by limitations in the read alignment algorithm, and require a sophisticated assembly of reads to identify.
+Reconstructing real variation and discriminating it from technical error is challenging, and modern variant detection tools employ a variety of statistical and heuristic techniques to achieve high precision and sensitivity (e.g.
+De Pristo et al.
 2011, Nielson et al.
@@
 2009) and the UnifiedGenotyper tool from the Genome Analysis ToolKit (GATK, De Pristo et al.
-2011) rely on the read aligner to generate candidate allele candidates, and utilize ad-hoc heuristics and thresholds to determine likely alleles (Nielsen et al.
+2011) rely on the read aligner to generate candidate allele candidates, and utilize several ad-hoc heuristics and thresholds to determine likely alleles (e.g.
+Nielsen et al.
 2011).
@@
 Our model consists of a transformer-based encoder and a simple, fixed-length decoder that produces two output sequences, one for each predicted haplotype.
-The encoder component is an unmodified transformer with GeLU activations as implemented in PyTorch 1.10 ((Vaswani et al.
+The encoder component is an unmodified transformer with GeLU activations as implemented in PyTorch 1.10 (Vaswani et al.
 2017, Paszke et al.
@@
 The embedded basecalls are `flattened' along the read dimension, producing an input token with size $dr$ where $r$ is the number of reads ($r=100$ for all experiments proposed here).
-Instead of the typical 1-dimensional positional encoding, we employed a 2-dimensional encoding, enabling the model to retain information about which read contained a given base across input tokens.
+Instead of the typical 1-dimensional positional encoding, we employed a 2-dimensional encoding, allowing the model to differentiate across both the token (position) and read dimensions.
+The 2D encoding implementation follows Wang \& Jyh-Charn (2021).
 
-The decoder consists of a single fully-connected layer followed by two additional fully connected components that each generate a single haplotype.
+The decoder consists of a single fully-connected layer with ReLU activation followed by two additional fully connected components that each generate a single haplotype.
 Predictions are discrete probability distributions over the four possible bases at each predicted output position, produced with a softmax function over the raw outputs.
@@
 2015).
-We used the scikit-learn implementation of the random forest with 100 trees, a maximum tree depth of 25, and 'balanced' class weighting option.
+We used the scikit-learn implementation of the random forest with 100 trees, a maximum tree depth of 25, and `balanced' class weighting option.
+
+\begin{center}
+	\begin{tabular}{ l }
+	 \textbf{Feature} \\
+	 Mean probability of base call at all positions affected by variant \\ 
+	 Number of reference bases affected \\
+	 Number of alternate bases affected \\
+	 Minimum probability of base call at all positions affected by variant \\
+	 Maximum probability of base call at all positions affected by variant \\
+	 Number of variants in \textit{cis} (same haplotype) with this variant \\
+	 Number of variants in \textit{trans} (other haplotype) with this variant \\ 
+	 Total number of variants in both \textit{cis} and \textit{trans} \\
+	 Distance in bases of call from beginning of window - minimum across windows \\
+	 Distance in bases of call from beginning of window - maximum across windows \\
+	 Read depth at variant position \\
+	 Gnomad genome frequency (v2.0.2) of variant \\
+	 Zygosity (heterozygous or homozygous) of variant \\
+	 Fraction of reads in BAM/CRAM supporting variant call \\
+	 Fraction of reads with mapping quality > 10 supporting variant call \\
+	 Total number of reads supporting variant call \\
+	\end{tabular}
+	\label{table:features}
+\end{center}
 
@@
 \vspace{8pt}
+Wang, Zelun, and Jyh-Charn Liu.
+"Translating math formula images to LaTeX sequences using deep neural networks with sequence-level training." International Journal on Document Analysis and Recognition (IJDAR) 24.1 (2021): 63-75.
+
+\vspace{8pt}
 Wu, Chuhan, et al. "Fastformer: Additive attention can be all you need." arXiv preprint arXiv:2108.09084 (2021).

### 2022-08-05 dnaseq2seq/buildclf.py

@@ -8,6 +8,7 @@ import scipy.stats as stats
 import pysam
 import pickle
 import sklearn
+import concurrent.futures
 from sklearn.ensemble import RandomForestClassifier
 
 from functools import lru_cache
@@ -136,6 +137,7 @@ def load_model(path):
     logger.info(f"Loading model from {path}")
     with open(path, 'rb') as fh:
         return pickle.load(fh)
+
 
 def train_model(conf, threads, var_freq_file, feat_csv=None, labels_csv=None):
     alltps = []

### 2022-08-05 dnaseq2seq/buildclf.py
Fixes for mp when training classifiers

@@ -282,7 +282,9 @@ def varstr(var):
 
 
 def extract_feats(args):
-    vcf, aln, var_freq_file, is_real = args
+    vcf, alnpath, var_freq_file, is_real, reference_filename = args
+    aln = pysam.AlignmentFile(alnpath, reference_filename=reference_filename)
+    var_freq_file = pysam.VariantFile(var_freq_file)
     allfeats = []
     feat_strs = []
     labels = []
@@ -315,7 +317,6 @@ def load_model(path):
 def train_model(conf, threads, var_freq_file, feat_csv=None, labels_csv=None, reference_filename=None):
     alltps = []
     allfps = []
-    var_freq_file = pysam.VariantFile(var_freq_file)
     if feat_csv:
         logger.info(f"Writing feature dump to {feat_csv}")
         feat_fh = open(feat_csv, "w")
@@ -327,24 +328,20 @@ def train_model(conf, threads, var_freq_file, feat_csv=None, labels_csv=None, re
         futs = []
         for sample in conf.keys():
             logger.info(f"Processing sample {sample}")
-            aln = pysam.AlignmentFile(conf[sample]['bam'], reference_filename=reference_filename)
+            print(conf[sample])
+            alnpath = conf[sample]['bam']
             tps = conf[sample].get('tps')
             if type(tps) == str:
-                futs.append(pool.submit(extract_feats, (tps, aln, var_freq_file, True)))
-                # alltps.extend(extract_feats(tps, aln, var_freq_file))
+                futs.append(pool.submit(extract_feats, (tps, alnpath, var_freq_file, True, reference_filename)))
             elif type(tps) == list:
                 for tp in tps:
-                    futs.append(pool.submit(extract_feats, (tp, aln, var_freq_file, True)))
-                    # alltps.extend(extract_feats(tp, aln, var_freq_file))?
+                    futs.append(pool.submit(extract_feats, (tp, alnpath, var_freq_file, True, reference_filename)))
             fps = conf[sample].get('fps')
             if type(fps) == str:
-                futs.append(pool.submit(extract_feats, (fps, aln, var_freq_file, False)))
-                # allfps.extend(extract_feats(fps, aln, var_freq_file))
+                futs.append(pool.submit(extract_feats, (fps, alnpath, var_freq_file, False, reference_filename)))
             elif type(fps) == list:
                 for fp in fps:
[…]

### 2022-08-06 dnaseq2seq/buildclf.py
Test classifier accuracy when training

@@ -8,6 +8,7 @@ import bisect
 import pysam
 import pickle
 import sklearn
+from sklearn.model_selection import train_test_split
 import concurrent.futures
 from sklearn.ensemble import RandomForestClassifier
 import scipy.stats as stats
@@ -364,8 +365,16 @@ def train_model(conf, threads, var_freq_file, feat_csv=None, labels_csv=None, re
     total_tps = sum(x for x in alllabels)
     logger.info(f"Loaded {total_tps} TP and {len(alllabels) - total_tps} FPs")
 
+    feat_train, feat_test, labels_train, labels_test = train_test_split(np.array(allfeatures), np.array(alllabels), test_size=0.05)
+    logger.info(f"Holding out {len(labels_test)} for model validation")
     clf = RandomForestClassifier(n_estimators=100, max_depth=25, random_state=0, max_features=None, class_weight="balanced", n_jobs=threads)
-    clf.fit(np.array(allfeatures), np.array(alllabels))
+    clf.fit(feat_train, labels_train)
+
+    test_preds = clf.predict(feat_test)
+    for t, p in zip(labels_test, test_preds):
+        print(f"{t}\t{p:4f}")
+    accuracy = labels_test * test_preds + (1 - labels_test) * (1-test_preds)
+    logger.info(f"Accuracy: {accuracy*100:.4f}")
     return clf
 
 

### 2022-08-09 manuscript/manuscript.tex

@@
 Regions of the reference genome overlapping the high-confidence regions from Genome-in-a-Bottle were subdivided into 150bp nonoverlapping windows, and these regions labelled according to the presence of variants.
-Separate labels were generated for regions conataining a single SNV, deletion, or insertion, as well as regions containing multiple insertion-deletions, or those containing variants intersecting low-complexity regions.
+Separate labels were generated for regions containing a single SNV, deletion, or insertion, as well as regions containing multiple insertion-deletions, or those containing variants intersecting low-complexity regions.
 We additionally included `true negative' regions where no known variant was present.
+In regions containing more than 100 overlapping reads, the reads were repeatedly downsampled to generate multiple training regions.
 For all analyses described here, we obtain training data only from the human autosomes 1-20, and hold out chromosomes 21 and 22 for model evaluation.
-We examine three different sizes of training data sets with 5.1M, 10.2M and 20.4M examples.
+Approximately 1.3M regions were obtained from each of the 17 samples, generating a full training data size of 24.3M regions.
+In addition to the "Full" training set we also examine a half-sized training data set of 12.1M regions, and quarter-sized set of 6.0M regions.
+The number of regions generated for each variant type is shown in Table \ref{table:regioncounts}.
 
@@
 
+\begin{center}
+	\begin{tabular}{ lc }
+	 \textbf{Region type} & \textbf{Number of regions (per sample)} \\
+	 No variant & 200,000 \\
+	 SNV & 200,000 \\
+	 Insertions & 200,000 \\
+	 Deletions & 200,000 \\
+	 Insertion/Deletion + SNV & 50,000 \\
+	 Insertion + Deletion & 30,000 \\
+	 SNV (low complexity) & 80,000 \\
+	 Insertion (low complexity) & 80,000 \\
+	 Deletion (low complexity) & 80,000 \\  
+	 Insertion + Deletion (low complexity) & 20,000 \\
+	 Insertion/Deletion + SNV (low complexity) & 80,000 \\
+	\end{tabular}
+	\label{table:regioncounts}
+\end{center}
+
 \subsection{Training procedure \& loss function}
 
-Models were trained for 30 epochs unless noted otherwise, using the Adam optimizer and a learning rate of $5e-5$ and batch size of 512. Models were implemented in PyTorch (v1.10, Paszke et al 2019).
+Models were trained for 50 epochs unless noted otherwise, using the Adam optimizer and a learning rate of $5e^{-5}$ and batch size of 512.
+Models were implemented in PyTorch (v1.10, Paszke et al 2019).
 
-For a given haplotype we use a standard cross-entropy loss function to compare the prediction to the target sequence.
+For a given haplotype we use a modified cross-entropy loss function to compare the predictions to the target sequence.
 However, two haplotypes exist for each region and they have no inherent order.
@@
 The above variant detection procedure has two shortcomings.
-First, if every potential variant is emitted, then specificity is poor.
+First, if every potential variant is emitted, then precision is poor.
 Second, the model does not innately produce well calibrated variant quality scores.
-To address these shortcomings we introduce a post-hoc random forest classifier that we train to discriminate true and false-positive variant calls (see table X for features).
+To address these shortcomings we introduce a post-hoc random forest classifier that we train to discriminate true and false-positive variant calls (see table \ref{table:features} for list of features).
 The resulting score is used as a final variant quality score suitable for tuning the tradeoff between sensitivity and specificity.
@@
 
-We explore the effects of model size and training data size on validation loss and variant detection accuracy. We evaluate three model sizes and two sets of training data.
+We explore the effects of model size and training data size on variant detection accuracy.
+Throughout, we calculate sensitivity and precision on a small set of evaluation regions from chromosomes 21 and 22 that were not used for training.
+Importantly, for performance reasons these sensitivity and precision calculations do \textit{not} include the full variant detection procedure with multiple overlapping windows and the post-calling classifier, and instead reflect results from a single inference procedure on the target window (accuracy of the full detection procedure is explored below).
+We evaluate three model sizes and three sets of training data (Full, Half, and Quarter-sized, see section Training Data above for details).
 
@@
 \end{center}
+
+During training, sensitivity and precision as measured on the holdout regions from chromosomes 21 and 22 increase for all variant classes, with returns dwindling after 30-40 epochs of training.
+Little difference was observed between the "full" and "half" sized training data sets, but both sensitivity and precision were lower for the quarter-sied training set.
+These results indicate that additional training examples beyond the "half" size ($12e^6$ example regions) offer little improvement in accuracy, and suggest that performance becomes limited by other factors, such as model capacity, inaccuracies in the training data set, or a non-optimal loss function.
+
+\begin{figure}[htp]
+	\includegraphics[scale=0.5]{./trainsize_effect.png}
+	\caption{ Sensitivity and precision curves for the 10M model during training for different training data sizes.
+Dashed lines are raw values, solid lines indicate rolling average with window size of 5.
+}
+	\label{fig:trainsize}
+\end{figure}
 
@@
 
+Limitations
+ - Low SNV precision
+ - not using strand bias?
+ -
+
 \section{Availability}

### 2022-08-31 scripts/run_pregen.sh
run scripts

@@ -19,9 +19,9 @@ ds2s=/uufs/chpc.utah.edu/common/home/u0379426/src/dnaseq2seq/dnaseq2seq/main.py
 #CONF=/uufs/chpc.utah.edu/common/home/u0379426/src/dnaseq2seq/wgs_multindel_conf.yaml
 #CONF=/uufs/chpc.utah.edu/common/home/u0379426/src/dnaseq2seq/wgs_multindel_splits_chrsE.yaml
 #CONF=/uufs/chpc.utah.edu/common/home/u0379426/src/dnaseq2seq/wgs_bigvars_conf.yaml
-CONF=/uufs/chpc.utah.edu/common/home/u0379426/src/dnaseq2seq/wgs_multindel_splits_chrsE.yaml
+CONF=/uufs/chpc.utah.edu/common/home/u0379426/src/dnaseq2seq/wgs_multindel_conf.yaml
 
-DEST=/uufs/chpc.utah.edu/common/home/arup-storage3/u0379426/wgs_pregen_mqfeat_chrsE/
+DEST=/uufs/chpc.utah.edu/common/home/arup-storage3/u0379426/wgs_pregen_mqfeat_chrs21and22_val
 
 BATCH_SIZE=512
 

### 2022-08-31 scripts/run_train_clone.sh

@@ -30,7 +30,7 @@ PYTHON=$HOME/miniconda3/envs/ds2s/bin/python
 CONF=/uufs/chpc.utah.edu/common/home/u0379426/src/dnaseq2seq/chpc_conf3.yaml
 
 #VAL_DIR=/uufs/chpc.utah.edu/common/home/arup-storage3/u0379426/pregen_wgs_readwindowfix_w150_chr21and22
-VAL_DIR=/uufs/chpc.utah.edu/common/home/arup-storage3/u0379426/test_pregen_mqfeat_kmers
+VAL_DIR=/uufs/chpc.utah.edu/common/home/u0379426/storage/wgs_pregen_mqfeat_chrs21and22_val
 
 #PREGEN_DIR=/uufs/chpc.utah.edu/common/home/arup-storage3/u0379426/wgs_pregen_halfhuge
 #PREGEN_DIR=/uufs/chpc.utah.edu/common/home/arup-storage3/u0379426/test_pregen_mqfeat_kmers
@@ -38,14 +38,15 @@ VAL_DIR=/uufs/chpc.utah.edu/common/home/arup-storage3/u0379426/test_pregen_mqfea
 #PREGEN_DIR=/uufs/chpc.utah.edu/common/home/arup-storage3/u0379426/pregen_wgs_w150_nochr21or22_big
 
 #PREGEN_DIR=/uufs/chpc.utah.edu/common/home/arup-storage3/u0379426/wgs_pregen_huge
-PREGEN_DIR=/uufs/chpc.utah.edu/common/home/arup-storage3/u0379426/wgs_pregen_halfhuge_lc_bigvars_fpfns
+#PREGEN_DIR=/uufs/chpc.utah.edu/common/home/arup-storage3/u0379426/wgs_pregen_halfhuge_lc_bigvars_fpfns
+PREGEN_DIR=/uufs/chpc.utah.edu/common/home/u0379426/storage/wgs_pregen_mqfeat_all
 
 LEARNING_RATE=0.00005
 
 CHECKPOINT_FREQ=1
 
 RUN_NAME="wgs_decoder_mqfeattest"
-RUN_NOTES="A test of the decoder model"
+RUN_NOTES="Decoder model with MapQ feature"
 
 set -x
 
@@ -68,7 +69,7 @@ cd ..
 
 echo "Branch: $GIT_BRANCH \n commit: $COMMIT \n" >> git_info.txt
 
[…]

### 2022-08-31 dnaseq2seq/train.py
35M params

@@ -330,12 +330,12 @@ def train_epochs(epochs,
                  wandb_notes="",
                  cl_args = {}
 ):
-    encoder_attention_heads = 4 # was 4
+    encoder_attention_heads = 8 # was 4
     decoder_attention_heads = 4 # was 4
     dim_feedforward = 512
-    encoder_layers = 4
-    decoder_layers = 2 # was 2
-    embed_dim_factor = 100 # was 100
+    encoder_layers = 6
+    decoder_layers = 4 # was 2
+    embed_dim_factor = 120 # was 100
     model = VarTransformer(read_depth=max_read_depth,
                             feature_count=feats_per_read, 
                             kmer_dim=util.FEATURE_DIM, # Number of possible kmers

### 2022-08-31 scripts/run_train_clone.sh
run scr

@@ -45,8 +45,8 @@ LEARNING_RATE=0.00005
 
 CHECKPOINT_FREQ=1
 
-RUN_NAME="wgs_decoder_mqfeattest"
-RUN_NOTES="Decoder model with MapQ feature"
+RUN_NAME="wgs_decoder_35m_mqfeattest"
+RUN_NOTES="Decoder model with MapQ feature, bigger 35M model"
 
 set -x
 

### 2022-08-31 manuscript/manuscript.tex
Move main.py out of bin/

@@
 \linespread{1.25}
-\textwidth=380pt
+\textwidth=400pt
 
-\title{Sequence modeling enables direct inference of germline haplotypes from short read sequencing data}
+\title{Jovian enables direct inference of germline haplotypes from short reads via sequence-to-sequence modeling}
 \author[1]{Brendan O'Fallon}
@@
 Variant detection tools typically rely on statistical algorithms such as de Bruijn graphs, Hidden Markov Models and regression models, often coupled with heuristic techniques and thresholds to identify variants.
-Here we describe a new approach that replaces these handcrafted statistical methods with a single, end-to-end deep learning that directly infers germline haplotypes from short read pileups.
-Our model frames variant detection as a sequence-to-sequence modeling task, akin to language translation, and employs a transformer-based architecture to translate alignment columns into two predicted haplotype sequences.
+Here we describe a new approach that replaces these handcrafted statistical methods with a single, end-to-end deep learning model that directly infers germline haplotypes from short read pileups.
+Our model, called Jovian, frames variant detection as a sequence-to-sequence modeling task, akin to language translation, and employs a transformer-based architecture to translate alignment columns into two predicted haplotype sequences.
 After training with 17 whole genome sequences from Genome-in-a-Bottle cell lines, we demonstrate that this method predicts variant genotypes and phase accurately, has the ability to realign complex and ambiguous read mappings to produce accurate haplotype predictions, and uses the local read context to inform predictions about a given position.
@@
 HELLO (Ramachandran et al.
-2020), designed to work on hybrid short- and long-read datasets, employs a mixture-of-experts approach with separate 1-dimensional convolutions across the read and allele dimensions of the input.
+2020), designed to work on hybrid short- and long-read datasets, employs a mixture-of-experts approach with separate 1-dimensional convolutions across the read and position dimensions of the input.
 Clair (Luo et al.
@@
 2019).
-Input tokens are the collection of bases that align to a given reference position across reads, with some modifications described below.
+Input tokens are the collection of bases that align to a given reference position, with some modifications described below.
 We add an additional fully connected layer prior to the transformer encoders which embeds the encoded basecalls in $d$ dimensions, where $d=12$ for the analyses here.
 The embedded basecalls are `flattened' along the read dimension, producing an input token with size $dr$ where $r$ is the number of reads ($r=100$ for all experiments proposed here).
-Instead of the typical 1-dimensional positional encoding, we employed a 2-dimensional encoding, allowing the model to differentiate across both the token (position) and read dimensions.
+Instead of the typical 1-dimensional positional encoding, we employ a 2-dimensional encoding, allowing the model to differentiate across both the token (position) and read dimensions.
 The 2D encoding implementation follows Wang \& Jyh-Charn (2021).
 
-The decoder consists of a single fully-connected layer with ReLU activation followed by two additional fully connected components that each generate a single haplotype.
+The decoder consists of a single fully-connected layer with ReLU activation followed by two additional fully-connected components that each generate a single haplotype.
+The decoder is not autoregressive and does not utilize special start-of-sequence or end-of-sequence tokens.
 Predictions are discrete probability distributions over the four possible bases at each predicted output position, produced with a softmax function over the raw outputs.
 The decoder is of fixed length and produces two haplotype sequences equal in length to the number of input tokens $g$.
-The full model architecture is shown in \ref{fig:architecture}.
+The full model architecture is shown in Figure \ref{fig:architecture}.
 
@@
 	\includegraphics[scale=0.5]{./jovian_architecture_diagram.png}
-	\caption{ Overview of network architecture, numbers in parentheses indicate size of tensors  }
+	\caption{ Overview of network architecture, numbers in parentheses indicate size of tensors}
 	\label{fig:architecture}
@@
 
-Input tensors are generated by selecting a 150bp region from a set of aligned reads in BAM / CRAM format.
+Input tensors were generated by selecting a 150bp region from a set of aligned reads in BAM / CRAM format.
 For each region, all reads overlapping the window were obtained from the corresponding alignment file.
 Regions containing more than the maximum number of reads (100 for all analyses here) were downsampled to the maximum size.
-Reads are then sorted by the reference coordinate of the first aligned base.
+Reads were sorted by the reference coordinate of the first aligned base.
 For each aligned base in the selected reads nine features are encoded; the first four are the one-hot encoded base call, followed by base quality, two flags indicating if base `consumed' reference base (i.e.
@@
 
-Target / label haplotype sequences were produced by obtaining truth variants from the Genome-in-a-Bottle VCF files for each sample and inserting the variants into the reference sequence.
+Target haplotype sequences were produced by obtaining truth variants from the Genome-in-a-Bottle VCF files for each sample and inserting the variants into the reference sequence.
 Two sequences were generated for each region, one representing each haplotype.
@@
 \begin{center}
+	\begin{table}
 	\begin{tabular}{ lc }
@@
 	\end{tabular}
+	\caption{Number of regions selected from each sample to generate training data}
 	\label{table:regioncounts}
+\end{table}
 \end{center}
@@
 
-Models were trained for 50 epochs unless noted otherwise, using the Adam optimizer and a learning rate of $5e^{-5}$ and batch size of 512.
-Models were implemented in PyTorch (v1.10, Paszke et al 2019).
+Models were trained for 50 epochs unless noted otherwise, using the Adam optimizer, a learning rate of $5e^{-5}$, and batch size of 512. Models were implemented in PyTorch (v1.10, Paszke et al 2019).
 
 For a given haplotype we use a modified cross-entropy loss function to compare the predictions to the target sequence.
-However, two haplotypes exist for each region and they have no inherent order.
-For instance, a single heterozygous variant can appear on either haplotype, and the model cannot learn predict which because the assigment is random.
+For each region two haplotypes exist and they have no inherent order.
+For instance, a single heterozygous variant can appear on either haplotype, and the model cannot learn to predict which because the assigment is random.
 To make the predictions opaque to haplotype order, we compute the loss in both haplotype configurations - predicted haplotype 0 with target haplotype 0 and 1 with 1, as well as predicted haplotype 1 with target haplotype 0, and 0 with 1, and backpropogate gradients only for the configuration with the lowest loss.
@@
 We used the scikit-learn implementation of the random forest with 100 trees, a maximum tree depth of 25, and `balanced' class weighting option.
+Features are listed in Table \ref{table:features}.
 
 \begin{center}
+	\begin{table}
 	\begin{tabular}{ l }
@@
 	\label{table:features}
+	\caption{ Features used to train classifier model for estimation of variant quality scores }
+\end{table}
 \end{center}
@@
 \begin{center}
-	\begin{tabular}{ lllll }
-	 Model name & Encoder layers & Attention heads & Hidden dimension & Embedding dimension \\ 
+	\begin{table}
+	\begin{tabular}{ ccccc }
+	 Model name & Encoder layers & Attn. heads & Hidden dim. & Embedding dim. \\ 
 	\hline
@@
 	\end{tabular}
+	\caption{ Model parameter combinations used for the three model sizes investigated }
+\end{table}
 \end{center}
-
-During training, sensitivity and precision as measured on the holdout regions from chromosomes 21 and 22 increase for all variant classes, with returns dwindling after 30-40 epochs of training.
-Little difference was observed between the "full" and "half" sized training data sets, but both sensitivity and precision were lower for the quarter-sied training set.
-These results indicate that additional training examples beyond the "half" size ($12e^6$ example regions) offer little improvement in accuracy, and suggest that performance becomes limited by other factors, such as model capacity, inaccuracies in the training data set, or a non-optimal loss function.
 
@@
 
+During training, sensitivity and precision increased for all variant classes, with returns dwindling after 30-40 epochs of training (Figure \ref{fig:trainsize}).
+Little difference was observed between the Full and Half sized training data sets, but both sensitivity and precision were lower for the Quarter-sized training set.
+These results indicate that additional training examples beyond the Half size ($12e^6$ example regions) offer little improvement in accuracy, and suggest that performance becomes limited by other factors, such as model capacity, inaccuracies in the training data set, or a non-optimal loss function.
+
+\begin{figure}[htp]
+	\includegraphics[scale=0.5]{./modelsize_effect.png}
+	\caption{ Sensitivity and precision curves for the Half-sized data set, for the 25M, 10M, and 5M model sizes.
+Dashed lines are raw values, solid lines indicate rolling average with window size of 5.
+}
+	\label{fig:modelsize}
+\end{figure}
+
+Sensitivity was similar across all model sizes evaluated for all variant classes, but precision increased as model size was increased from 5M to 10M (Figure \ref{fig:modelsize}).
+However, precision did not increase further as model size was increased from 10M to 25M.
+Together with the training data size analysis, the results indicate that neither increasing model capacity beyond 10M parameters or training data size beyond approximately 12M examples results in improved model performance.
+
 \subsection{Variant detection accuracy}
@@
 2018).
-For this comparison we used 25M model trained for 25 epochs on the 'huge' training set (~$20^6$ training examples).
+For this comparison we used the 10M model trained on the Full set.
 We selected filtering criteria for each caller by reviewing ROC curves produced by \textit{vcfeval} and selecting values close to the $F_1$ maximizing value across samples.
-Jovian calls were filtered at 0.25, HaplotypeCaller at 60 (phred-scaled), Strelka at 4 (phred-scaled), and DeepVariant at 0.
+Jovian calls were filtered at quality 0.25, HaplotypeCaller at 60 (phred-scaled), Strelka at 4 (phred-scaled), and DeepVariant at 0.
 Sensitivity, precision, and related calculations were computed with the \textit{hap.py} tool (Krushe et al.
-2019) using Genome-in-a-Bottle (v.4.2.1) as the benchmark variant set.
+2019) using Genome-in-a-Bottle (v4.2.1) as the benchmark variant set.
 
@@
 	\includegraphics[scale=0.5]{./sensitivity-precision.png}
-	\caption{ Sensitivity and precision of Jovian variant calls (10M parameter model, large training size) compared to other callers }
+	\caption{ Sensitivity and precision of Jovian variant calls (10M parameter model, Full training size) compared to other callers }
 	\label{fig:sens-ppv}
@@
 
-Across all regions, Jovian's sensitivity and precision for indels is slightly higher than that for HaplotypeCaller, and slightly lower than DeepVariant or Strelka2 (\ref{fig:sens-ppv}).
+Across all regions, Jovian's sensitivity and precision for indels is slightly higher than that for HaplotypeCaller, and slightly lower than DeepVariant or Strelka2 (Figure \ref{fig:sens-ppv}).
 For SNV calls Jovian has the highest sensitivity (99.1\%) and precision similar to HaplotypeCaller (99.5\%), but precision is lower than DeepVariant and Strelka2 (99.8\% and 99.9\%, respectively).
 
-Across indel size classes Jovian has high sensitivity for deletions less than about 15bp, but relatively low sensitivity for larger deletions (~75\%).
-For insertions Jovian has sensitivity similar to other callers at smaller sizes, but again reduced sensitivity for the largest class (~68\%).
+Across indel size classes Jovian has high sensitivity for deletions less than about 15bp, but relatively low sensitivity for larger deletions ($\approx$75\%).
+For insertions Jovian has sensitivity similar to other callers at smaller sizes, but again reduced sensitivity for the largest class ($\approx$68\%).
 Precision follows a similar pattern, but with relatively high precision for smaller indel sizes and reduced precision at larger sizes.
@@
 	\includegraphics[scale=0.5]{./accuracy_indel_size.png}
-	\caption{ Sensitivity and precision of Jovian indel calls by size class (10M parameter model, large training size) compared to other callers }
+	\caption{ Sensitivity and precision of Jovian indel calls by size class (10M parameter model, Full training size) compared to other callers }
 	\label{fig:indelsize-sens-ppv}
@@
 To check the accuracy of predicted genotypes and phasing of variants we compared calls in the chromosome 21 and 22 evaluation regions to the known genotype and phase from Genome-in-a-Bottle (GIAB) variant data.
-Approximately 99.5\% of homozygous and 99.0\% of heterozygous calls matched the true genotype, indicating that our model learns to construct both genotypes accurately.
+Approximately 99.5\% of homozygous and 99.0\% of heterozygous calls matched the true genotype.
 For phasing accuracy, we identified pairs of heterozygous variants less than 100 base pairs apart and compared their phasing predictions with the known phasing information from GIAB.
@@
 Phase predictions matched GIAB phase for approximately 96.9\% of SNV pairs and 99.3\% of indel pairs.
+Together, these analyses indicate that the model learns to reconstruct variant phase and genotype accurately, producing pairs of haplotype sequences that match those present in the sample with high fidelity.
 
@@
 
-To identify variants requiring complex reconstruction, we first obtained all indel variants from Genome-in-a-Bottle greater than 4 bp from the validation regions on chromosomes 21 and 22.
-For each variant, we examine the reads in the alignment file that overlap the variant position, and flag any with with more than 4 distinct indel start sites or at least 20\% reads with more than 10bp soft-clipped as requiring 'complex' reconstruction.
-These variants are moved to a separate truth set.
-After running our variant calling procedure on the alignment file, we compute the fraction of variants from the truth set that were identified by our model.
+To identify variants requiring complex reconstruction, we first obtained all indel variants from Genome-in-a-Bottle greater than 4 bp in length from the validation regions on chromosomes 21 and 22.
+For each variant, we examine the reads in the alignment file that overlap the variant position, and flag any with more than 4 distinct indel start sites or at least 20\% reads with more than 10bp soft-clipped as requiring `complex' reconstruction.
+These variants were moved to a separate truth set.
+After running our variant calling procedure on the alignment file, we computed the fraction of variants from the truth set that were identified by our model.
 This procedure was repeated for all 17 samples in our data set.
 
-Our model was able to reconstruct the correct haplotypes for ~89.5\% (mid 50\% 82.5-94.8) of these variants, somewhat lower than other callers (\ref{table:complex-sensitivity}).
+Our model was able to reconstruct the correct haplotypes for $\approx$89.5\% (mid 50\% 82.5-94.8) of these variants, somewhat lower than other callers (table \ref{table:complex-sensitivity}).
 These results indicate that the model learns a reconstruction procedure more nuanced than a simple column-by-column examination of reads.
@@
 \begin{center}
+\begin{table}	 
 	\begin{tabular}{ ccc }
@@
 	 \hline
+
 	\end{tabular}
 	\label{table:complex-sensitivity}
+	\caption{ Sensitivity of detection algorithms for variants requiring complex reconstruction }
+\end{table}
 \end{center}
 
-\subsection{Use of local context}
+\subsection{Use of local context for SNV prediction}
 
-We were interested to see how model variant calling was informed by contextual information from nearby alignment columns.
-In particular, the model must learn to discriminate real variation from noise at low Variant Allele Frequency (VAF, the fraction of reads supporting a given allele candidate), and this model decision process may involve the use of local context information.
-For example, the model may be less likely to call a SNV in a region containing many reference mismatches, because the presence of multiple mismatches suggests poor sequencing quality or other technical artifacts.
-However, the call probability is likely to be modified by other factors, including VAF and the base call quality associated with the prospective variant.
-
-To examine the influence of the number of nearby mismatches, VAF, and base quality on the likelihood that a prospective SNV was called, we first collected genomic positions where 10 - 30\% of reads supported a non-reference basecall.
+Variant callers must integrate multiple types of information to produce accurate calls, including the fraction of reads supporting a given call (Variant Allele Frequency, VAF),base and mapping qualities, read position, and potentially information from nearby positions.
+To examine the influence of the number of nearby reference mismatches, VAF, and base quality on the likelihood that a prospective SNV was called, we first collected genomic positions in which 10 - 30\% of reads supported a non-reference basecall.
 We then tallied the number of genomic positions at which 10\% - 20\% of reads supported a non-reference call in a window 20bp upstream and downstream of the candidate.
@@
 \begin{figure}[htp]
-	\includegraphics[scale=0.25]{./variant_context_BQ.png}
+	\includegraphics[scale=0.22]{./variant_context_BQ.png}
 	\caption{ Effect of number of nearby reference mismatches on the likelihood that a poorly supported candidate variant will be called. }
@@
 When the prospective variant has poor base quality scores, the association disappears and the likelihood that the variant is called remains uniformly low.
-Together, these results indicate that the models learns a nuanced SNV detection procedure that incorporates multiple elements, including base quality, VAF, and information from local surrounding bases to discriminate actual genetic variation from technical artifacts and sequencing errors.
+Overall, these results indicate that the model learns a nuanced SNV detection procedure that discards most prospective SNVs with low base quality, but selectively retains those with higher base quality provided that not too many nearby positions contain reference mismatches.
 
@@
 To assess the impact of the two-dimensional positional encoding, we replaced the 2D encoding module with traditional 1D encoding, and left other model parameters unmodified.
-We conducted two training runs, one with the 10M model and smaller training data size, and the other with the 25M model and medium training data size.
+We conducted two training runs, one with the 10M model and Quarter training data size, and the other with the 25M model and Half training data size.
 Both training runs were conducted for 25 epochs.
@@
 
-Two-dimensional encoding allows the model to track which read contains a given base across genomic positions.
+By applying positional encoding across both the read and position (input token) dimensions, two-dimensional encoding allows the model to track which read contains a given base across positions.
 We hypothesize that read-level tracking is important for precision because mismapped or poorly aligned reads generate many reference mismatches, most of which are likely to be errors.
@@
 For instance, our model learns to reconstruct large, complex variants in the presence of ambiguous mapping and extensive soft clipping with a sensitivity near 90\%.
-In addition, the model learns to employ local context and read-specific features to predict phase and genotype accurately, and pairs of variants are phased correctly in ~97-99\% of cases where variants are within 100bp.
+In addition, the model learns to employ local context and read-specific features to predict phase and genotype accurately, and pairs of variants are phased correctly in 97-99\% of cases where variants are within 100bp.
 The model also learns to consider the surrounding read context when predicting the presence of a SNV.
@@
 Overall variant detection accuracy is similar to other modern callers.
-At our chosen quality cutoffs (selected to be near the $F_1$-maximizing value for each caller), Jovian's performance on insertion-deletion variants outperforms HaplotypeCaller, but falls short of DeepVariant and Strelka2 (\ref{fig:sens-ppv}).
+At our chosen quality cutoffs (selected to be near the $F_1$-maximizing value for each caller), Jovian's performance on insertion-deletion variants outperforms HaplotypeCaller, but falls short of DeepVariant and Strelka2 (Figure \ref{fig:sens-ppv}).
 The differences are partially due to Jovian's relatively poor sensitivity for indels larger than 15bp (mean 75.6\% for deletions, 68.8\% for insertions) and low sensitivity in tandem repeat regions (mean sensitivity 84.0\%).
@@
 
-In this work we have used a standard cross-entropy loss function that simply compares each prediction at base $i$ to the label at base $i$.
+In this work we have used a standard cross-entropy loss function that simply compares each prediction at base $i$ to the label at base $i$ for each haplotype.
 While training with this loss yields accurate results, it has two shortcomings.
@@
 
-For this work we have employed an unmodified transformer model with a very simple decoder consisting of just two fully connected layers, and we leave to future work exploration of the many transformer variations proposed in recent years (e.g.
+Our model uses an unmodified transformer encoder with a very simple decoder consisting of just two fully connected layers, and we leave to future work exploration of the many transformer variations proposed in recent years (e.g.
 Dosovitskiy 2018, Fedus et al.
@@
 
-In his essay `The Bitter Lesson' (Sutton 2019), noted researcher Richard Sutton commented that methods that leverage computation and search are often more successful in the long term when compared to methods that directly model human knowledge.
-Top performing chess engines, for instance, do not directly contain rules or strategies developed by humans and instead play millions of games to discover a successful approach.
-In the context of genomic variant discovery, approaches that enable machines to \textit{learn} the correct method for assembling reads are likely to yield better results.
-
-Limitations
- - Low SNV precision
- - not using strand bias?
- -
-
 \section{Availability}
 
- Source code for is available via git at https://github.com/ARUP-NGS/somewhere
+ Source code for is available via git at https://github.com/brendanofallon/jovian
 

### 2022-09-01 manuscript/manuscript.tex
dockerfile, readme, remove ssw_aligner dep

@@
 2009) and the UnifiedGenotyper tool from the Genome Analysis ToolKit (GATK, De Pristo et al.
-2011) rely on the read aligner to generate candidate allele candidates, and utilize several ad-hoc heuristics and thresholds to determine likely alleles (e.g.
+2011) rely on the read aligner to generate candidate alleles, and utilize several ad-hoc heuristics and thresholds to determine which alleles are most likely (e.g.
 Nielsen et al.
@@
 For this special row we inserted a base quality of 100 for every position and did not set any of the other flags.
-Resulting tensors had dimension $[g, r, 9]$, where $g$ represents genomic position, $r$ represents read, and 9 is the number of features.
+Resulting tensors had dimension $[g, r, 9]$, where $g$ indexes genomic positions, $r$ indexes reads, and 9 is the number of features.
 Positions in the input not corresponding to an aligned base were all set to zero.

### 2022-09-01 dnaseq2seq/call.py

@@ -1,7 +1,18 @@
-import itertools
+"""
+This program is free software: you can redistribute it and/or modify it under the terms of the GNU General Public
+License as published by the Free Software Foundation, either version 3 of the License, or (at your option) any
+later version.
+
+This program is distributed in the hope that it will be useful, but WITHOUT ANY WARRANTY; without even the implied
+warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the GNU General Public License for more details.
+You should have received a copy of the GNU General Public License along with this program.
+If not, see <https://www.gnu.org/licenses/>.
+"""
+
+
 import os
 import time
-import math
+
 import datetime
 import logging
 import string
@@ -15,7 +26,6 @@ from concurrent.futures import ProcessPoolExecutor
 import torch
 import torch.multiprocessing as mp
 import pysam
-import pyranges as pr
 import pandas as pd
 
 from dnaseq2seq.model import VarTransformer

### 2022-09-01 dnaseq2seq/loader.py

@@ -1,4 +1,15 @@
 
+"""
+This program is free software: you can redistribute it and/or modify it under the terms of the GNU General Public
+License as published by the Free Software Foundation, either version 3 of the License, or (at your option) any
+later version.
+
+This program is distributed in the hope that it will be useful, but WITHOUT ANY WARRANTY; without even the implied
+warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the GNU General Public License for more details.
+You should have received a copy of the GNU General Public License along with this program.
+If not, see <https://www.gnu.org/licenses/>.
+"""
+
 import logging
 
 logger = logging.getLogger(__name__)

### 2022-09-01 dnaseq2seq/model.py

@@ -1,4 +1,14 @@
 
+"""
+This program is free software: you can redistribute it and/or modify it under the terms of the GNU General Public
+License as published by the Free Software Foundation, either version 3 of the License, or (at your option) any
+later version.
+
+This program is distributed in the hope that it will be useful, but WITHOUT ANY WARRANTY; without even the implied
+warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the GNU General Public License for more details.
+You should have received a copy of the GNU General Public License along with this program.
+If not, see <https://www.gnu.org/licenses/>.
+"""
 
 import logging
 

### 2022-09-01 dnaseq2seq/swloss.py

@@ -1,4 +1,15 @@
 
+"""
+This program is free software: you can redistribute it and/or modify it under the terms of the GNU General Public
+License as published by the Free Software Foundation, either version 3 of the License, or (at your option) any
+later version.
+
+This program is distributed in the hope that it will be useful, but WITHOUT ANY WARRANTY; without even the implied
+warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the GNU General Public License for more details.
+You should have received a copy of the GNU General Public License along with this program.
+If not, see <https://www.gnu.org/licenses/>.
+"""
+
 import torch
 import torch.nn as nn
 

### 2022-09-01 dnaseq2seq/vcf.py

@@ -1,8 +1,9 @@
 
 import numpy as np
-import ssw_aligner
 from dataclasses import dataclass
 import pysam
+
+from skbio.alignment import StripedSmithWaterman
 
 @dataclass
 class Variant:
@@ -91,13 +92,12 @@ def align_sequences(query, target):
     """
     Return Smith-Watterman alignment of both sequences
     """
-    aln = ssw_aligner.local_pairwise_align_ssw(query,
-                                               target,
-                                               gap_open_penalty=3,
-                                               gap_extend_penalty=1,
-                                               match_score=2,
-                                               mismatch_score=-1)
-    return aln
+    ssw = StripedSmithWaterman(query,
+                               gap_open_penalty=3,
+                               gap_extend_penalty=1,
+                               match_score=2,
+                               mismatch_score=-1)
+    return ssw(target)
 
 
 def _mismatches_to_vars(query, target, offset, probs):
@@ -475,3 +475,8 @@ def vars_to_vcf(vcf_file, variants):
     for var in variants:
         r = create_vcf_rec(var, vcf_file)
         vcf_file.write(r)
+
+
+if __name__=="__main__":
+    aln = align_sequences("ACTGACTG", "ACGACTG")
+    print(aln)

### 2022-09-01 setup.py

@@ -101,7 +101,7 @@ setup(
     license="",
     install_requires=requires,
     scripts=[
-        "dnaseq2seq/bin/main.py",
+        "dnaseq2seq/main.py",
     ],
     cmdclass={"install": DNAseq2seqInstallCommand},
     tests_require=["pytest"],

### 2022-09-02 dnaseq2seq/main.py
Add file back in

@@ -1,0 +1,236 @@
+#!/usr/bin/env python
+
+"""
+This program is free software: you can redistribute it and/or modify it under the terms of the GNU General Public
+License as published by the Free Software Foundation, either version 3 of the License, or (at your option) any
+later version.
+
+This program is distributed in the hope that it will be useful, but WITHOUT ANY WARRANTY; without even the implied
+warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the GNU General Public License for more details.
+You should have received a copy of the GNU General Public License along with this program.
+If not, see <https://www.gnu.org/licenses/>.
+"""
+
+import sys
+import logging
+import random
+from collections import defaultdict
+from pathlib import Path
+from string import ascii_letters, digits
+import lz4.frame
+import tempfile
+from datetime import datetime
+import re
+import torch
+
+from concurrent.futures.process import ProcessPoolExecutor
+import argparse
+import yaml
+
+from dnaseq2seq import util
+from dnaseq2seq import loader
+from dnaseq2seq.train import train
+from dnaseq2seq.call import call
+
+
+logging.basicConfig(format='[%(asctime)s]  %(name)s  %(levelname)s  %(message)s',
+                    datefmt='%m-%d %H:%M:%S',
+                    level=logging.INFO) # handlers=[RichHandler()])
+logger = logging.getLogger(__name__)
+
+DEVICE = torch.device("cuda") if hasattr(torch, 'cuda') and torch.cuda.is_available() else torch.device("cpu")
+
+def load_conf(confyaml):
+    logger.info(f"Loading configuration from {confyaml}")
+    conf = yaml.safe_load(open(confyaml).read())
+    assert 'reference' in conf, "Expected 'reference' entry in training configuration"
+    assert 'data' in conf, "Expected 'data' entry in training configuration"
+    return conf
+
+
+
+def default_vals_per_class():
+    """
+    Multiprocess will instantly deadlock if a lambda or any callable not defined on the top level of the module is given
+    as the 'factory' argument to defaultdict - but we have to give it *some* callable that defines the behavior when the key
+    is not present in the dictionary, so this returns the default "vals_per_class" if a class is encountered that is not 
+    specified in the configuration file. I don't think there's an easy way to make this user-settable, unfortunately
+    """
+    return 0
+
+
[…]

### 2022-09-02 dnaseq2seq/call.py
10m params

@@ -131,16 +131,16 @@ def reconcile_current_window(prev_win, current_win):
 def load_model(model_path):
     
     # 25m
-    attention_heads = 8
-    encoder_layers = 8
-    transformer_dim = 400
-    embed_dim_factor = 100
+    #attention_heads = 8
+    #encoder_layers = 8
+    #transformer_dim = 400
+    #embed_dim_factor = 100
     
     # 10m
-    #attention_heads = 6
-    #encoder_layers = 6
-    #transformer_dim = 250
-    #embed_dim_factor = 100
+    attention_heads = 6
+    encoder_layers = 6
+    transformer_dim = 250
+    embed_dim_factor = 100
     model = VarTransformer(read_depth=100,
                            feature_count=9,
                            out_dim=4,

### 2022-09-02 dnaseq2seq/main.py

@@ -18,6 +18,7 @@ from collections import defaultdict
 from pathlib import Path
 from string import ascii_letters, digits
 import lz4.frame
+import os
 import tempfile
 from datetime import datetime
 import re
@@ -212,8 +213,8 @@ def main():
     trainparser.set_defaults(func=train)
 
     callparser = subparser.add_parser("call", help="Call variants")
-    callparser.add_argument("-m", "--model-path", help="Stored model", required=True)
-    callparser.add_argument("-c", "--classifier-path", help="Stored variant classifier model", default=None, type=str)
+    callparser.add_argument("-m", "--model-path", help="Stored model", default=os.getenv('JOVIAN_MODEL'))
+    callparser.add_argument("-c", "--classifier-path", help="Stored variant classifier model", default=os.getenv('JOVIAN_CLASSIFIER'))
     callparser.add_argument("-r", "--reference-fasta", help="Path to Fasta reference genome", required=True)
     callparser.add_argument("-b", "--bam", help="Input BAM file", required=True)
     callparser.add_argument("-d", "--bed", help="bed file defining regions to call", required=False)

### 2022-09-06 manuscript/manuscript.tex
Manuscript update, main.py

@@
 Our model, called Jovian, frames variant detection as a sequence-to-sequence modeling task, akin to language translation, and employs a transformer-based architecture to translate alignment columns into two predicted haplotype sequences.
-After training with 17 whole genome sequences from Genome-in-a-Bottle cell lines, we demonstrate that this method predicts variant genotypes and phase accurately, has the ability to realign complex and ambiguous read mappings to produce accurate haplotype predictions, and uses the local read context to inform predictions about a given position.
+After training with 17 whole genome sequences from Genome-in-a-Bottle cell lines, we demonstrate that this method learns to realign complex and ambiguous read mappings to produce accurate haplotype predictions, predicts variant genotypes and phase accurately, and leverages the local read context to inform predictions about a given position.
 We also demonstrate that a 2-dimensional positional encoding significantly improved precision of the detected variants.

### 2022-09-06 dnaseq2seq/main.py

@@ -170,7 +170,7 @@ def alphanumeric_no_spaces(name):
 
 
 def main():
-    parser = argparse.ArgumentParser()
+    parser = argparse.ArgumentParser(description='NGS variant detection with')
     subparser = parser.add_subparsers()
 
     genparser = subparser.add_parser("pregen", help="Pre-generate tensors from BAMs")
@@ -224,11 +224,11 @@ def main():
     callparser.set_defaults(func=call)
 
     args = parser.parse_args()
-    args.cmdline = " ".join(sys.argv[1:])
-    args.cl_args = vars(args).copy()  # command line copy for logging
-    if len(vars(args)) == 0 or getattr(args, 'func') is None:
-        print(args.usage())
+    if len(vars(args)) == 0 or hasattr(args, 'func'):
+        print(parser.usage)
     else:
+        args.cmdline = " ".join(sys.argv[1:])
+        args.cl_args = vars(args).copy()  # command line copy for logging
         args.func(**vars(args))
 
 

### 2022-09-06 dnaseq2seq/main.py
Fix help msg

@@ -170,7 +170,7 @@ def alphanumeric_no_spaces(name):
 
 
 def main():
-    parser = argparse.ArgumentParser(description='NGS variant detection with')
+    parser = argparse.ArgumentParser(description='NGS variant detection with transformers')
     subparser = parser.add_subparsers()
 
     genparser = subparser.add_parser("pregen", help="Pre-generate tensors from BAMs")
@@ -224,8 +224,8 @@ def main():
     callparser.set_defaults(func=call)
 
     args = parser.parse_args()
-    if len(vars(args)) == 0 or hasattr(args, 'func'):
-        print(parser.usage)
+    if len(vars(args)) == 0 or not hasattr(args, 'func'):
+        parser.print_help()
     else:
         args.cmdline = " ".join(sys.argv[1:])
         args.cl_args = vars(args).copy()  # command line copy for logging

### 2022-10-14 manuscript/manuscript.tex
No vaftgt or parallel decomp

@@
 \title{Jovian enables direct inference of germline haplotypes from short reads via sequence-to-sequence modeling}
-\author[1]{Brendan O'Fallon}
-\author[1]{Ashini Bolia}
-\author[1]{Jacob Durtschi}
-\author[1]{Luobin Yang}
-\author[1]{Eric Frederickson}
-\author[1]{Katherine Noble}
-\author[1]{Joshua Coleman}
-\author[1]{Hunter Best}
+\author{Author names witheld during review}
+%\author[1]{Brendan O'Fallon}
+%\author[1]{Ashini Bolia}
+%\author[1]{Jacob Durtschi}
+%\author[1]{Luobin Yang}
+%\author[1]{Eric Frederickson}
+%\author[1]{Katherine Noble}
+%\author[1]{Joshua Coleman}
+%\author[1]{Hunter Best}
 
-\affil[1]{ARUP Institute for Clinical and Experimental Pathology, Salt Lake City, UT}
+%\affil[1]{ARUP Institute for Clinical and Experimental Pathology, Salt Lake City, UT}
 \date{}
@@
 Our model, called Jovian, frames variant detection as a sequence-to-sequence modeling task, akin to language translation, and employs a transformer-based architecture to translate alignment columns into two predicted haplotype sequences.
-After training with 17 whole genome sequences from Genome-in-a-Bottle cell lines, we demonstrate that this method learns to realign complex and ambiguous read mappings to produce accurate haplotype predictions, predicts variant genotypes and phase accurately, and leverages the local read context to inform predictions about a given position.
+After training with 17 whole genome sequences from Genome-in-a-Bottle cell lines, we demonstrate that this method learns to realign complex and ambiguous read mappings to produce accurate haplotype predictions, infers variant genotype and phase accurately, and leverages the local read context to inform predictions about a given position.
 We also demonstrate that a 2-dimensional positional encoding significantly improved precision of the detected variants.
-Compared to other callers, sensitivity and precision is higher than GATK HaplotypeCaller, but lower than DeepVariant and Strelka2.
+Compared to other callers, sensitivity and precision are higher than GATK HaplotypeCaller, but lower than DeepVariant and Strelka2.
 \end{abstract}
@@
 	 Fraction of reads in BAM/CRAM supporting variant call \\
-	 Fraction of reads with mapping quality > 10 supporting variant call \\
+	 Fraction of reads with mapping quality $>$ 10 supporting variant call \\
 	 Total number of reads supporting variant call \\
@@
 2021).
-Our simple decoder has a fixed length equal to the number of input tokens (genomic positions), but a natural improvement would be to add a more sophisticated decoder capable of producing a variable length sequence to handle larger insertion and deletion events.
+Our simple decoder has a fixed length equal to the number of input tokens (genomic positions), but a natural improvement would be to add a more sophisticated autoregressive decoder capable of producing a variable length sequence to handle larger insertion and deletion events.
 In addition, newer transformer models that avoid the $O(n^2)$ scaling with the number of input tokens may be able to handle longer input sequence lengths, and therefore may be able to detect larger variants.

### 2022-10-14 dnaseq2seq/loader.py

@@ -136,11 +136,13 @@ def decompress_multi(paths, threads):
     :returns : List of Tensors (all on CPU)
     """
     start = datetime.now()
-    with mp.Pool(threads) as pool:
-        result = pool.map(decomp_single, paths)
-        elapsed = datetime.now() - start
-        logger.info(f"Decompressed {len(result)} items in {elapsed.total_seconds():.3f} seconds ({elapsed.total_seconds()/len(result):.3f} secs per item)")
-        return result
+
+    result = [decomp_single(p) for p in paths]
+    elapsed = datetime.now() - start
+    logger.info(
+        f"Decompressed {len(result)} items in {elapsed.total_seconds():.3f} seconds ({elapsed.total_seconds() / len(result):.3f} secs per item)"
+    )
+    return result
 
 
 class PregenLoader:
@@ -164,8 +166,8 @@ class PregenLoader:
         if pathpairs:
             self.pathpairs = pathpairs
         else:
-            self.pathpairs = util.find_files(self.datadir, self.src_prefix, self.tgt_prefix, self.vaftgt_prefix)
-            self.pathpairs = random.shuffle(self.pathpairs)
+            self.pathpairs = util.find_files(self.datadir, self.src_prefix, self.tgt_prefix)
+            random.shuffle(self.pathpairs)
         self.threads = threads
         self.max_decomped = max_decomped_batches # Max number of decompressed items to store at once - increasing this uses more memory, but allows increased parallelization
         logger.info(f"Creating PreGen data loader with {self.threads} threads")
@@ -197,7 +199,7 @@ class PregenLoader:
         sequentially
         :param batch_size: The number of samples in a minibatch.
         """
-        src, tgt, vaftgt = [], [], []
+        src, tgt = [], []
         for i in range(0, len(self.pathpairs), self.max_decomped):
             decomp_start = datetime.now()
             paths = self.pathpairs[i:i+self.max_decomped]
@@ -205,10 +207,10 @@ class PregenLoader:
             decomp_end = datetime.now()
             decomp_time = (decomp_end - decomp_start).total_seconds()
 
-            for j in range(0, len(decomped), 3):
+            for j in range(0, len(decomped), 2):
                 src.append(decomped[j])
                 tgt.append(decomped[j+1])
-                vaftgt.append(decomped[j+2])
+                # vaftgt.append(decomped[j+2])
 
             total_size = sum([s.shape[0] for s in src])
[…]

### 2022-10-14 dnaseq2seq/util.py

@@ -45,7 +45,7 @@ def concat_metafile(sample_metafile, dest_metafh):
     os.unlink(sample_metafile)
 
 
-def find_files(datadir, src_prefix='src', tgt_prefix='tgt', vaftgt_prefix='vaftgt'):
+def find_files(datadir, src_prefix='src', tgt_prefix='tgt'):
     """
     Examine files in datadir and match up all src / tgt / vaftgt files and store them as tuples in a list
     :returns : List of (src, tgt, vaftgt) tuples of matched files
@@ -57,7 +57,6 @@ def find_files(datadir, src_prefix='src', tgt_prefix='tgt', vaftgt_prefix='vaftg
         suffix = src.name.split("_")[-1]
         pairs.append((src,
                      f"{datadir}/{tgt_prefix}_{suffix}",
-                     f"{datadir}/{vaftgt_prefix}_{suffix}"
                       ))
     return pairs
 

### 2022-10-14 scripts/run_train_clone.sh
tr scr

@@ -48,8 +48,8 @@ LEARNING_RATE=0.00005
 
 CHECKPOINT_FREQ=1
 
-RUN_NAME="wgs_decoder_biggermodel"
-RUN_NOTES="Decoder model 8 encoder and 6 decoder layers"
+RUN_NAME="wgs_decoder_biggermodel_cont"
+RUN_NOTES="Decoder model 8 encoder and 6 decoder layers, continued"
 
 set -x
 
@@ -84,6 +84,7 @@ $PYTHON $ds2s train \
     --checkpoint-freq $CHECKPOINT_FREQ \
     -o ${RUN_NAME}.model \
     --threads 4 \
+    -i /uufs/chpc.utah.edu/common/home/arup-storage3/u0379426/variant_transformer_runs/wgs_decoder_biggermodel/wgs_decoder_biggermodel_epoch6.model \
     --max-decomp-batches 4 \
     --wandb-run-name $RUN_NAME \
     --wandb-notes "$RUN_NOTES"

### 2022-10-14 dnaseq2seq/loader.py
Use a processpool

@@ -19,11 +19,10 @@ import math
 from pathlib import Path
 from collections import defaultdict
 from itertools import chain
-import gzip
 import lz4.frame
 from datetime import datetime
 import traceback as tb
-
+from concurrent.futures import ProcessPoolExecutor
 import io
 
 import scipy.stats as stats
@@ -31,6 +30,7 @@ import numpy as np
 import torch
 import torch.multiprocessing as mp
 import pysam
+
 
 from bam import (
     target_string_to_tensor,
@@ -136,8 +136,17 @@ def decompress_multi(paths, threads):
     :returns : List of Tensors (all on CPU)
     """
     start = datetime.now()
+    result = []
+    futs = []
+    with ProcessPoolExecutor(threads) as pool:
+        for path in paths:
+            fut = pool.submit(decomp_single, path)
+            futs.append(fut)
 
-    result = [decomp_single(p) for p in paths]
+    for f in fut:
+        result.append(f.result(timeout=120))
+
+    # result = [decomp_single(p) for p in paths]
     elapsed = datetime.now() - start
     logger.info(
         f"Decompressed {len(result)} items in {elapsed.total_seconds():.3f} seconds ({elapsed.total_seconds() / len(result):.3f} secs per item)"

### 2022-10-14 dnaseq2seq/loader.py
Merge in new processpoolstuff

@@ -122,11 +122,8 @@ class WeightedLoader:
 
 
 def decomp_single(path):
-    if str(path).endswith('.lz4'):
-        with open(path, 'rb') as fh:
-            return torch.load(io.BytesIO(lz4.frame.decompress(fh.read())), map_location='cpu')
-    else:
-        return torch.load(path, map_location='cpu')
+    with open(path, 'rb') as fh:
+        return lz4.frame.decompress(fh.read())
 
 
 def decompress_multi(paths, threads):
@@ -143,10 +140,13 @@ def decompress_multi(paths, threads):
             fut = pool.submit(decomp_single, path)
             futs.append(fut)
 
-    for f in fut:
-        result.append(f.result(timeout=120))
+    for f in futs:
+        data = f.result(timeout=20)
+        result.append(torch.load(io.BytesIO(data), map_location='cpu'))
 
-    # result = [decomp_single(p) for p in paths]
+    #for i, r in enumerate(result):
+    #    logger.info(f"Result item {i}: {r.shape}")
+        
     elapsed = datetime.now() - start
     logger.info(
         f"Decompressed {len(result)} items in {elapsed.total_seconds():.3f} seconds ({elapsed.total_seconds() / len(result):.3f} secs per item)"
@@ -169,7 +169,6 @@ class PregenLoader:
         self.datadir = Path(datadir) if datadir else None
         self.src_prefix = src_prefix
         self.tgt_prefix = tgt_prefix
-        self.vaftgt_prefix = vaftgt_prefix
         if pathpairs and datadir:
             raise ValueError(f"Both datadir and pathpairs specified for PregenLoader - please choose just one")
         if pathpairs:
@@ -220,7 +219,6 @@ class PregenLoader:
             for j in range(0, len(decomped), 2):
                 src.append(decomped[j])
                 tgt.append(decomped[j+1])
-                # vaftgt.append(decomped[j+2])
 
             total_size = sum([s.shape[0] for s in src])
             if total_size < batch_size:
@@ -230,7 +228,6 @@ class PregenLoader:
             # Make a big tensor.
             src_t = torch.cat(src, dim=0)
             tgt_t = torch.cat(tgt, dim=0)
-            # vaftgt_t = torch.cat(vaftgt, dim=0)
 
             nbatch = total_size // batch_size
             remain = total_size % batch_size
@@ -253,7 +250,6 @@ class PregenLoader:
                 # The remaining data points will be in next batch. 
[…]

### 2022-10-14 dnaseq2seq/train.py

@@ -447,7 +447,7 @@ def train_epochs(epochs,
 
     if val_dir:
         logger.info(f"Using validation data in {val_dir}")
-        val_loader = loader.PregenLoader(device=DEVICE, datadir=val_dir, threads=4, tgt_prefix="tgkmers")
+        val_loader = loader.PregenLoader(device=DEVICE, datadir=val_dir, threads=8, tgt_prefix="tgkmers")
     else:
         logger.info(f"No val. dir. provided retaining a few training samples for validation")
         valpaths = dataloader.retain_val_samples(fraction=0.05)

### 2022-10-14 dnaseq2seq/train.py
max decomp batches = threads/2

@@ -447,7 +447,7 @@ def train_epochs(epochs,
 
     if val_dir:
         logger.info(f"Using validation data in {val_dir}")
-        val_loader = loader.PregenLoader(device=DEVICE, datadir=val_dir, threads=8, tgt_prefix="tgkmers")
+        val_loader = loader.PregenLoader(device=DEVICE, datadir=val_dir, max_decomped_batches=4, threads=8, tgt_prefix="tgkmers")
     else:
         logger.info(f"No val. dir. provided retaining a few training samples for validation")
         valpaths = dataloader.retain_val_samples(fraction=0.05)

### 2022-10-14 scripts/run_train_clone.sh

@@ -6,7 +6,7 @@
 #SBATCH --account=arup-gpu-np
 #SBATCH --partition=arup-gpu-np
 #SBATCH --mem=64G
-#SBATCH --cpus-per-task=4
+#SBATCH --cpus-per-task=8
 #SBATCH --time=5-0
 #SBATCH --mail-type=END,FAIL
 #SBATCH --mail-user=brendan.ofallon@aruplab.com

### 2022-10-14 dnaseq2seq/loader.py
try new method

@@ -126,7 +126,7 @@ def decomp_single(path):
         return lz4.frame.decompress(fh.read())
 
 
-def decompress_multi(paths, threads):
+def decompress_multi_ppe(paths, threads):
     """
     Read & decompress all of the items in the paths with lz4, then load them into Tensors
     but keep them on the CPU
@@ -146,6 +146,24 @@ def decompress_multi(paths, threads):
 
     #for i, r in enumerate(result):
     #    logger.info(f"Result item {i}: {r.shape}")
+        
+    elapsed = datetime.now() - start
+    logger.info(
+        f"Decompressed {len(result)} items in {elapsed.total_seconds():.3f} seconds ({elapsed.total_seconds() / len(result):.3f} secs per item)"
+    )
+    return result
+
+def decompress_multi_map(paths, threads):
+    """
+    Read & decompress all of the items in the paths with lz4, then load them into Tensors
+    but keep them on the CPU
+    :returns : List of Tensors (all on CPU)
+    """
+    start = datetime.now()
+    result = []
+    with mp.Pool(threads) as pool:
+        result = pool.map(decomp_single, path)
+
         
     elapsed = datetime.now() - start
     logger.info(
@@ -180,6 +198,8 @@ class PregenLoader:
         self.max_decomped = max_decomped_batches # Max number of decompressed items to store at once - increasing this uses more memory, but allows increased parallelization
         logger.info(f"Creating PreGen data loader with {self.threads} threads")
         logger.info(f"Found {len(self.pathpairs)} batches in {datadir}")
+        logger.info(f"Possible sharing strategies: {mp.get_all_sharing_strategies()}")
+        logger.info(f"Current sharing strategy: {mp.get_sharing_strategy()}")
         if not self.pathpairs:
             raise ValueError(f"Could not find any files in {datadir}")
 
@@ -212,7 +232,7 @@ class PregenLoader:
         for i in range(0, len(self.pathpairs), self.max_decomped):
             decomp_start = datetime.now()
             paths = self.pathpairs[i:i+self.max_decomped]
-            decomped = decompress_multi(chain.from_iterable(paths), self.threads)
+            decomped = decompress_multi_map(chain.from_iterable(paths), self.threads)
             decomp_end = datetime.now()
             decomp_time = (decomp_end - decomp_start).total_seconds()
 

### 2022-10-14 scripts/run_train_clone.sh

@@ -48,8 +48,8 @@ LEARNING_RATE=0.00005
 
 CHECKPOINT_FREQ=1
 
-RUN_NAME="wgs_decoder_biggermodel_cont"
-RUN_NOTES="Decoder model 8 encoder and 6 decoder layers, continued"
+RUN_NAME="test_deadlock4"
+RUN_NOTES="deadlock testing"
 
 set -x
 
@@ -72,7 +72,7 @@ cd ..
 
 echo "Branch: $GIT_BRANCH \n commit: $COMMIT \n" >> git_info.txt
 
-export ENABLE_WANDB=1
+export ENABLE_WANDB=
 
 $PYTHON $ds2s train \
     -c $CONF \
@@ -83,7 +83,7 @@ $PYTHON $ds2s train \
     --learning-rate $LEARNING_RATE \
     --checkpoint-freq $CHECKPOINT_FREQ \
     -o ${RUN_NAME}.model \
-    --threads 4 \
+    --threads 8 \
     -i /uufs/chpc.utah.edu/common/home/arup-storage3/u0379426/variant_transformer_runs/wgs_decoder_biggermodel/wgs_decoder_biggermodel_epoch6.model \
     --max-decomp-batches 4 \
     --wandb-run-name $RUN_NAME \

### 2022-10-14 dnaseq2seq/loader.py
try new method

@@ -162,7 +162,7 @@ def decompress_multi_map(paths, threads):
     start = datetime.now()
     result = []
     with mp.Pool(threads) as pool:
-        result = pool.map(decomp_single, path)
+        result = pool.map(decomp_single, paths)
 
         
     elapsed = datetime.now() - start

### 2022-10-14 dnaseq2seq/loader.py
try new method

@@ -153,6 +153,7 @@ def decompress_multi_ppe(paths, threads):
     )
     return result
 
+
 def decompress_multi_map(paths, threads):
     """
     Read & decompress all of the items in the paths with lz4, then load them into Tensors
@@ -160,11 +161,12 @@ def decompress_multi_map(paths, threads):
     :returns : List of Tensors (all on CPU)
     """
     start = datetime.now()
-    result = []
+    decompressed = []
     with mp.Pool(threads) as pool:
-        result = pool.map(decomp_single, paths)
-
-        
+        decompressed = pool.map(decomp_single, paths)
+    
+    result = [torch.load(io.BytesIO(d), map_location='cpu') for d in decompressed]
+           
     elapsed = datetime.now() - start
     logger.info(
         f"Decompressed {len(result)} items in {elapsed.total_seconds():.3f} seconds ({elapsed.total_seconds() / len(result):.3f} secs per item)"

### 2022-10-14 dnaseq2seq/loader.py
try new method

@@ -123,7 +123,7 @@ class WeightedLoader:
 
 def decomp_single(path):
     with open(path, 'rb') as fh:
-        return lz4.frame.decompress(fh.read())
+        return torch.load(io.BytesIO(lz4.frame.decompress(fh.read()), map_location='cpu')
 
 
 def decompress_multi_ppe(paths, threads):
@@ -163,9 +163,9 @@ def decompress_multi_map(paths, threads):
     start = datetime.now()
     decompressed = []
     with mp.Pool(threads) as pool:
-        decompressed = pool.map(decomp_single, paths)
+        result = pool.map(decomp_single, paths)
     
-    result = [torch.load(io.BytesIO(d), map_location='cpu') for d in decompressed]
+    #result = [torch.load(io.BytesIO(d), map_location='cpu') for d in decompressed]
            
     elapsed = datetime.now() - start
     logger.info(

### 2022-10-14 dnaseq2seq/loader.py
try new method

@@ -123,7 +123,7 @@ class WeightedLoader:
 
 def decomp_single(path):
     with open(path, 'rb') as fh:
-        return torch.load(io.BytesIO(lz4.frame.decompress(fh.read()), map_location='cpu')
+        return torch.load(io.BytesIO(lz4.frame.decompress(fh.read())), map_location='cpu')
 
 
 def decompress_multi_ppe(paths, threads):

### 2022-10-14 dnaseq2seq/loader.py
try new method

@@ -201,6 +201,7 @@ class PregenLoader:
         logger.info(f"Creating PreGen data loader with {self.threads} threads")
         logger.info(f"Found {len(self.pathpairs)} batches in {datadir}")
         logger.info(f"Possible sharing strategies: {mp.get_all_sharing_strategies()}")
+        mp.set_sharing_strategy("file_system")
         logger.info(f"Current sharing strategy: {mp.get_sharing_strategy()}")
         if not self.pathpairs:
             raise ValueError(f"Could not find any files in {datadir}")

### 2022-10-14 dnaseq2seq/loader.py
try new method

@@ -121,9 +121,9 @@ class WeightedLoader:
             yield self.src[idx, :, :, :].to(self.device), self.tgt[idx, :, :].to(self.device)
 
 
-def decomp_single(path):
+def decomp_single(path, queue):
     with open(path, 'rb') as fh:
-        return torch.load(io.BytesIO(lz4.frame.decompress(fh.read())), map_location='cpu')
+        queue.put(torch.load(io.BytesIO(lz4.frame.decompress(fh.read())), map_location='cpu'))
 
 
 def decompress_multi_ppe(paths, threads):
@@ -135,14 +135,18 @@ def decompress_multi_ppe(paths, threads):
     start = datetime.now()
     result = []
     futs = []
+    q = mp.Queue()
+    dfunc = functools.partial(decomp_single, queue=q)
     with ProcessPoolExecutor(threads) as pool:
         for path in paths:
-            fut = pool.submit(decomp_single, path)
+            fut = pool.submit(dfunc, path)
             futs.append(fut)
 
     for f in futs:
-        data = f.result(timeout=20)
-        result.append(torch.load(io.BytesIO(data), map_location='cpu'))
+        f.result(timeout=20)
+        r = queue.get()
+        logger.info(f"Got : {r}")
+        result.append(r)
 
     #for i, r in enumerate(result):
     #    logger.info(f"Result item {i}: {r.shape}")

### 2022-10-14 dnaseq2seq/loader.py
try new method

@@ -24,6 +24,7 @@ from datetime import datetime
 import traceback as tb
 from concurrent.futures import ProcessPoolExecutor
 import io
+import functools
 
 import scipy.stats as stats
 import numpy as np
@@ -205,7 +206,7 @@ class PregenLoader:
         logger.info(f"Creating PreGen data loader with {self.threads} threads")
         logger.info(f"Found {len(self.pathpairs)} batches in {datadir}")
         logger.info(f"Possible sharing strategies: {mp.get_all_sharing_strategies()}")
-        mp.set_sharing_strategy("file_system")
+        #mp.set_sharing_strategy("file_system")
         logger.info(f"Current sharing strategy: {mp.get_sharing_strategy()}")
         if not self.pathpairs:
             raise ValueError(f"Could not find any files in {datadir}")
@@ -239,7 +240,7 @@ class PregenLoader:
         for i in range(0, len(self.pathpairs), self.max_decomped):
             decomp_start = datetime.now()
             paths = self.pathpairs[i:i+self.max_decomped]
-            decomped = decompress_multi_map(chain.from_iterable(paths), self.threads)
+            decomped = decompress_multi_ppe(chain.from_iterable(paths), self.threads)
             decomp_end = datetime.now()
             decomp_time = (decomp_end - decomp_start).total_seconds()
 

### 2022-10-14 dnaseq2seq/loader.py
try new method

@@ -124,7 +124,9 @@ class WeightedLoader:
 
 def decomp_single(path, queue):
     with open(path, 'rb') as fh:
+        logger.info("PUtting item in queue...")
         queue.put(torch.load(io.BytesIO(lz4.frame.decompress(fh.read())), map_location='cpu'))
+    logger.info("func returning...")
 
 
 def decompress_multi_ppe(paths, threads):
@@ -143,9 +145,11 @@ def decompress_multi_ppe(paths, threads):
             fut = pool.submit(dfunc, path)
             futs.append(fut)
 
+    logger.info(f"Submitted all the funcs, now waiting for results, length: {len(q)}")
     for f in futs:
+        logger.info("Waiting for item...")
         f.result(timeout=20)
-        r = queue.get()
+        r = q.get()
         logger.info(f"Got : {r}")
         result.append(r)
 

### 2022-10-14 dnaseq2seq/loader.py
try new method

@@ -135,6 +135,7 @@ def decompress_multi_ppe(paths, threads):
     but keep them on the CPU
     :returns : List of Tensors (all on CPU)
     """
+    logger.info("Hey we are in the function")
     start = datetime.now()
     result = []
     futs = []

### 2022-10-14 dnaseq2seq/loader.py
try new method

@@ -123,8 +123,9 @@ class WeightedLoader:
 
 
 def decomp_single(path, queue):
+    logger.info("Entering function")
     with open(path, 'rb') as fh:
-        logger.info("PUtting item in queue...")
+        logger.info("Putting item in queue...")
         queue.put(torch.load(io.BytesIO(lz4.frame.decompress(fh.read())), map_location='cpu'))
     logger.info("func returning...")
 
@@ -141,8 +142,10 @@ def decompress_multi_ppe(paths, threads):
     futs = []
     q = mp.Queue()
     dfunc = functools.partial(decomp_single, queue=q)
+    logger.info(f"Making the pool")
     with ProcessPoolExecutor(threads) as pool:
         for path in paths:
+            logger.info(f"Submitting path: {path}")
             fut = pool.submit(dfunc, path)
             futs.append(fut)
 

### 2022-10-14 dnaseq2seq/loader.py
try new method

@@ -125,8 +125,10 @@ class WeightedLoader:
 def decomp_single(path, queue):
     logger.info("Entering function")
     with open(path, 'rb') as fh:
+        logger.info("Decompressing...")
+        d = io.BytesIO(lz4.frame.decompress(fh.read()))
         logger.info("Putting item in queue...")
-        queue.put(torch.load(io.BytesIO(lz4.frame.decompress(fh.read())), map_location='cpu'))
+        queue.put(torch.load(d, map_location='cpu'))
     logger.info("func returning...")
 
 

### 2022-10-15 dnaseq2seq/loader.py
profile loading

@@ -189,6 +189,34 @@ def decompress_multi_map(paths, threads):
     return result
 
 
+def decomp_profile(paths, threads):
+    result = []
+    read_sum = 0
+    decomp_sum = 0
+    load_sum = 0
+    for path in paths:
+        start = datetime.now()
+        with open(path, 'rb') as fh:
+            r = fh.read()
+        read_dt = datetime.now()
+        d = io.BytesIO(lz4.frame.decompress(r))
+        decomp_dt = datetime.now()
+        t = torch.load(d, map_location='cpu')
+        load_dt = datetime.now()
+        result.append(t)        
+        read_sum += read_dt - start
+        decomp_sum += decomp_dt - read_dt
+        load_sum += load_dt - decomp-dt
+
+    tot = read_sum + decomp_sum + load_sum
+    logger.info(f"Decomped {len(paths)} items")
+    logger.info(f"Total decomp time (secs): {tot.total_seconds() :.6f}")
+    logger.info(f"Read frac: {read_sum.total_seconds() / tot.total_seconds() * 100 :.3f}")
+    logger.info(f"Decomp frac: {decomp_sum.total_seconds() / tot.total_seconds() * 100 :.3f}")
+    logger.info(f"Load frac: {load_sum.total_seconds() / tot.total_seconds() * 100 :.3f}")
+    return result
+
+
 class PregenLoader:
 
     def __init__(self, device, datadir, threads, max_decomped_batches=10, src_prefix="src", tgt_prefix="tgt", vaftgt_prefix="vaftgt", pathpairs=None):
@@ -250,7 +278,7 @@ class PregenLoader:
         for i in range(0, len(self.pathpairs), self.max_decomped):
             decomp_start = datetime.now()
             paths = self.pathpairs[i:i+self.max_decomped]
-            decomped = decompress_multi_ppe(chain.from_iterable(paths), self.threads)
+            decomped = decompress_profile(chain.from_iterable(paths), self.threads)
             decomp_end = datetime.now()
             decomp_time = (decomp_end - decomp_start).total_seconds()
 

### 2022-10-15 dnaseq2seq/loader.py
profile loading

@@ -278,7 +278,7 @@ class PregenLoader:
         for i in range(0, len(self.pathpairs), self.max_decomped):
             decomp_start = datetime.now()
             paths = self.pathpairs[i:i+self.max_decomped]
-            decomped = decompress_profile(chain.from_iterable(paths), self.threads)
+            decomped = decomp_profile(chain.from_iterable(paths), self.threads)
             decomp_end = datetime.now()
             decomp_time = (decomp_end - decomp_start).total_seconds()
 

### 2022-10-15 scripts/run_train_clone.sh

@@ -48,8 +48,8 @@ LEARNING_RATE=0.00005
 
 CHECKPOINT_FREQ=1
 
-RUN_NAME="test_deadlock4"
-RUN_NOTES="deadlock testing"
+RUN_NAME="profile_loading"
+RUN_NOTES="profiling exprs"
 
 set -x
 

### 2022-10-15 dnaseq2seq/loader.py
profile loading

@@ -204,9 +204,10 @@ def decomp_profile(paths, threads):
         t = torch.load(d, map_location='cpu')
         load_dt = datetime.now()
         result.append(t)        
-        read_sum += read_dt - start
-        decomp_sum += decomp_dt - read_dt
-        load_sum += load_dt - decomp-dt
+        
+        read_sum = read_sum + (read_dt - start)
+        decomp_sum = decomp_sum + (decomp_dt - read_dt)
+        load_sum = load_sum + (load_dt - decomp_dt)
 
     tot = read_sum + decomp_sum + load_sum
     logger.info(f"Decomped {len(paths)} items")

### 2022-10-15 dnaseq2seq/loader.py
profile loading

@@ -191,9 +191,9 @@ def decompress_multi_map(paths, threads):
 
 def decomp_profile(paths, threads):
     result = []
-    read_sum = 0
-    decomp_sum = 0
-    load_sum = 0
+    read_sum = datetime.timedelta(0)
+    decomp_sum = datetime.timedelta(0)
+    load_sum = datetime.timedelta(0)
     for path in paths:
         start = datetime.now()
         with open(path, 'rb') as fh:

### 2023-03-31 dnaseq2seq/train.py
Remove SW loss stuff

@@ -28,7 +28,6 @@ import vcf
 import loader
 import util
 from model import VarTransformer
-from swloss import SmithWatermanLoss
 
 ENABLE_WANDB = os.getenv('ENABLE_WANDB', False)
 
@@ -442,20 +441,8 @@ def train_epochs(epochs,
         logger.info("Creating N.L.L. Loss loss function")
         criterion = nn.NLLLoss()
         scheduler = torch.optim.lr_scheduler.StepLR(optimizer, 1.0, gamma=0.995)
-    elif lossfunc == 'sw':
-        gap_open_penalty = -5
-        gap_exend_penalty = -1
-        temperature = 1.0
-        trim_width = 100
-        logger.info(f"Creating Smith-Waterman loss function with gap open: {gap_open_penalty} extend: {gap_exend_penalty} temp: {temperature:.4f}, trim_width: {trim_width}")
-        criterion = SmithWatermanLoss(gap_open_penalty=gap_open_penalty,
-                                    gap_extend_penalty=gap_exend_penalty,
-                                    temperature=temperature,
-                                    trim_width=trim_width,
-                                    device=DEVICE,
-                                    reduction=None,
-                                    window_mode="random")
-        scheduler = torch.optim.lr_scheduler.StepLR(optimizer, 1.0, gamma=0.90)
+    else:
+        raise NotImplemented("Only cross entropy loss supported for now")
 
     trainlogpath = str(model_dest).replace(".model", "").replace(".pt", "") + "_train.log"
     logger.info(f"Training log data will be saved at {trainlogpath}")

### 2023-03-31 dnaseq2seq/call.py
Try fixing multialt issue

@@ -535,6 +535,40 @@ def call_and_merge(batch, batch_offsets, regions, model, reference, max_batch_si
     return hap0, hap1
 
 
+def merge_multialts(v0, v1):
+    """
+    Merge two VcfVar objects into a single one with two alts
+
+    ATCT   G
+    AT     G
+      -->
+    ATCT   G,GCT
+
+    """
+    assert v0.pos == v1.pos
+    assert v0.het and v1.het
+    if v0.ref == v1.ref:
+        v0.alts.append(v1.alt)
+        v0.qual = (v0.qual + v1.qual) / 2  # Average quality??
+        v0.genotype = (1,2)
+        return v0
+    elif v0.alts[0] == v1.alts[0]:
+        longer, shorter = sorted([v0, v1], key=lambda x: len(x.ref))
+        extra_ref = longer.ref[len(shorter.ref):]
+        newalt = shorter.alts[0] + extra_ref
+        longer.alts.append(newalt)
+        longer.qual = (longer.qual + shorter.qual) / 2
+        longer.genotype = (1, 2)
+        return longer
+    else:
+        raise Exception(f"What? Two variant with different refs and alts at same position: {v0} and {v1}")
+
+
+
+
+
+
+
 def vars_hap_to_records(
     chrom, window_idx, vars_hap0, vars_hap1, aln, reference, classifier_model, vcf_template, var_freq_file
 ):
@@ -550,10 +584,23 @@ def vars_hap_to_records(
         reference=reference
     )
 
+    bypos = defaultdict(list)
+    for v in vcf_vars:
+        bypos[v.pos].append(v)
+
+    mergedvars = []
+    for pos, vars in bypos.items():
+        if len(vars) == 2:
+            multialt = merge_multialts(vars[0], vars[1])
+            mergedvars.append(multialt)
+        else:
+            mergedvars.append(vars[0])
+
+
     # covert variants to pysam vcf records
     vcf_records = [
         vcf.create_vcf_rec(var, vcf_template)
-        for var in sorted(vcf_vars, key=lambda x: x.pos)
+        for var in sorted(mergedvars, key=lambda x: x.pos)
     ]
 
     for rec in vcf_records:

### 2023-03-31 dnaseq2seq/vcf.py

@@ -53,7 +53,7 @@ class VcfVar:
     chrom: str
     pos: int
     ref: str
-    alt: str
+    # alt: str
     quals: list
     qual: float
     filter: list
@@ -72,6 +72,11 @@ class VcfVar:
     genotype: tuple
     het: bool
     duplicate: bool
+    alts: list
+
+    @property
+    def alt(self):
+        return self.alts[0]
 
 
 def _cigtups(cigstr):
@@ -273,6 +278,7 @@ def var_depth(var, chrom, aln):
 def vcf_vars(vars_hap0, vars_hap1, chrom, window_idx, aln, reference, mindepth=30):
     """
     From hap0 and hap1 lists of vars (pos ref alt qual) create vcf variant record information for entire call window
+    This creates VcfVar objects, which are almost like pysam.VariantRecord objects but not quite
     :param vars_hap0:
     :param vars_hap1:
     :param chrom:
@@ -290,7 +296,7 @@ def vcf_vars(vars_hap0, vars_hap1, chrom, window_idx, aln, reference, mindepth=3
             chrom=chrom,
             pos=var[0] + 1,
             ref=var[1],
-            alt=var[2],
+            alts=[var[2]],
             quals=[call.qual for call in vars_hap0[var]],
             qual=float(np.mean([call.qual for call in vars_hap0[var]])),
             filter=[],
@@ -319,7 +325,7 @@ def vcf_vars(vars_hap0, vars_hap1, chrom, window_idx, aln, reference, mindepth=3
             chrom=chrom,
             pos=var[0] + 1,
             ref=var[1],
-            alt=var[2],
+            alts=[var[2]],
             quals=[call.qual for call in vars_hap1[var]],
             qual=float(np.mean([call.qual for call in vars_hap1[var]])),
             filter=[],
@@ -357,8 +363,9 @@ def vcf_vars(vars_hap0, vars_hap1, chrom, window_idx, aln, reference, mindepth=3
 
     # combine haplotypes
     assert len(set(vcfvars_hap0) & set(vcfvars_hap1)) == 0, (
-        "There should be no vars shared between vcfvars_hap0 and vcfvars_hap1 (hom vars should all be in hap0"
+        "There should be no vars shared between vcfvars_hap0 and vcfvars_hap1 (hom vars should all be in hap0)"
     )
+
     vcfvars = {**vcfvars_hap0, **vcfvars_hap1}
 
     for key, var in vcfvars.items():
[…]

### 2023-03-31 dnaseq2seq/call.py
Fix multialt stuff

@@ -535,14 +535,14 @@ def call_and_merge(batch, batch_offsets, regions, model, reference, max_batch_si
     return hap0, hap1
 
 
-def merge_multialts(v0, v1):
+def merge_multialts(reference, v0, v1):
     """
     Merge two VcfVar objects into a single one with two alts
 
     ATCT   G
-    AT     G
+    A     GCAC
       -->
-    ATCT   G,GCT
+    ATCT   G,GCACT
 
     """
     assert v0.pos == v1.pos
@@ -552,19 +552,15 @@ def merge_multialts(v0, v1):
         v0.qual = (v0.qual + v1.qual) / 2  # Average quality??
         v0.genotype = (1,2)
         return v0
-    elif v0.alts[0] == v1.alts[0]:
-        longer, shorter = sorted([v0, v1], key=lambda x: len(x.ref))
+
+    else:
+        shorter, longer = sorted([v0, v1], key=lambda x: len(x.ref))
         extra_ref = longer.ref[len(shorter.ref):]
         newalt = shorter.alts[0] + extra_ref
         longer.alts.append(newalt)
         longer.qual = (longer.qual + shorter.qual) / 2
         longer.genotype = (1, 2)
         return longer
-    else:
-        raise Exception(f"What? Two variant with different refs and alts at same position: {v0} and {v1}")
-
-
-
 
 
 
@@ -591,7 +587,7 @@ def vars_hap_to_records(
     mergedvars = []
     for pos, vars in bypos.items():
         if len(vars) == 2:
-            multialt = merge_multialts(vars[0], vars[1])
+            multialt = merge_multialts(reference, vars[0], vars[1])
             mergedvars.append(multialt)
         else:
             mergedvars.append(vars[0])

### 2023-03-31 dnaseq2seq/vcf.py

@@ -378,7 +378,7 @@ def vcf_vars(vars_hap0, vars_hap1, chrom, window_idx, aln, reference, mindepth=3
             var.filter.append("SingleCallHom")
 
         # adjust insertions and deletions so no blank ref or alt
-        if var.ref == "" or any(v == "" for v in var.alts[0]):
+        if var.ref == "" or any(v == "" for v in var.alts):
             leading_ref_base = reference.fetch(reference=var.chrom, start=var.pos - 2, end=var.pos - 1)
             var.pos = var.pos - 1
             var.ref = leading_ref_base + var.ref

### 2023-03-31 manuscript/manuscript.tex

@@
 \title{Jovian enables direct inference of germline haplotypes from short reads via sequence-to-sequence modeling}
-\author{Author names witheld during review}
-%\author[1]{Brendan O'Fallon}
-%\author[1]{Ashini Bolia}
-%\author[1]{Jacob Durtschi}
-%\author[1]{Luobin Yang}
-%\author[1]{Eric Frederickson}
-%\author[1]{Katherine Noble}
-%\author[1]{Joshua Coleman}
-%\author[1]{Hunter Best}
+\author[1]{Brendan O'Fallon}
+\author[1]{Ashini Bolia}
+\author[1]{Jacob Durtschi}
+\author[1]{Luobin Yang}
+\author[1]{Eric Frederickson}
+\author[1]{Katherine Noble}
+\author[1]{Joshua Coleman}
+\author[1]{Hunter Best}
 
-%\affil[1]{ARUP Institute for Clinical and Experimental Pathology, Salt Lake City, UT}
+\affil[1]{ARUP Institute for Clinical and Experimental Pathology, Salt Lake City, UT}
 \date{}
@@
 \begin{abstract}
-	Detection of germline variants in next-generation sequencing data is an essential component of modern genomics analysis.
+	Detection of germline variants in next-generation sequencing data is an essential component of modern genomics.
 Variant detection tools typically rely on statistical algorithms such as de Bruijn graphs, Hidden Markov Models and regression models, often coupled with heuristic techniques and thresholds to identify variants.
@@
 
-Accurate inference of the DNA sequence variants in a sample is a fundamental element of genomics analysis.
+Inference of the DNA sequence variants in a sample is one of the most common and important tasks in modern genomics.
 Tools that aim to identify variants must overcome several challenges to produce an accurate set of variant calls.
@@
 
-Here, we describe a new approach that replaces both the candidate allele generation and the likelihood calculation with a single, learned model.
+Here, we describe a new approach that replaces both the candidate allele generation and the likelihood calculation with a single learned model.
 We recast variant detection as a sequence-to-sequence translation problem in which the input sequence is the series of bases aligning to a single genomic location, and the output sequences are the two predicted haplotypes.
@@
 Our analysis demonstrates that the model learns many features of modern, state of the art variant callers.
-For instance, our model learns to reconstruct large, complex variants in the presence of ambiguous mapping and extensive soft clipping with a sensitivity near 90\%.
+For instance, our model learns to reconstruct large, complex variants in the presence of ambiguous mapping and extensive soft-clipping with a sensitivity near 90\%.
 In addition, the model learns to employ local context and read-specific features to predict phase and genotype accurately, and pairs of variants are phased correctly in 97-99\% of cases where variants are within 100bp.

### 2023-11-21 manuscript/manuscript.tex
Add jitter to pregen

@@
 
-\title{Jovian enables direct inference of germline haplotypes from short reads via sequence-to-sequence modeling}
+\title{Jovian enables direct inference of germline haplotypes from short reads via generative modeling}
 \author[1]{Brendan O'Fallon}
@@
 \begin{abstract}
-	Detection of germline variants in next-generation sequencing data is an essential component of modern genomics.
+	Detection of germline variants in next-generation sequencing data is an fundamenal component of modern genomics.
 Variant detection tools typically rely on statistical algorithms such as de Bruijn graphs, Hidden Markov Models and regression models, often coupled with heuristic techniques and thresholds to identify variants.
@@
 2016).
-In addition, large or complex sequence variants may be obscured by limitations in the read alignment algorithm, and require a sophisticated assembly of reads to identify.
+In addition, large or complex sequence variants may be obscured by limitations in the read alignment algorithm, and require a sophisticated reassembly of reads to identify.
 Reconstructing real variation and discriminating it from technical error is challenging, and modern variant detection tools employ a variety of statistical and heuristic techniques to achieve high precision and sensitivity (e.g.
@@
 
-Here, we describe a new approach that replaces both the candidate allele generation and the likelihood calculation with a single learned model.
-We recast variant detection as a sequence-to-sequence translation problem in which the input sequence is the series of bases aligning to a single genomic location, and the output sequences are the two predicted haplotypes.
+Here, we describe a new approach that replaces both the candidate allele generation and the likelihood calculation with a single generative model that is learned end-to-end.
+We recast variant detection as a generative modeling problem in which the outputs are the two haplotype sequences present in the sample.
+From these generated haplotypes we recover accurate variant calls.
+Overall, our model architecture resembles that of modern Large Language Models (LLMs).
+LLMs typically take as input a series of word fragments (tokens) and iteratively create new output tokens, where the model predicts each output token using the list of input tokens as well as the output tokens already generated.
+In contrast, our model uses columns of aligned bases as input tokens, and predicts two output sequences, one for each diploid haplotype present in the sample.
 The sequence-to-sequence translation is learned by a transformer-based deep neural network (Vaswani et al.
@@
 
-Our model consists of a transformer-based encoder and a simple, fixed-length decoder that produces two output sequences, one for each predicted haplotype.
-The encoder component is an unmodified transformer with GeLU activations as implemented in PyTorch 1.10 (Vaswani et al.
+Our model consists of a transformer-based encoder and dual, parallel decoders that each produce a single output haplotype sequence.
+The encoder component is an unmodified transformer with GeLU activations as implemented in PyTorch 2.10 (Vaswani et al.
 2017, Paszke et al.
@@
 The embedded basecalls are `flattened' along the read dimension, producing an input token with size $dr$ where $r$ is the number of reads ($r=100$ for all experiments proposed here).
-Instead of the typical 1-dimensional positional encoding, we employ a 2-dimensional encoding, allowing the model to differentiate across both the token (position) and read dimensions.
+Instead of the typical 1-dimensional positional encoding we employ a 2-dimensional encoding, allowing the model to differentiate across both the token (position) and read dimensions.
 The 2D encoding implementation follows Wang \& Jyh-Charn (2021).
 
-The decoder consists of a single fully-connected layer with ReLU activation followed by two additional fully-connected components that each generate a single haplotype.
-The decoder is not autoregressive and does not utilize special start-of-sequence or end-of-sequence tokens.
-Predictions are discrete probability distributions over the four possible bases at each predicted output position, produced with a softmax function over the raw outputs.
-The decoder is of fixed length and produces two haplotype sequences equal in length to the number of input tokens $g$.
+The two decoders each consist of stacked transformer decoder layers with GeLU activations.
+Each decoder generates a k-mer of length 4 on each prediction step.
+Because there are $4^4=256$ possible 4-mers the decoders have dimension 256, which is not directly compatible with the encoder dimension, so we add an additional linear layer between the encoder outputs and the decoder layers to convert the model dimensions accordingly.
 The full model architecture is shown in Figure \ref{fig:architecture}.
@@
 
-Training data was obtained from 17 whole genomes sequenced from 5 Genome-in-a-Bottle cell lines.
+Training data was obtained from 28 whole genomes sequenced from 7 Genome-in-a-Bottle cell lines.
 Eight samples were prepared with Illumina Nextera DNA Flex kit and the remainder were prepared with the Illumina TruSeq PCR-free kit.
@@
 
-Models were trained for 50 epochs unless noted otherwise, using the Adam optimizer, a learning rate of $5e^{-5}$, and batch size of 512. Models were implemented in PyTorch (v1.10, Paszke et al 2019).
+Models were trained for 50 epochs unless noted otherwise, using the Adam optimizer, a learning rate of $5e^{-5}$, and batch size of 512. Models were implemented in PyTorch (v2.10, Paszke et al 2019).
 

### 2023-11-21 dnaseq2seq/loader.py

@@ -71,7 +71,7 @@ class LazyLoader:
     Useful for 'pre-gen' where we just want to iterate over everything once and save it to a file
     """
 
-    def __init__(self, bam, bed, vcf, reference, reads_per_pileup, samples_per_pos, vals_per_class):
+    def __init__(self, bam, bed, vcf, reference, reads_per_pileup, samples_per_pos, vals_per_class, max_jitter_bases):
         self.bam = bam
         self.bed = bed
         self.vcf = vcf
@@ -79,19 +79,21 @@ class LazyLoader:
         self.reads_per_pileup = reads_per_pileup
         self.samples_per_pos = samples_per_pos
         self.vals_per_class = vals_per_class
+        self.max_jitter_bases = max_jitter_bases
 
     def iter_once(self, batch_size):
         logger.info(f"Encoding tensors from {self.bam} and {self.vcf}")
         for src, tgt, vaftgt, varsinfo in encode_chunks(self.bam,
-                                          self.reference,
-                                          self.bed,
-                                          self.vcf,
-                                          batch_size,
-                                          self.reads_per_pileup,
-                                          self.samples_per_pos,
-                                          self.vals_per_class,
-                                          max_to_load=1e9):
-                yield src, tgt, vaftgt, varsinfo
+                                                        self.reference,
+                                                        self.bed,
+                                                        self.vcf,
+                                                        batch_size,
+                                                        self.reads_per_pileup,
+                                                        self.samples_per_pos,
+                                                        self.vals_per_class,
+                                                        max_to_load=1e9,
+                                                        max_jitter_bases=self.max_jitter_bases):
+            yield src, tgt, vaftgt, varsinfo
 
 
 class WeightedLoader:
@@ -503,7 +505,7 @@ def upsample_labels(bed, vals_per_class):
     return rows
 
 
-def load_from_csv(bampath, refpath, bed, vcfpath, max_reads_per_aln, samples_per_pos, vals_per_class):
+def load_from_csv(bampath, refpath, bed, vcfpath, max_reads_per_aln, samples_per_pos, vals_per_class, max_jitter_bases=0):
     """
[…]

### 2023-11-21 scripts/run_pregen.sh
script updates

@@ -21,9 +21,9 @@ ds2s=/uufs/chpc.utah.edu/common/home/u0379426/src/jovian/dnaseq2seq/main.py
 #CONF=/uufs/chpc.utah.edu/common/home/u0379426/src/dnaseq2seq/wgs_lcbig_sus_chrs21and22.yaml
 #CONF=/uufs/chpc.utah.edu/common/home/u0379426/src/dnaseq2seq/wgs_lcbig_sus_5more_chrsE.yaml
 #CONF=/uufs/chpc.utah.edu/common/home/u0379426/src/dnaseq2seq/decoder_fpfn_chr1_conf.yaml
-CONF=/uufs/chpc.utah.edu/common/home/u0379426/src/jovian/pregen_confs/pregen_d128_medium_chrE.yaml
+CONF=/uufs/chpc.utah.edu/common/home/u0379426/src/jovian/pregen_confs/pregen_d128_nosus_chrE.yaml
 
-DEST=/scratch/general/vast/u0379426/pregen/wgs_d128_medium_chrE
+DEST=/scratch/general/vast/u0379426/pregen/wgs_d128_nosus_chrE
 
 BATCH_SIZE=512
 READ_DEPTH=128
